### 2019-01-03 paper/main.tex
adding LaTeX template to be used in paper submission

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[short Title]{This is a title}
\author[Holt \textit{et~al}.]{James M. Holt\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
$^{\text{\sf 2}}$Department, Institution, City, Post Code,
Country.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} TODO.\\
\textbf{Results:} TODO. \\
\textbf{Availability:} TODO. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text. Figure~\ref{fig:01}
shows that the above method  Text Text Text Text  Text Text Text
Text Text Text  Text Text.  \citep{Bag01} wants to know about
{\ldots}{\ldots} text follows.
\begin{equation}
\sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
\end{equation}

%\enlargethispage{12pt}

\section{Approach}

Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
.....

\begin{methods}
\section{Methods}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text
text\vspace*{1pt}

\begin{itemize}
\item for bulleted list, use itemize
\item for bulleted list, use itemize
\item for bulleted list, use itemize\vspace*{1pt}
\end{itemize}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
Text Text Text Text Text Text Text Text Text Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text\vadjust{\newpage}.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
Text Text Text Text Text Text Text Text Text Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

\subsection{This is subheading}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text
text. Text Text Text Text Text Text  Text Text Text Text Text Text
Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

\subsubsection{This is subsubheading}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text\break
text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text
textText Text Text Text Text Text  Text Text Text Text Text Text
Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

\enlargethispage{6pt}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\begin{table}[!t]
\processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule head1 &
head2 & head3 & head4\\\midrule
row1 & row1 & row1 & row1\\
row2 & row2 & row2 & row2\\
row3 & row3 & row3 & row3\\
row4 & row4 & row4 & row4\\\botrule
\end{tabular}}{This is a footnote}
\end{table}

\end{methods}

\begin{figure}[!tpb]%figure1
\fboxsep=0pt\colorbox{gray}{\begin{minipage}[t]{235pt} \vbox to 100pt{\vfill\hbox to
235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}\vfill}
\end{minipage}}
%\centerline{\includegraphics{fig01.eps}}
\caption{Caption, caption.}\label{fig:01}
\end{figure}

%\begin{figure}[!tpb]%figure2
%%\centerline{\includegraphics{fig02.eps}}
%\caption{Caption, caption.}\label{fig:02}
%\end{figure}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\subsection{Test1}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\section{Discussion}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text.

Table~\ref{Tab:01} shows that Text Text Text Text Text  Text Text
Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
the above method Text Text. Text Text Text  Text Text Text Text
Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
method Text Text. Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
Text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}

(Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text
Text Text Text Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\begin{enumerate}
\item this is item, use enumerate
\item this is item, use enumerate
\item this is item, use enumerate
\end{enumerate}

Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
above method  Text Text Text Text Text Text Text Text Text Text
Text Text.  \citealp{Boffelli03} might want to know about text
text text text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text\break Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text\vspace*{-10pt}

\section*{Acknowledgements}

Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text
text text text\vspace*{-12pt}

\section*{Funding}

This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\begin{thebibliography}{}

\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

\bibitem[Bag {\it et~al}., 2001]{Bag01}
Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
in nigral dopaminergic neurnol cell: correlation with the known
pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
Res.}, \textbf{110}(Suppl. 1), 76--84.

\bibitem[Lehmann, 1986]{Leh86}
Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

\bibitem[Crenshaw and Jones, 2003]{Cre03}
Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
cancer management: one tumor, one chip. \textit{Bioinformatics},
doi:10.1093/bioinformatics/btn000.

\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
???--???.

\bibitem[Bardet, 1920]{Bar20}
Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
polydactylie et retinite pigmentaire (contribution a l'etude des
formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
institution, Paris, France.

\end{thebibliography}
\end{document}

### 2019-01-03 paper/main.tex
minor laying out of thoughts

@@
 \section{Introduction}
+WGS is currently being used as a molecular diagnostic tool for rare disease; blood draw + phenotypes are provided to analysis centers and these go through a standard alignment/variant calling pipeline
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text. Figure~\ref{fig:01}
-shows that the above method  Text Text Text Text  Text Text Text
-Text Text Text  Text Text.  \citep{Bag01} wants to know about
-{\ldots}{\ldots} text follows.
-\begin{equation}
-\sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
-\end{equation}
+Because of size of data, variants are typically filtered down based on rare disease expectations; these variants are manually checked by two or more analysts and reported to clinical site
+
+Manual curation is tedious and prone to error due to variant fatigue; thus tools that further reduce manual time on a case are paramount
 
@@
 
-Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-.....
+Here we test methods that take phenotype, gene, transcript, and variant level annotations from the filtered variant list and predict whether a variant will ultimately be reported back to the clinical site.
+
+The methods provide a boolean predictor and a ranking system.
 
@@
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text
-text\vspace*{1pt}
+\subsection{Variant annotation}
+CODI does most of it; we use PyxisMap and HPOUtil for the rest
 
-\begin{itemize}
-\item for bulleted list, use itemize
-\item for bulleted list, use itemize
-\item for bulleted list, use itemize\vspace*{1pt}
-\end{itemize}
+\subsection{Data cleaning}
+Numerical data is given a number with out-of-bounds defaults
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-Text Text Text Text Text Text Text Text Text Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
+Categorical data is XXX (what do we decide to do with it?)
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text\vadjust{\newpage}.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-Text Text Text Text Text Text Text Text Text Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
+\subsection{Model training and tuning}
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-\subsection{This is subheading}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text
-text. Text Text Text Text Text Text  Text Text Text Text Text Text
-Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-\subsubsection{This is subsubheading}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text\break
-text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text
-textText Text Text Text Text Text  Text Text Text Text Text Text
-Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-\enlargethispage{6pt}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\begin{table}[!t]
-\processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule head1 &
-head2 & head3 & head4\\\midrule
-row1 & row1 & row1 & row1\\
-row2 & row2 & row2 & row2\\
-row3 & row3 & row3 & row3\\
-row4 & row4 & row4 & row4\\\botrule
-\end{tabular}}{This is a footnote}
-\end{table}
+We tested a lot of models and tuned them
 
@@
 
-\begin{figure}[!tpb]%figure1
-\fboxsep=0pt\colorbox{gray}{\begin{minipage}[t]{235pt} \vbox to 100pt{\vfill\hbox to
-235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}\vfill}
-\end{minipage}}
-%\centerline{\includegraphics{fig01.eps}}
-\caption{Caption, caption.}\label{fig:01}
-\end{figure}
+\section{Discussion}
+Results go here
 
-%\begin{figure}[!tpb]%figure2
-%%\centerline{\includegraphics{fig02.eps}}
-%\caption{Caption, caption.}\label{fig:02}
-%\end{figure}
+Mathy stuff: Precision, recall, AUC, ROC, etc.
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\subsection{Test1}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\section{Discussion}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text.
-
-Table~\ref{Tab:01} shows that Text Text Text Text Text  Text Text
-Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
-the above method Text Text. Text Text Text  Text Text Text Text
-Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
-method Text Text. Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-Text.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Practical stuff: in a case, this pushed reported variants to rank X out of Y (or something like this)
 
@@
 
-(Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text
-Text Text Text Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\begin{enumerate}
-\item this is item, use enumerate
-\item this is item, use enumerate
-\item this is item, use enumerate
-\end{enumerate}
-
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
-above method  Text Text Text Text Text Text Text Text Text Text
-Text Text.  \citealp{Boffelli03} might want to know about text
-text text text Text Text Text Text Text Text  Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text\break Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text\vspace*{-10pt}
+We did a thing
 
 \section*{Acknowledgements}
+TODO
 
@@
 \section*{Funding}
+TODO
 

### 2019-01-04 scripts/TestLearners.py
updated to break down by case and send some simple rank values to screen

@@ -38,7 +38,10 @@ PYXIS_DATE='12192018'
 
 def getPyxisMapResults():
     '''
-    Script for downloading the PyxisMap results, update the date constant when needed
+    Script for pre-computing HPO-based results to the "pyxis_ranks_<PYXIS_DATE>" subfolder specified in the 
+    constant above.  This will do two things: 1) hit the PyxisMap endpoint specified in PyxisMapUtil and
+    2) perform the cosine score computation from HPOUtil.  Outputs are saved to files so we don't run this 
+    every time we want to train a new model.
     '''
     ontFN = '/Users/matt/githubProjects/LayeredGraph/HPO_graph_data/hp.obo'
     annotFN = '/Users/matt/githubProjects/LayeredGraph/HPO_graph_data/ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt'
@@ -50,6 +53,7 @@ def getPyxisMapResults():
     node, edges, parents, altIDMap = OntologyUtil.loadGraphStructure(ontFN)
     hpoOnt = HPOUtil.HPOUtil(ontFN, annotFN)
     
+    #load case data and go through them searching for HPO terms
     caseData = SummaryDBUtil.loadSummaryDatabase(altIDMap)
     for sl in sorted(caseData.keys()):
         #TODO: figure out how to properly handle this
@@ -104,9 +108,15 @@ def getPyxisMapResults():
                 else:
                     print('Failed to retrieve results for '+sl+' from HPOUtil')
 
-def runAnalysis():
+def loadFormattedData():
     '''
-    Core test function
+    Core test function.  This function is primarily about loading and formatting data prior to training/testing any 
+    classifiers.
+    @return - tuple (xFinal, yFinal, featureLabels, startIndices)
+        xFinal - an NxM matrix where N is the number of variants in our test/training set and M is the number of features; contains feature values
+        yFinal - an N length array where N is the number of variants in our test/training set; 1 if the variant was reported
+        featureLabels - an M length array containing feature labels for our output benefit
+        startIndices - a (C+1) length array containing the start indices of variants from an individual case
     '''
     pyxisRootDir = '/Users/matt/githubProjects/VarSight/pyxis_ranks_'+PYXIS_DATE
     
@@ -231,13 +241,16 @@ def runAnalysis():
     allCatValues = []
     allCatBreak = {}
     allClassifications = []
+    
+    #this will track the breaks between cases
+    startIndices = [0]
+
[…]

### 2019-01-04 paper/data_template.tex
dumping output to a JSON that is later used by LaTeX and Jinja2

Classifier&TPR&FPR&AUC&CV10 Acc. \\ \hline
{% for clf_label in CLF_LABELS %}
    {{ clf_label }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_TPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_FPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].ROC_AUC }}&{{ '%0.2f+-%0.2f' % (CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[0], CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[1]) }} \\
{% endfor %}

### 2019-01-04 scripts/SummaryDBUtil.py

@@ -23,11 +23,42 @@ def loadSummaryDatabase(altIDMap):
         ...
     }
     '''
+    #need to get UDN to SL##s
+    TRACKING_FILENAME = '/Users/matt/githubProjects/VarSight/data/UDN Research Tracking_v1.xlsx'
+    wbt = openpyxl.load_workbook(TRACKING_FILENAME)
+    mainSheet = wbt['Original case order']
+    mainDict = {}
+    udnToSL = {}
+    ignoreSet = set([None, 'SANGER ONLY', 'N/A', 'SANGER_ONLY', 'NO DNA LEFT'])
+    for i, row in enumerate(mainSheet):
+        if i == 0:
+            continue
+        elif i == 1:
+            for j, c in enumerate(row):
+                if c.value != None:
+                    mainDict[c.value] = j
+        else:
+            sl_raw = row[mainDict['HA_SL#']].value
+            udnID = row[mainDict['UDN#']].value
+
+            if (sl_raw not in ignoreSet) and udnID != None:
+                for slid in sl_raw.split('/'):
+                    sl = slid.rstrip('R')
+                    if len(sl) != 8:
+                        raise Exception('Unexpected slid:'+sl+' from '+sl_raw)
+                    
+                    if udnID in udnToSL:
+                        udnToSL[udnID].append(sl)
+                    else:
+                        udnToSL[udnID] = [sl]
+    
     #core info to read and fill in
-    SUMMARY_DB_FILENAME = '/Users/matt/githubProjects/VarSight/data/UDN Summary Database_v3.xlsx'
+    SUMMARY_DB_FILENAME = '/Users/matt/githubProjects/VarSight/data/UDN Summary Database_v4.xlsx'
     wb = openpyxl.load_workbook(SUMMARY_DB_FILENAME)
     ret = {}
     
+    '''
+    #this was the version used with v3, formatting changed though
     #first load HPO terms for each patient
     hpoSheet = wb['Patient HPO terms']
     hpoDict = {}
@@ -56,6 +87,36 @@ def loadSummaryDatabase(altIDMap):
                             else:
                                 #get the alternate ID if it exists, otherwise use the normal
                                 ret[slid]['hpoTerms'].add(altIDMap.get(h, h))
+    '''
+
+    #first load HPO terms for each patient
+    hpoSheet = wb['Phenotips_HPO_04Jan19']
+    hpoDict = {}
+    for i, row in enumerate(hpoSheet):
+        if i == 0:
[…]

### 2019-01-04 scripts/TestLearners.py

@@ -34,7 +34,7 @@ from sklearn.model_selection import StratifiedKFold
 from sklearn.model_selection import train_test_split
 from sklearn.neural_network import MLPClassifier
 
-PYXIS_DATE='12192018'   
+PYXIS_DATE='01042019'   
 
 def getPyxisMapResults():
     '''
@@ -112,11 +112,12 @@ def loadFormattedData():
     '''
     Core test function.  This function is primarily about loading and formatting data prior to training/testing any 
     classifiers.
-    @return - tuple (xFinal, yFinal, featureLabels, startIndices)
+    @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts)
         xFinal - an NxM matrix where N is the number of variants in our test/training set and M is the number of features; contains feature values
         yFinal - an N length array where N is the number of variants in our test/training set; 1 if the variant was reported
         featureLabels - an M length array containing feature labels for our output benefit
         startIndices - a (C+1) length array containing the start indices of variants from an individual case
+        allRepDicts - an N length array of None or dictionaries; if a dictionary, it contains data on a returned variant
     '''
     pyxisRootDir = '/Users/matt/githubProjects/VarSight/pyxis_ranks_'+PYXIS_DATE
     
@@ -241,6 +242,7 @@ def loadFormattedData():
     allCatValues = []
     allCatBreak = {}
     allClassifications = []
+    allRepDicts = []
     
     #this will track the breaks between cases
     startIndices = [0]
@@ -251,6 +253,8 @@ def loadFormattedData():
             print('Skipping '+sl+', no primaries')
             continue
         primarySet = set([p['variant'] for p in caseData[sl]['primaries']])
+        primaryDict = {p['variant'] : p for p in caseData[sl]['primaries']}
+        assert(len(primaryDict) == len(caseData[sl]['primaries']))
         
         #if there is no CODICEM dump, we obviously can't do anything either
         codiDump = CodiDumpUtil.loadCodiDump(sl, fieldsOnly, geneFieldsOnly, seqFieldsOnly, transLabelsOnly)
@@ -290,6 +294,7 @@ def loadFormattedData():
         catValues = []
         catBreak = {}
         classifications = []
+        repDicts = []
         
         print('Primary set: ', primarySet)
         foundPrimaries = set([])
@@ -304,6 +309,9 @@ def loadFormattedData():
[…]

### 2019-01-04 paper/data_template.tex
base level table is filled out, probably time to start writing :(

@@
-Classifier&TPR&FPR&AUC&CV10 Acc. \\ \hline
+\multirow{2}{*}{Classifier}&\multicolumn{4}{c||}{Performance}&\multicolumn{4}{c}{Rank Median(Mean)} \\ \cline{2-9}
+&TPR&FPR&AUC&CV10 Acc.&Overall (n={{ TEST_COUNTS.OVERALL }})&
+VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
+Likely Path. (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Pathogenic (n={{ TEST_COUNTS.PATHOGENIC }}) \\ \hline
 {% for clf_label in CLF_LABELS %}
-    {{ clf_label }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_TPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_FPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].ROC_AUC }}&{{ '%0.2f+-%0.2f' % (CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[0], CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[1]) }} \\
+    {{ clf_label }}&
+    {{ '%0.2f' % CLASSIFIERS[clf_label].TEST_TPR_RATE }}&
+    {{ '%0.2f' % CLASSIFIERS[clf_label].TEST_FPR_RATE }}&
+    {{ '%0.2f' % CLASSIFIERS[clf_label].ROC_AUC }}&
+    {{ '%0.2f+-%0.2f' % (CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[0], CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[1]) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.OVERALL), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.OVERALL)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.LIKELY_PATHOGENIC), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.LIKELY_PATHOGENIC)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.PATHOGENIC), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.PATHOGENIC)) }} \\
 {% endfor %}

### 2019-01-04 scripts/TestLearners.py

@@ -507,6 +507,7 @@ def runClassifiers(values, classifications, featureLabels, startIndices, allRepD
     '''
     
     resultsDict = {}
+    resultsDict['FEATURE_LABELS'] = featureLabels
     print('Values:', values.shape)
     
     #V1
@@ -773,6 +774,12 @@ def runClassifiers(values, classifications, featureLabels, startIndices, allRepD
                 print('\t\tstdev:', np.std(rDict[pDict[v]]))
                 print('\t\tmedian:', np.median(rDict[pDict[v]]))
                 resultsDict['CLASSIFIERS'][clf_label]['TEST_RANKINGS'][v] = rDict[pDict[v]]
+            
+            #this will get written multiple times but that's okay
+            resultsDict['TEST_COUNTS'] = {}
+            resultsDict['TEST_COUNTS']['OVERALL'] = len(ranks)
+            for p in pList:
+                resultsDict['TEST_COUNTS'][p] = len(rDict[pDict[p]])
 
         if currentMode == EXACT_MODE:
             #cross validation scores
@@ -851,6 +858,10 @@ def jsonDumpFix(o):
     raise TypeError(str(o)+' '+str(type(o)))
 
 def generateLaTeXResult(d):
+    '''
+    This function takes a dictionary of values from our results and makes the calls to render those results through
+    Jinja2 into LaTeX files for the paper.
+    '''
     from jinja2 import Environment, FileSystemLoader
     rootPath = '/Users/matt/githubProjects/VarSight/paper/'
     env = Environment(loader=FileSystemLoader(rootPath))
@@ -880,7 +891,7 @@ def runAnalysis():
     fp.close()
 
     print(json.dumps(resultsDict, indent=4, sort_keys=True))
-
+    resultsDict['np'] = np
     generateLaTeXResult(resultsDict)
 
 if __name__ == '__main__':

### 2019-01-08 scripts/CodiDumpUtil.py
removed some prints that were annoying

@@ -21,7 +21,7 @@ def loadCodiDump(sl, EXTRA_FIELDS, GENE_FIELDS, SEQ_FIELDS, TRANSCRIPT_FIELDS):
     #make sure the file exists
     fn = '/Users/matt/githubProjects/VarSight/CLI_primary_V6/'+sl+'_results.json'
     if not os.path.exists(fn):
-        print('Missing file: '+fn)
+        #print('Missing file: '+fn)
         return []
     
     #load the json file

### 2019-01-08 scripts/SummaryDBUtil.py
big changes in where inputs came from due to events

@@ -2,6 +2,7 @@
 This file should contain helper functions to load the UDN summary DB information
 '''
 
+import csv
 import openpyxl
 
 def loadSummaryDatabase(altIDMap):
@@ -56,38 +57,6 @@ def loadSummaryDatabase(altIDMap):
     SUMMARY_DB_FILENAME = '/Users/matt/githubProjects/VarSight/data/UDN Summary Database_v4.xlsx'
     wb = openpyxl.load_workbook(SUMMARY_DB_FILENAME)
     ret = {}
-    
-    '''
-    #this was the version used with v3, formatting changed though
-    #first load HPO terms for each patient
-    hpoSheet = wb['Patient HPO terms']
-    hpoDict = {}
-    for i, row in enumerate(hpoSheet):
-        if i == 0:
-            for j, c in enumerate(row):
-                if c.value != None:
-                    hpoDict[c.value] = j
-        else:
-            slids = row[hpoDict['SL#']].value
-            hpo = row[hpoDict['Phenotips HPO']].value
-            
-            if slids != None:
-                for slid in slids.split(','):
-                    if hpo != None:
-                        #create this entry if it doesn't exist
-                        if (slid not in ret):
-                            ret[slid] = {
-                                "hpoTerms" : set([])
-                            }
-                        
-                        #add each HPO term
-                        for h in hpo.split(','):
-                            if h[0:3] != 'HP:' or len(h) != 10:
-                                print('Ignoring term:', h)
-                            else:
-                                #get the alternate ID if it exists, otherwise use the normal
-                                ret[slid]['hpoTerms'].add(altIDMap.get(h, h))
-    '''
 
     #first load HPO terms for each patient
     hpoSheet = wb['Phenotips_HPO_04Jan19']
@@ -118,6 +87,8 @@ def loadSummaryDatabase(altIDMap):
                             #get the alternate ID if it exists, otherwise use the normal
                             ret[slid]['hpoTerms'].add(altIDMap.get(h, h))
     
+    '''
+    #this was the original primary getter, but it changed, so I need to re-parse from a CSV
     #second load the primaries
     primarySheet = wb['Primary variants']
     ALLOWED_PATHOGENICITY = set(['PATHOGENIC', 'LIKELY_PATHOGENIC', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE'])
@@ -149,7 +120,50 @@ def loadSummaryDatabase(altIDMap):
[…]

### 2019-01-08 paper/data_template.tex
updated data outputs to include controls

--- 2019-01-04 paper/data_template.tex
@@
-Classifier&TPR&FPR&AUC&CV10 Acc. \\ \hline
+\multirow{2}{*}{Metric/Classifier}&\multicolumn{4}{c||}{Performance}&\multicolumn{4}{c}{Case Rank - Median(Mean)} \\ \cline{2-9}
+&CV10 Acc.&TPR&FPR&AUC&Overall (n={{ TEST_COUNTS.OVERALL }})&
+VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
+Likely Path. (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Pathogenic (n={{ TEST_COUNTS.PATHOGENIC }}) \\ \hhline{=|=|=|=|=||=|=|=|=}
+{% for cs_label in CS_LABELS %}
+    {{ cs_label }}&
+    --&
+    --&
+    --&
+    {{ '%0.2f' % COMPARISON[cs_label].ROC_AUC }}&
+    {{ '%0.1f (%0.2f)' % (np.median(COMPARISON[cs_label].TEST_RANKINGS.OVERALL), np.mean(COMPARISON[cs_label].TEST_RANKINGS.OVERALL)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(COMPARISON[cs_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE), np.mean(COMPARISON[cs_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(COMPARISON[cs_label].TEST_RANKINGS.LIKELY_PATHOGENIC), np.mean(COMPARISON[cs_label].TEST_RANKINGS.LIKELY_PATHOGENIC)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(COMPARISON[cs_label].TEST_RANKINGS.PATHOGENIC), np.mean(COMPARISON[cs_label].TEST_RANKINGS.PATHOGENIC)) }} \\
+{% endfor %}
+\hhline{=|=|=|=|=||=|=|=|=}
 {% for clf_label in CLF_LABELS %}
-    {{ clf_label }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_TPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_FPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].ROC_AUC }}&{{ '%0.2f+-%0.2f' % (CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[0], CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[1]) }} \\
+    {{ clf_label }}&
+    {{ '%0.2f+-%0.2f' % (CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[0], CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[1]) }}&
+    {{ '%0.2f' % CLASSIFIERS[clf_label].TEST_TPR_RATE }}&
+    {{ '%0.2f' % CLASSIFIERS[clf_label].TEST_FPR_RATE }}&
+    {{ '%0.2f' % CLASSIFIERS[clf_label].ROC_AUC }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.OVERALL), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.OVERALL)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.LIKELY_PATHOGENIC), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.LIKELY_PATHOGENIC)) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.PATHOGENIC), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.PATHOGENIC)) }} \\
 {% endfor %}

### 2019-01-08 paper/main.tex
dumping output to a JSON that is later used by LaTeX and Jinja2
base level table is filled out, probably time to start writing :(
updated data outputs to include controls

@@
 \documentclass{bioinfo}
+\usepackage{multirow}
+\usepackage{hhline}
 \copyrightyear{2015} \pubyear{2015}
@@
 \section{Introduction}
+
 WGS is currently being used as a molecular diagnostic tool for rare disease; blood draw + phenotypes are provided to analysis centers and these go through a standard alignment/variant calling pipeline
@@
 \section{Approach}
+In this paper, we tested the application of classification algorithms for predicting whether a given variant in a patient sample would be returned on a clinical report.
+To do this, we used patient samples from the UDN \citep{ramoni2017} who received genome sequencing and received at least one variant on their clinical report.
+We then extracted data directly from the annotation and filtering tool used by our case analysts in order to replicate their standard view of the patient.
+Additionally, we extracted a list of phenotype terms provided by the clinical site to pair with each patient.
+These phenotype terms were processed with two scoring systems to provide phenotype scores for each variant.
+Finally, each variant was either labeled as ``returned" or ``not returned" depending on whether it was ultimately sent back to the clinical site.
 
-Here we test methods that take phenotype, gene, transcript, and variant level annotations from the filtered variant list and predict whether a variant will ultimately be reported back to the clinical site.
-
-The methods provide a boolean predictor and a ranking system.
+Given the above variant information, we split the data into training and testing sets for measuring the performance of classifiers.
+We tested four classifiers that are readily available in the sklearn (CITE:XXX) and imblearn (CITE:XXX) Python modules.
+After tuning each classifier, we generated summaries of the performance of each measure from a boolean classification perspective and from an ordered variant ranking perspective.
 
@@
 \section{Methods}
+\subsection{Patient selection}
+All samples were selected from the cohort of genome sequencing samples originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB) Sequencing Center of the Undiagnosed Diseases Network \citep{ramoni2017}.
+The DNA was prepared from blood samples (with few exceptions) and sequenced via standard operation protocols for use as a Laboratory-Developed Test (LDT) in the HAIB CAP/CLIA lab.
+After sequencing, we followed GATK best practices \citep{depristo2011} to align to the hg19 reference with BWA-mem (CITE: XXX).
+Aligned sequences we processed via GATK for base quality score recalibration, indel realignment, and duplicate removal.
+Finally, SNV and indel variants were joint genotyped, again according to GATK best practices \citep{depristo2011}.
+The end result of this pipeline is a Variant Call Format (VCF) file used in the following sections.
+
+After VCF generation, clinical analysts followed various published recommendations (e.g. \citealp{worthey2017, roy2018}) to annotate and filter variants from proband samples.  
+%Specifically, we used a tool called CODICEM (CITE:XXX) to annotate, filter, and ultimately generate a report for each proband sample.  
+In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believe the variant to be directly related to the patient's phenotype.
+Secondary or incidental variant findings were excluded for our analysis.
+We required each sample included in our analysis to be a proband sample and to have at least one primary variant that was returned on the final clinical report.
+At the time of writing, this amounted to XXX primary variants from a total of XXX proband samples.
+
+TODO: modify above to reflect the need for phenotips also
 
 \subsection{Variant annotation}
-CODI does most of it; we use PyxisMap and HPOUtil for the rest
+For variant annotation, we used the tool CODICEM (CITE:XXX) that was developed at HudsonAlpha Institute for Biotechnology.
+In short, the software loads patient variants from a VCF and annotates the variants with over fifty annotations that a case analyst can use to interpret a patient's variant.
+These annotations include: variant level annotations such as CADD (CITE:XXX), conservation scores (CITE:XXX, phylop, phastcon, and GERP), and population frequencies (CITE:XXX gnomad); gene level annotations such as haploinsufficiency scores (CITE: XXX, HIS and GHIS), intolerance scores (CITE: XXX, RVIS), and disease associations (CITE: XXX, HGMD, ClinVar, OMIM); and transcript level annotations such as amino acid change scores (CITE: XXX, PolyPhen, SIFT, PROVEAN) and splice site impact scores (CITE: XXX, AdaBoost and RandomForest splice predictors).
+Additionally, if the variant has been previously curated through HGMD or ClinVar (CITE: XXX), those annotations are also made available to the analysts.
+
+The above annotations are all agnostic of the patient phenotype.
+In order to incorporate patient phenotype information, we used two methods to rank genes using the Human Phenotype Ontology (HPO) (CITE: XXX) terms provided by the clinical sites.
+We then annotated each variant with the scores from their corresponding gene.
+The first method uses base annotations provided by the HPO to calculate a simple cosine score (CITE: XXX) between the patient phenotype and each gene.
+This method tends to be more conservative because it relies on curated annotations from the HPO.
+The second method, a tool called PyxisMap, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator (CITE: XXX) and performs a Random-Walk with Restart (CITE:XXX) on the HPO ontology.
+This methods has the added benefit of incorporating fresh gene-phenotype connections from newer papers, but tends to be more error prone due to the imprecision of the text-mining.
 
@@
 
+\subsection{RENDERED DATA}
+\begin{table*}
+\centering
+\begin{tabular}{c|c|c|c|c||c|c|c|c}
+\input{data_rendered.tex}
+\end{tabular}
+\end{table*}
+
+more results go here
+
+\subsection{blah blah}
 Mathy stuff: Precision, recall, AUC, ROC, etc.
@@
 
-\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
+%CODICEM? - http://envisiongenomics.com/codicem-analysis-platform/
 
-\bibitem[Bag {\it et~al}., 2001]{Bag01}
-Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
+%GATK best practices
+\bibitem[DePristo {\it et~al}., 2011]{depristo2011} DePristo, Mark A., \textit{et~al.} "A framework for variation discovery and genotyping using next-generation DNA sequencing data." {\it Nature genetics} 43.5 (2011): 491.
 
-\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-in nigral dopaminergic neurnol cell: correlation with the known
-pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-Res.}, \textbf{110}(Suppl. 1), 76--84.
+%UDN paper
+\bibitem[Ramoni {\it et~al}., 2017]{ramoni2017} Ramoni, Rachel B.
+\textit{et~al.} "The undiagnosed diseases network: accelerating discovery about health and disease." {\it The American Journal of Human Genetics} 100.2 (2017): 185-192.
 
-\bibitem[Lehmann, 1986]{Leh86}
-Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
+%
+\bibitem[Roy {\it et~al}., 2018]{roy2018} Roy, Somak, \textit{et~al.} "Standards and guidelines for validating next-generation sequencing bioinformatics pipelines: a joint recommendation of the Association for Molecular Pathology and the College of American Pathologists." {\it The Journal of Molecular Diagnostics} 20.1 (2018): 4-27.
 
-\bibitem[Crenshaw and Jones, 2003]{Cre03}
-Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-cancer management: one tumor, one chip. \textit{Bioinformatics},
-doi:10.1093/bioinformatics/btn000.
-
-\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-???--???.
-
-\bibitem[Bardet, 1920]{Bar20}
-Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-polydactylie et retinite pigmentaire (contribution a l'etude des
-formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-institution, Paris, France.
+%Worthey WGS pipeline
+\bibitem[Worthey, 2017]{worthey2017} Worthey, Elizabeth A.
+"Analysis and Annotation of Whole-Genome or Whole-Exome Sequencing Derived Variants for Clinical Diagnosis." {\it Current protocols in human genetics} 95.1 (2017): 9-24.
 

### 2019-01-08 scripts/TestLearners.py

@@ -110,8 +110,7 @@ def getPyxisMapResults():
 
 def loadFormattedData():
     '''
-    Core test function.  This function is primarily about loading and formatting data prior to training/testing any 
-    classifiers.
+    This function is primarily about loading and formatting data prior to training/testing any classifiers.
     @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts)
         xFinal - an NxM matrix where N is the number of variants in our test/training set and M is the number of features; contains feature values
         yFinal - an N length array where N is the number of variants in our test/training set; 1 if the variant was reported
@@ -250,8 +249,15 @@ def loadFormattedData():
     for sl in sorted(caseData.keys()):
         #we only need to load those which actually have a return
         if len(caseData[sl]['primaries']) == 0:
-            print('Skipping '+sl+', no primaries')
+            #I don't think we care about seeing these at all
+            #print(sl, 'NO_PRIMARIES', sep=',')
             continue
+        
+        if len(caseData[sl]['hpoTerms']) == 0:
+            #we don't care to see these either, just incomplete data :(
+            #print(sl, 'NO_HPOTERMS', sep=',')
+            continue
+
         primarySet = set([p['variant'] for p in caseData[sl]['primaries']])
         primaryDict = {p['variant'] : p for p in caseData[sl]['primaries']}
         assert(len(primaryDict) == len(caseData[sl]['primaries']))
@@ -259,20 +265,18 @@ def loadFormattedData():
         #if there is no CODICEM dump, we obviously can't do anything either
         codiDump = CodiDumpUtil.loadCodiDump(sl, fieldsOnly, geneFieldsOnly, seqFieldsOnly, transLabelsOnly)
         if len(codiDump) == 0:
-            print('Skipping '+sl+', empty or absent CODICEM dump')
+            print(sl, 'NO_CODICEM_DUMP', sep=',')
             continue
         
         #finally, make sure we have a PyxisMap dump we can use
         pyxisFN = pyxisRootDir+'/'+sl+'_pyxis.json'
         if not os.path.exists(pyxisFN):
-            print('Skipping '+sl+', no PyxisMap dump available')
+            print(sl, 'NO_PYXISMAP_DUMP', sep=',')
             continue
         fp = open(pyxisFN, 'r')
[…]

### 2019-01-08 scripts/SummaryDBUtil.py
cleaning the file and print statements

@@ -86,41 +86,6 @@ def loadSummaryDatabase(altIDMap):
                         else:
                             #get the alternate ID if it exists, otherwise use the normal
                             ret[slid]['hpoTerms'].add(altIDMap.get(h, h))
-    
-    '''
-    #this was the original primary getter, but it changed, so I need to re-parse from a CSV
-    #second load the primaries
-    primarySheet = wb['Primary variants']
-    ALLOWED_PATHOGENICITY = set(['PATHOGENIC', 'LIKELY_PATHOGENIC', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE'])
-    primaryDict = {}
-    for i, row in enumerate(primarySheet):
-        if i == 0:
-            for j, c in enumerate(row):
-                if c.value != None:
-                    primaryDict[c.value] = j
-        else:
-            slid = row[primaryDict['SL#']].value
-            genes = row[primaryDict['GENE']].value
-            coords = row[primaryDict['GENOMIC COORDINATES']].value
-            pathLevel = row[primaryDict['REPORTED PATHOGENICITY']].value
-            
-            if (slid != None and genes != None and pathLevel != None and coords != None and
-                pathLevel in ALLOWED_PATHOGENICITY):
-                #strip out all g.
-                coordsMod = coords.replace('g.', '')
-                
-                if (slid not in ret):
-                    ret[slid] = {}
-                if ("primaries" not in ret[slid]):
-                    ret[slid]['primaries'] = []
-                    
-                print(slid, genes, pathLevel, coordsMod, sep='\t')
-                ret[slid]['primaries'].append({
-                    "genes" : list(set(genes.split(','))),
-                    "variant" : coordsMod,
-                    "path_level" : pathLevel
-                })
-    '''
 
     #second, load the primaries
     fpcsv = open('/Users/matt/githubProjects/VarSight/CODI_metadata/3Jan2019_UDN_Variants_all.csv', 'rt')
@@ -155,7 +120,7 @@ def loadSummaryDatabase(altIDMap):
                     pDict['path_level'] = revOrderPath[weakestPL]
                     break
             else:
-                print(slid, genes, pathLevel, coordsMod, sep='\t')
+                #print(slid, genes, pathLevel, coordsMod, sep='\t')
                 ret[slid]['primaries'].append({
                     "genes" : list(set(genes.split(','))),
                     "variant" : coordsMod,

### 2019-01-08 scripts/TestLearners.py
added argparse to make my life easier

@@ -2,6 +2,7 @@
 Primary analysis script for testing various learning methods
 '''
 
+import argparse as ap
 import datetime
 import json
 import matplotlib.pyplot as plt
@@ -34,7 +35,13 @@ from sklearn.model_selection import StratifiedKFold
 from sklearn.model_selection import train_test_split
 from sklearn.neural_network import MLPClassifier
 
+#CONSTANTS
 PYXIS_DATE='01042019'   
+
+#mode for analyzing the classifiers
+EXACT_MODE=0
+GRID_MODE=1
+RANDOM_MODE=2
 
 def getPyxisMapResults():
     '''
@@ -499,8 +506,9 @@ def loadFormattedData():
 
     return (xFinal, yFinal, featureLabels, startIndices, allRepDicts)
 
-def runClassifiers(values, classifications, featureLabels, startIndices, allRepDicts):
+def runClassifiers(args, values, classifications, featureLabels, startIndices, allRepDicts):
     '''
+    @param args - any arguments from the command line argparse can be accessed here
     @param values - a matrix with R rows and C columns, where there are "C" features
     @param classifications - an array of length R corresponding to the above values
     @param featureLabels - C length array containing labels (strings) for the features
@@ -587,10 +595,7 @@ def runClassifiers(values, classifications, featureLabels, startIndices, allRepD
     resultsDict['TEST_CASES'] = len(test_indices)-1
 
     #class_weight="balanced" is when there is a major imbalance between the number of true negatives and true positives
-    EXACT_MODE=0
-    GRID_MODE=1
-    RANDOM_MODE=2
-    currentMode=GRID_MODE
+    currentMode=args.training_mode
 
     #record the training method used
     resultsDict['TRAINING_MODE'] = ['EXACT_MODE', 'GRID_MODE', 'RANDOM_MODE'][currentMode]
@@ -965,16 +970,16 @@ def generateLaTeXResult(d):
     fp.write(rendered)
     fp.close()
 
-def runAnalysis():
+def runAnalysis(args):
     '''
     Core function for joining our pieces together
     '''
     resultJsonFN = '/Users/matt/githubProjects/VarSight/paper/results.json'
-    REGENERATE_DATA = True
+    REGENERATE_DATA = args.regenerate
 
     if REGENERATE_DATA or not os.path.exists(resultJsonFN):
         (xFinal, yFinal, featureLabels, startIndices, allRepDicts) = loadFormattedData()
[…]

### 2019-01-14 paper/data_template.tex
v0 of the paper, need to get edits and flesh out some TODOs

--- 2019-01-04 paper/data_template.tex
@@
-Classifier&TPR&FPR&AUC&CV10 Acc. \\ \hline
+\multirow{2}{*}{Ranking System}&\multicolumn{4}{c||}{Case Rank - Median (Mean)}&\multicolumn{4}{c}{Percentage in Top X Variants - X=(1, 10, 20)} \\ \cline{2-9}
+&All (n={{ TEST_COUNTS.OVERALL }})&
+VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
+LP (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Path. (n={{ TEST_COUNTS.PATHOGENIC }})&
+All (n={{ TEST_COUNTS.OVERALL }})&
+VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
+LP (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Path. (n={{ TEST_COUNTS.PATHOGENIC }}) \\ \hhline{=|=|=|=|=||=|=|=|=}
+{% for cs_label in CS_LABELS %}
+    {{ cs_label }}&
+    {% for LEVEL in ['OVERALL', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'LIKELY_PATHOGENIC', 'PATHOGENIC'] %}
+        {{ '%0.1f (%0.2f)' % (np.median(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]), np.mean(COMPARISON[cs_label].TEST_RANKINGS[LEVEL])) }}&
+    {% endfor %}
+    {% for LEVEL in ['OVERALL', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'LIKELY_PATHOGENIC', 'PATHOGENIC'] %}
+        {{ '%d, %d, %d' % 
+            (100.0*np.sum(np.array(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]) <= 1) / len(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]) <= 10) / len(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]) <= 20) / len(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]))
+        }}
+        {% if LEVEL != 'PATHOGENIC' %}
+            &
+        {% endif %}
+    {% endfor %}
+    \\
+{% endfor %}
+\hhline{=|=|=|=|=||=|=|=|=}
+
+Exomiser$^*$&
+{% for LEVEL in ['OVERALL', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'LIKELY_PATHOGENIC', 'PATHOGENIC'] %}
+    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER.TEST_RANKINGS[LEVEL])) }}&
+{% endfor %}
+{% for LEVEL in ['OVERALL', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'LIKELY_PATHOGENIC', 'PATHOGENIC'] %}
+    {{ '%d, %d, %d' % 
+        (100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER.TEST_RANKINGS[LEVEL]))
+    }}
+    {% if LEVEL != 'PATHOGENIC' %}
+        &
+    {% endif %}
+{% endfor %}
+\\ \hhline{=|=|=|=|=||=|=|=|=}
+
 {% for clf_label in CLF_LABELS %}
-    {{ clf_label }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_TPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].TEST_FPR_RATE }}&{{ '%0.2f' % CLASSIFIERS[clf_label].ROC_AUC }}&{{ '%0.2f+-%0.2f' % (CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[0], CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[1]) }} \\
+    {{ clf_label }}&
+    {% for LEVEL in ['OVERALL', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'LIKELY_PATHOGENIC', 'PATHOGENIC'] %}
+        {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL])) }}&
     {% endfor %}
+    {% for LEVEL in ['OVERALL', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'LIKELY_PATHOGENIC', 'PATHOGENIC'] %}
+        {{ '%d, %d, %d' % 
+            (100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 1) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 10) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 20) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]))
+        }}
+        {% if LEVEL != 'PATHOGENIC' %}
+            &
+        {% endif %}
+    {% endfor %}
+    \\
+{% endfor %}

### 2019-01-14 paper/main.tex
v0 of the paper, need to get edits and flesh out some TODOs

--- 2019-01-03 paper/main.tex
@@
 \documentclass{bioinfo}
+\usepackage{caption}
+\usepackage{graphicx}
+\usepackage{multirow}
+\usepackage{hhline}
 \copyrightyear{2015} \pubyear{2015}
@@
 
-\title[short Title]{This is a title}
-\author[Holt \textit{et~al}.]{James M. Holt\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
-\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
-$^{\text{\sf 2}}$Department, Institution, City, Post Code,
-Country.}
+\title[VarSight: Prioritizing Clinically Reported Variants]{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
+\author[Holt \textit{et~al}.]{James M.
+Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille Birch\,$^{\text{\sfb 1}}$, Donna Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa Wilk\,$^{\text{\sfb 1}}$, Julie Anderson\,$^{\text{\sfb 1}}$, Jeremy Harris\,$^{\text{\sfb 1}}$, Jacob Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A.
+Worthey\,$^{\text{\sfb 1}}$}
+\address{$^{\text{\sf 1}}$Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA % and \\
+%$^{\text{\sf 2}}$Department, Institution, City, Post Code, Country.}
+}
 
@@
 
-\abstract{\textbf{Motivation:} TODO.\\
-\textbf{Results:} TODO. \\
-\textbf{Availability:} TODO. \\
+\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the major goal is to identify one or more variants that cause disease.
+Typically, this is done through filtering and then prioritization of variants for manual curation.
+However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.
+Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability of critical importance.
+\\
+\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will be ultimately clinically reported and returned to a patient.
+To test the classifiers, we performed a retrospective study on variants that were clinically reported to patients in the Undiagnosed Diseases Network.
+We measured the classifier against another variant prioritization algorithm along with two single-measure controls to show how the classifiers tend to outperform the other methods.
+The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.
+\\
+\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
 \textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
@@
 \section{Introduction}
-WGS is currently being used as a molecular diagnostic tool for rare disease; blood draw + phenotypes are provided to analysis centers and these go through a standard alignment/variant calling pipeline
+Genome and exome sequencing are both currently being used as molecular diagnostic tools for patients with rare, undiagnosed diseases \citep{ramoni2017} (TODO: XXX: CITE MORE, CSER maybe?).
+Typically, these technologies are applied clinically following the workflows consisting of blood draw, sequencing, alignment, variant calling, variant annotation, variant filtering, and variant prioritization \citep{worthey2017, roy2018}.
+Then, clinical analysts usually perform the more manual processes of inspecting and then clinically reporting variants based on the known set of patient phenotypes.
 
-Because of size of data, variants are typically filtered down based on rare disease expectations; these variants are manually checked by two or more analysts and reported to clinical site
+In general, commonly used pipelines exist for the steps from sequencing through variant calling \citep{rehm2013, cornish2015}.
+Despite differences in performance, most of these standards ingest the same information to create a list of variants from sequencing data.
+In contrast, methods for variant annotation and/or variant filtering are quite diverse \citep{wang2010, hu2013, jager2014, desvignes2018}.
+These methods use a wide range of input sources including but not limited to population allele frequencies \citep{lek2016}, conservation scores \citep{cooper2005, siepel2006, petrovski2013}, haploinsufficiency scores \citep{huang2010, steinberg2015}, deleteriousness scores \citep{steinberg2015, rentzsch2018}, transcript impact scores \citep{kumar2009, choi2012, adzhubei2013, dong2014, jian2014} , and previously associated disease annotation \citep{stenson2003, hamosh2005, landrum2015}.
+Variant prioritization is also quite diverse with some methods relying on the variant annotations to order variants \citep{hu2013} and some relying on patient phenotype to order the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.
+There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.
 
-Manual curation is tedious and prone to error due to variant fatigue; thus tools that further reduce manual time on a case are paramount
+Given a prioritized list of variants, analysts manually inspect those variants and curate a list of variants to ultimately report to the ordering physician.
+Unfortunately, manual curation is time consuming and exhausting.
+An analyst must inspect each variant and the associated metadata while simultaneously maintaining a mental picture of the patient's phenotype, leading to what we colloquially refer to as ``variant fatigue".
+This variant fatigue means that variants at the end of the prioritized list are often regarded with less scrutiny and/or a less accurate patient model than those near the beginning.
+Additionally, if causative variants can be identified earlier due to a high rank, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.
+Finally, accurate prioritization is a step towards the ultimate goal of automatically identifying all variants that cause a patient's primary phenotypes.
 
-%\enlargethispage{12pt}
+One of the issues with previously published ranking methods is that they were primarily tested on simulated datasets with known, single-gene, pathogenic variants injected into real or simulated background genomic datasets.
+Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few noisy terms added or removed.
+
+In this paper, we focus on real patient data from the multi-site collaboration of the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.
+Patients accepted into the UDN are believed to have rare, undiagnosed diseases of genetic origin.
+Because the UDN is not focused on a single particular disease, the patient population has a diverse range of phenotypes represented.
+Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as phenotypes caused by old age).
+Finally, analysts have the discretion to send back variants with different levels of pathogenicity as specified by the ACMG guidelines \cite{richards2015}.
+The summation of these real-world problems means that prioritizing variants is challenging due to noisy phenotype inputs and variation in the pathogenicity of clinically reported outputs.
 
 \section{Approach}
+In this paper, we tested the application of classification algorithms for two purposes: 1) predicting whether a variant observed by an analyst would be clinically reported and 2) prioritizing all variants seen by clinical analysts.
+In particular, we focused our analyses on real patients with a diverse collection of rare, undiagnosed diseases that were admitted to the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.
+We limited our patients to those who received whole genome sequencing and received at least one primary variant (i.e.
+not secondary or incidental) on their clinical report.
+We extracted data directly from the same annotation and filtering tool used by the analysts in order to replicate their data view of each variant in a patient.
+Additionally, we incorporated phenotype information into the models using two scoring systems that are based on ranking genes by their association to a set of patient phenotypes.
+Finally, each variant was either labeled as ``returned" or ``not returned" depending on whether it was ultimately reported back to the clinical site.
 
-Here we test methods that take phenotype, gene, transcript, and variant level annotations from the filtered variant list and predict whether a variant will ultimately be reported back to the clinical site.
-
-The methods provide a boolean predictor and a ranking system.
+Given the above variant information, we split the data into training and testing sets for measuring the performance of classifiers to predict whether a variant would be clinically reported or not.
+We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.
+Each classifier calculated probabilities of a variant belong to the ``returned" class, allowing for both classification analysis and ranking of the variants by their calculated probabilities.
+After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant ranking perspective.
+All of the scripts to train classifiers, test classifiers, and format results are contained in the VarSight repository.
 
@@
 \section{Methods}
+\subsection{Data sources}
+All samples were selected from the cohort of genome sequencing samples originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB) Sequencing Center of the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.
+In short, the UDN accepts patients with rare, undiagnosed diseases that are believed to have a genetic origin.
+The UDN is not restricted to a particular disease, so there are a diverse set of diseases and phenotypes represented across the whole population.
+The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) some patients have multiple diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, and 3) some patients exhibit more or less phenotypes than are associated with the classic disease presentation.
+For more details on the UDN, refer to \citealp{ramoni2017}.
 
-\subsection{Variant annotation}
-CODI does most of it; we use PyxisMap and HPOUtil for the rest
+DNA for these UDN patients was prepared from blood samples (with few exceptions) and sequenced via standard operation protocols for use as a Laboratory-Developed Test (LDT) in the HAIB CAP/CLIA lab.
+The analyses presented in this paper are based on data that is or will be deposited in the dbGaP database under dbGaP accession phs001232.v1.p1 by the UDN.
+
+%dbGaP citing info: https://www.ncbi.nlm.nih.gov/books/NBK45306/
+%link to study for my sanity: https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs001232.v1.p1&phv=278114&phd=7085&pha=&pht=6078&phvf=5&phdf=&phaf=&phtf=&dssp=1&consent=&temp=1
+
+\subsection{Alignment and variant calling}
+After sequencing, we followed GATK best practices \citep{depristo2011} to align to the hg19 reference with BWA-mem \citep{li2013}.
+Aligned sequences we processed via GATK for base quality score recalibration, indel realignment, and duplicate removal.
+Finally, SNV and indel variants were joint genotyped, again according to GATK best practices \citep{depristo2011}.
+The end result of this pipeline is a Variant Call Format (VCF) file used in the following sections.
+
+\subsection{Variant annotation and filtering}
+After VCF generation, the clinical analysts followed various published recommendations (e.g.
+\citealp{worthey2017, roy2018}) to annotate and filter variants from proband samples.
+For variant annotation and filtering, we used the same tool that our analysts used during their initial analyses.
+The tool, Codicem \citep{envision2018}, loads patient variants from a VCF and annotates the variants with over fifty annotations that the analysts can use to interpret pathogenicity.
+These annotations include: variant level annotations such as CADD \citep{rentzsch2018}, conservation scores \citep{cooper2005, siepel2006}, and population frequencies \citep{lek2016}; gene level annotations such as haploinsufficiency scores \citep{huang2010, steinberg2015}, intolerance scores \citep{petrovski2013}, and disease associations \citep{stenson2003, hamosh2005, landrum2015}; and transcript level annotations such as protein change scores \citep{kumar2009, choi2012, adzhubei2013, dong2014} and splice site impact scores \citep{jian2014}.
+Additionally, if the variant has been previously curated in another patient through HGMD or ClinVar \citep{stenson2003, landrum2015}, those annotations are also made available to the analysts.
+
+Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.
+We used the latest version of their primary clinical filter to replicate the standard filtering process of the analysts.
+In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.
+For the specific details on the filter used, refer to TODO:XXX (I think the source code?).
+In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).
+
+\subsection{Phenotype annotation}
+The Codicem annotations are all agnostic of the patient phenotype.
+As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases and phenotypes tied to UDN patients.
+In order to incorporate patient phenotype information, we used two distinct methods to rank genes based on the Human Phenotype Ontology (HPO) \citep{kohler2018}.
+We then annotated each variant with the best scores from their corresponding gene(s).
+
+The first method uses base annotations provided by the HPO to calculate a simple cosine score \citep{kohler2017} between the patient's phenotypes and each gene.
+This method tends to be more conservative because it relies solely on curated annotations from the HPO.
+The second method, a tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.
+The PyxisMap method has the added benefit of incorporating fresh gene-phenotype connections from newer papers, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.
+Each method generates a single numerical feature that is used in the following analyses.
+
+\subsection{Patient selection}
+In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believed the variant to be directly related to the patient's phenotype.
+Note that secondary and/or incidental findings are specifically not included in this list.
+The analysts assigned each primary variant a rank from variant of uncertain significance (VUS), likely pathogenic, or pathogenic based on the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.
+
+We required the following for each proband sample included in our analyses: 1) at least one clinically reported primary variant that came through the primary clinical filter (i.e.
+it was not found through some other targeted search) and 2) a set of phenotypes annotated using the Human Phenotype Ontology \citep{kohler2018}.
+At the time of writing, this amounted to 378 primary, reported variants spanning a total of 237 proband samples.
 
 \subsection{Data cleaning}
-Numerical data is given a number with out-of-bounds defaults
+For the purposes of classification, all annotations needed to be cleaned and stored as numerical features.
+For numerical annotations, we simply copied the annotation over as a single value feature.
+Default values were assigned to missing annotations that were outside the expected value range.
+Additionally, these values were always on the less impactful side of the spectrum (e.g.
+a default conservation score would err on the side of not being conserved).
+The one exception to this rule was for variant allele frequencies where a variant absent from the database was considered to have an allele frequency of 0.0.
 
-Categorical data is XXX (what do we decide to do with it?)
+For categorical data, we used a two step approach to cleaning the data: bin-count encoding and principal component analysis.
+First, we chose to use a bin-count because there are many categories where multiple categorical labels may be present at different quantities.
+For example, a single ClinVar variant may have multiple entries where different sites have selected different levels of pathogenicity.
+In this situation, we desired to capture not only the categorical label as a feature, but also the number of times that label occurred in the annotations.
+Second, we found that the bin-count encoding tended to create many extra features (one per category per annotation type) that were ignored and/or diluted the useful features from the pool of features.
+To reduce this dilution, we used principal component analysis \citep{jolliffe2011} to reduce the dimensions of each category and stored at most two features per categorical feature.
+For specifics on variance explained by the top two PCA components per category, please see Supplementary document XXX (TODO: are we going to make this or no?)
 
 \subsection{Model training and tuning}
+As noted earlier, there are generally hundreds of variants per proband that make it through the filter, but only a few are ever clinically reported.
+As a result, there is a major imbalance in the number of true positives (variants clinically reported) and true negatives (variants seen, but not clinically reported).
+Across all 237 proband samples, there were a total of 378 clinically reported variants and another 87819 variants that were seen but not reported.
+We split the data into training and test sets on a per-proband basis with the primary goal of roughly balancing the total number of true positives in each set.
+Additionally, the cases were assigned to a particular set by chronological order of analysis in order to reduce any chronological biases that may be introduced by expanding scientific knowledge (i.e.
+there are roughly equal proportions of ``early" or ``late" proband samples from the UDN in each set).
+In the training set, there were a total of 189 returned variants and 44593 not returned variants spanning 120 different probands.
+In the test set, there were a total of 189 returned variants and 43226 not returned variants spanning 117 different probands.
+In our results, the returned test variants are further stratified in their reported levels of pathogenicity.
 
-We tested a lot of models and tuned them
+We then selected four readily available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.
+For each model, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.
+For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e.
+a weighted accuracy score where inputs are weighted by their inverse class frequency).
+For each model type, we saved the hyperparameters and model with the best average balanced accuracy score.
+These four tuned models were then tested against the unseen set of test proband cases.
 
@@
 \section{Discussion}
-Results go here
+\subsection{Classifier Statistics}
+\begin{figure*}
+\centerline{
+\includegraphics[width=0.45\textwidth]{codi_rf_roc.png}
+\includegraphics[width=0.45\textwidth]{codi_rf_pr.png}
+}
+\caption{Receiver operator curves.
+This figure shows the receiver operator curves for the test data for each single-measure statistic and for each trained classifier.
+On the left, we show the false positive rate against the true positive rate.
+On the right, we show the recall against the precision.
+Area under the receiver-operator curve (AUROC) is reported beside each method in the legend.
+In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly.
+LogisticRegression tends to perform slightly worse overall, but still better than either single-measure statistic.}
+\label{fig:roc}
+\end{figure*}
 
-Mathy stuff: Precision, recall, AUC, ROC, etc.
+For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.
+We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data.
+The values for each model are shown in Table \ref{tab:classifier_results}.
+Additionally, receiver operator curves for FPR v.
+TPR and recall v.
+precision were created for the four models.
+We compared this to two single-measure statistics (CADD Scaled and HPO-cosine) in Figure \ref{fig:roc}.
 
-Practical stuff: in a case, this pushed reported variants to rank X out of Y (or something like this)
+From these metrics, the two random forest models and the EasyEnsembleClassifier seem to perform similarly whereas LogisticRegression performs slightly worse across the board.
+All four classifiers have higher AUROC than either single-measure statistic.
+However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was only 0.2109).
+This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.
+
+\begin{table}[!t]
+\processtable{Classifier performance statistics.
+For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy, true positive rate, false positive rate, and area under the receiver operator curve.
+In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.
+\label{tab:classifier_results}}{
+\begin{tabular}{c|c|c|c|c}
+\hline
+\input{classifier_rendered.tex}
+\hline
+\end{tabular}}{}
+\end{table}
+
+\subsection{Ranking Statistics}
+\begin{table*}[!t]
+\centering
+\processtable{Ranking performance statistics.
+This table shows the ranking performance statistics for all methods evaluated on our test set.
+CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.
+Exomiser is an external tool that only reported ranks for a subset of the filtered variants.
+The bottom four rows are the tuned, binary classification methods tested in this paper.
+The ``Case Rank" columns show the median and mean ranks for all variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).
+The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.
+All values were generated using only the test data that was unseen during training.
+\label{tab:performance}}{
+\begin{tabular}{c|c|c|c|c||c|c|c|c}
+\hline
+\input{data_rendered.tex}
+\hline
+\end{tabular}}{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
+\end{table*}
+
+In addition to the model performance statistics, we also quantified the performance of each classifier as a ranking system.
+For each proband, we calculated the probability of each class (reported or not reported) for each variant and ordered them from highest to lowest probability of being reported.
+We then calculated median and mean rank statistics for the reported variants.
+Additionally, we quantified the percentage of variants that were ranked in the top 1, 10, and 20 variants in each case.
+While the models were trained as a binary classification system, we broke down the results further to demonstrate differences between variants that were clinically reported as a variant of uncertain significance (VUS), likely pathogenic, and pathogenic.
+
+For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).
+We used followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.
+For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.
+We then parsed the output JSON file created by Exomiser into a ranked order of variants.
+Of note, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 20-50 variants from the pre-filtered VCF.
+For primary variants that were unranked by Exomiser, we conservatively assumed that the Exomiser ranked those variants as just outside the range of what was returned (e.g.
+if Exomiser ranked 15 variants, we assumed a missing variant was ranked 16th).
+
+Finally, we added two control scores for comparison: CADD scaled and HPO-cosine.
+These scores were inputs to each classifier, but also represent some common ways variants one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).
+The results for the two control scores, Exomiser, and all four classifiers are shown in Table \ref{tab:performance}.
+
+In the overall data, all four classifiers outperform the single-measure statistics and Exomiser across the board, with the one exception being Exomiser's mean (this measure is likely heavily biased due to our conservative ranking of unranked variants, harsher penalties for unranked variants would drastically change this result).
+As expected, all methods perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.
+Of the methods tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.
+
+\subsection{Random Forest Feature Importances}
+After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
+Both random forest models calculate a feature importance array after training that sum to 1.0.
+By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.
+Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ as reported by each classifier.
+
+Interestingly, the two strongest individual features by far were the phenotype-based measures (HPO-cosine and PyxisMap) making up $\sim$33-40\% of the feature importance.
+These were followed by CADD Scaled ($\sim$7-10\%) and three conservation scores ($\sim$14\%).
+GnomAD-related fields ($\sim$8\%), GERP rsScore ($\sim$2-3\%), and HGMD-related fields ($\sim$4\%) made up the rest of features with higher than expected importance.
+These results would suggest that phenotype-based metrics, deleteriousness scores, and conservation scores are of very high importance for predicting if a variant will be clinically returned to a rare disease patient.
+
+\begin{table}[!t]
+\processtable{Random forest feature importances.
+This table shows the top feature importances reported by the two random forest algorithms we tested.
+We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance.
+\label{tab:features}}{
+\begin{tabular}{l|c|c}
+\hline
+\input{features_rendered.tex}
+\hline
+\end{tabular}}{}
+\end{table}
 
 \section{Conclusion}
+We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report.
+We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process used for the Undiagnosed Diseases Network (UDN).
+From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.
+However, when evaluated as a ranking system, all four methods out-performed single-measure ranking systems and Exomiser.
+We consider the best classifier to be the EasyEnsembleClassifier from {\it imblearn} that had a median rank of 7.0 for all reported variants while ranking 73\% in the top 20 for the case.
+For ``Pathogenic" variants, the median rank was 2.0 and 97\% of those variants were ranked in the top 20 for the case.
+While not perfect, the ability for the algorithms to push reported variants to very early in the list is quite promising.
 
-We did a thing
+We expect these classification algorithms could be improved or refined in a variety of ways.
+First, adding new features and/or removing unused features could lead to improvements in the algorithm.
+The top two most important features were both phenotype related suggesting that a highly accurate phenotype scoring system would greatly benefit these algorithms.
+Many features had a low importance, so pruning them may improve the overall results.
+Additionally, some of the features represent data that is not freely available to the research community, so pruning or replacing those features with publicly accessible sources would likely influence the results.
+Second, there may be a better classification algorithm for this type of data.
+The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may improve the result.
+
+Ultimately, we believe the trained classifiers in VarSight are a significant step forward in handling real-world data for rare, undiagnosed disease patients.
+Despite the variability in phenotype and allowance of variants of uncertain significance in reporting, the models trained in VarSight still ranked reported variants highly, especially variants labeled as ``Pathogenic".
+Going forward, we expect the standard for variant ranking systems will behave similar to those in VarSight, fully integrating all available phenotype and variant annotations into one model.
 
 \section*{Acknowledgements}
-TODO
-
-Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text
-text text text\vspace*{-12pt}
+We are grateful for the participation of patients and family members and all of their referring clinicians.
+We would like to acknowledge all of the teams within the UDN including the coordinating center and all of the clinical sites working hard to provide definitive diagnoses for UDN patients.
+We would also like to acknowledge the UDN WGS core headed by Dr.
+Shawn Levy.
 
 \section*{Funding}
-TODO
-
-This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}
+This work was supported in part by the Intramural Research Program of the National Human Genome Research Institute and the NIH Common Fund through the Office of Strategic Coordination and Office of the NIH Director.
+Research reported in this manuscript was supported by the NIH Common Fund through the Office of Strategic Coordination and Office of the NIH Director under award numbers U01HG007530, U01HG007674, U01HG007703, U01HG007709, U01HG007672, U01HG007690, U01HG007708, U01HG007942, U01HG007943, U54NS093793, and U01TR001395.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the NIH.
 
@@
 
-\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
+%PolyPhen
+\bibitem[Adzhubei {\it et~al}., 2013]{adzhubei2013} Adzhubei, Ivan, Daniel M.
+Jordan, and Shamil R.
+Sunyaev.
+"Predicting functional effect of human missense mutations using PolyPhen?2." {\it Current protocols in human genetics} 76.1 (2013): 7-20.
 
-\bibitem[Bag {\it et~al}., 2001]{Bag01}
-Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
+%PROVEAN - variant-transcript prediction
+\bibitem[Choi, 2012]{choi2012} Choi, Yongwook.
+"A fast computation of pairwise sequence alignment scores between a protein and a set of single-locus variants of another protein." {\it Proceedings of the ACM Conference on Bioinformatics, Computational Biology and Biomedicine}.
+ACM, 2012.
 
-\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-in nigral dopaminergic neurnol cell: correlation with the known
-pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-Res.}, \textbf{110}(Suppl. 1), 76--84.
+%GERP - conservation
+\bibitem[Cooper {\it et~al}., 2005]{cooper2005} Cooper, Gregory M., \textit{et al.} "Distribution and intensity of constraint in mammalian genomic sequence." {\it Genome research} 15.7 (2005): 901-913.
 
-\bibitem[Lehmann, 1986]{Leh86}
-Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
+%variant calling pipeline comparisons
+\bibitem[Cornish {\it et~al}., 2015]{cornish2015} Cornish, Adam, and Chittibabu Guda.
+"A comparison of variant calling pipelines using genome in a bottle as a reference." {\it BioMed research international 2015} (2015).
 
-\bibitem[Crenshaw and Jones, 2003]{Cre03}
-Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-cancer management: one tumor, one chip. \textit{Bioinformatics},
-doi:10.1093/bioinformatics/btn000.
+%GATK best practices
+\bibitem[DePristo {\it et~al}., 2011]{depristo2011} DePristo, Mark A., \textit{et~al.} "A framework for variation discovery and genotyping using next-generation DNA sequencing data." {\it Nature genetics} 43.5 (2011): 491.
 
-\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-???--???.
+%VarAFT - variant annotator
+\bibitem[Desvignes {\it et~al}., 2018]{desvignes2018} Desvignes, Jean-Pierre, \textit{et~al.} "VarAFT: a variant annotation and filtration system for human next generation sequencing data." {\it Nucleic acids research} (2018).
 
-\bibitem[Bardet, 1920]{Bar20}
-Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-polydactylie et retinite pigmentaire (contribution a l'etude des
-formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-institution, Paris, France.
+%MetaSVM - variant-transcript annotation
+\bibitem[Dong {\it et~al}., 2014]{dong2014} Dong, Chengliang, \textit{et~al.} "Comparison and integration of deleteriousness prediction methods for nonsynonymous SNVs in whole exome sequencing studies." {\it Human molecular genetics} 24.8 (2014): 2125-2137.
+
+%Codicem citation?
+\bibitem[Envision, 2018]{envision2018} Envision Genomics.
+"Codicem Analysis Platform." {\it Envision Genomics}.
+URL: \href{http://envisiongenomics.com/codicem-analysis-platform/}{http://envisiongenomics.com/codicem-analysis-platform/}.
+
+%OMIM
+\bibitem[Hamosh {\it et~al}., 2005]{hamosh2005} Hamosh, Ada, \textit{et~al.} "Online Mendelian Inheritance in Man (OMIM), a knowledgebase of human genes and genetic disorders." {\it Nucleic acids research} 33.suppl$\_$1 (2005): D514-D517.
+
+%VAAST - variant annotation & prioritization
+\bibitem[Hu {\it et~al}., 2013]{hu2013} Hu, Hao, \textit{et~al.} "VAAST 2.0: improved variant classification and disease-gene identification using a conservation-controlled amino acid substitution matrix." {\it Genetic epidemiology} 37.6 (2013): 622-634.
+
+%HIS - haploinsufficiency 
+\bibitem[Huang {\it et~al}., 2010]{huang2010} Huang, Ni, \textit{et~al.} "Characterising and predicting haploinsufficiency in the human genome." {\it PLoS genetics} 6.10 (2010): e1001154.
+
+%JANNOVAR - variant annotator, mostly for transcripts
+\bibitem[Jäger {\it et~al}., 2014]{jager2014} Jäger, Marten, \textit{et~al.} "Jannovar: A Java Library for Exome Annotation." {\it Human mutation} 35.5 (2014): 548-555.
+
+%AdaBoost and RandomForest predictions
+\bibitem[Jian {\it et~al}., 2014]{jian2014} Jian, Xueqiu, Eric Boerwinkle, and Xiaoming Liu.
+"In silico prediction of splice-altering single nucleotide variants in the human genome." {\it Nucleic acids research} 42.22 (2014): 13534-13544.
+
+%highly cite PCA from book
+\bibitem[Jolliffe, 2011]{jolliffe2011} Jolliffe, Ian. "Principal component analysis." {\it International encyclopedia of statistical science}. Springer, Berlin, Heidelberg, 2011. 1094-1096.
+
+%phenomizer - phenotype ranker
+\bibitem[Köhler {\it et~al}., 2009]{kohler2009} Köhler, Sebastian, \textit{et~al.} "Clinical diagnostics in human genetics with semantic similarity searches in ontologies." {\it The American Journal of Human Genetics} 85.4 (2009): 457-464.
+
+%cosine & jaccard scores
+\bibitem[Köhler, 2017]{kohler2017} Koehler, Sebastian. "Ontology-based similarity calculations with an improved annotation model." {\it bioRxiv} (2017): 199554.
+
+%HPO paper to cite
+\bibitem[Köhler {\it et~al}., 2018]]{kohler2018} Köhler, Sebastian, \textit{et~al.} "Expansion of the Human Phenotype Ontology (HPO) knowledge base and resources." {\it Nucleic acids research} (2018).
+
+%SIFT - variant-transcript prediction
+\bibitem[Kumar {\it et~al}., 2009]{kumar2009} Kumar, Prateek, Steven Henikoff, and Pauline C.
+Ng.
+"Predicting the effects of coding non-synonymous variants on protein function using the SIFT algorithm." {\it Nature protocols} 4.7 (2009): 1073.
+
+%ClinVar - latest pub.
+\bibitem[Landrum {\it et~al}., 2015]{landrum2015} Landrum, Melissa J., \textit{et~al.} "ClinVar: public archive of interpretations of clinically relevant variants." {\it Nucleic acids research} 44.D1 (2015): D862-D868.
+
+%ExAC & gnomad paper to cite
+\bibitem[Lek {\it et~al}., 2016]{lek2016} Lek, Monkol, \textit{et~al.} "Analysis of protein-coding genetic variation in 60,706 humans." {\it Nature} 536.7616 (2016): 285.
+
+%imblearn
+\bibitem[Lemaître {\it et~al}., 2017]{lemaitre2017} Lemaître, Guillaume, Fernando Nogueira, and Christos K.
+Aridas.
+"Imbalanced-learn: A python toolbox to tackle the curse of imbalanced datasets in machine learning." {\it The Journal of Machine Learning Research} 18.1 (2017): 559-563.
+
+%BWA-MEM
+\bibitem[Li, 2013]{li2013} Li, Heng. "Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM." {\it arXiv} preprint arXiv:1303.3997 (2013).
+
+%RWR algorithm
+\bibitem[Page {\it et~al}., 1999]{page1999} Page, Lawrence, \textit{et~al.} "The PageRank citation ranking: Bringing order to the web." {\it Stanford InfoLab}, 1999.
+
+%sklearn
+\bibitem[Pedregosa {\it et~al}., 2011]{pedregosa2011} Pedregosa, Fabian, \textit{et~al.} "Scikit-learn: Machine learning in Python." {\it Journal of machine learning research} 12.Oct (2011): 2825-2830.
+
+%RVIS - intolerance score
+\bibitem[Petrovski {\it et~al}., 2013]{petrovski2013} Petrovski, Slavé, \textit{et~al.} "Genic intolerance to functional variation and the interpretation of personal genomes." {\it PLoS genetics} 9.8 (2013): e1003709.
+
+%UDN paper
+\bibitem[Ramoni {\it et~al}., 2017]{ramoni2017} Ramoni, Rachel B.
+\textit{et~al.} "The undiagnosed diseases network: accelerating discovery about health and disease." {\it The American Journal of Human Genetics} 100.2 (2017): 185-192.
+
+%phenotype based prioritization
+\bibitem[Rao {\it et~al}., 2018]{rao2018} Rao, Aditya, \textit{et~al.} "Phenotype-driven gene prioritization for rare diseases using graph convolution on heterogeneous networks." {\it BMC medical genomics} 11.1 (2018): 57.
+
+%ACMG clinical standards paper
+\bibitem[Rehm {\it et~al}., 2013]{rehm2013} Rehm, Heidi L., \textit{et~al.} "ACMG clinical laboratory standards for next-generation sequencing." {\it Genetics in medicine} 15.9 (2013): 733.
+
+%latest CADD paper
+\bibitem[Rentzsch {\it et~al}., 2018]{rentzsch2018} Rentzsch, Philipp, \textit{et~al.} "CADD: predicting the deleteriousness of variants throughout the human genome." {\it Nucleic acids research} (2018).
+
+%ACMG guidelines
+\bibitem[Richards {\it et~al}., 2015]{richards2015} Richards, Sue, \textit{et~al.} "Standards and guidelines for the interpretation of sequence variants: a joint consensus recommendation of the American College of Medical Genetics and Genomics and the Association for Molecular Pathology." {\it Genetics in medicine} 17.5 (2015): 405.
+
+%
+\bibitem[Roy {\it et~al}., 2018]{roy2018} Roy, Somak, \textit{et~al.} "Standards and guidelines for validating next-generation sequencing bioinformatics pipelines: a joint recommendation of the Association for Molecular Pathology and the College of American Pathologists." {\it The Journal of Molecular Diagnostics} 20.1 (2018): 4-27.
+
+%PhastCons and PhyloP - conservation scores
+\bibitem[Siepel {\it et~al}., 2006]{siepel2006} Siepel, Adam, Katherine S.
+Pollard, and David Haussler.
+"New methods for detecting lineage-specific selection." {\it Annual International Conference on Research in Computational Molecular Biology}.
+Springer, Berlin, Heidelberg, 2006.
+
+%Phevor - variant & phenotype ranker; removed reference in favor of review paper
+\bibitem[Singleton {\it et~al}., 2014]{singleton2014} Singleton, Marc V., \textit{et~al.} "Phevor combines multiple biomedical ontologies for accurate identification of disease-causing alleles in single individuals and small nuclear families." {\it The American Journal of Human Genetics} 94.4 (2014): 599-610.
+
+%Exomiser - phenotype & variant ranker; removed reference in favor of review paper
+\bibitem[Smedley {\it et~al}., 2015a]{smedley2015a}Smedley, Damian, \textit{et~al.} "Next-generation diagnostics and disease-gene discovery with the Exomiser." {\it Nature protocols} 10.12 (2015): 2004.
+
+%review of phenotype & variant rankers
+\bibitem[Smedley {\it et~al}., 2015b]{smedley2015b} Smedley, Damian, and Peter N.
+Robinson.
+"Phenotype-driven strategies for exome prioritization of human Mendelian disease genes." {\it Genome medicine} 7.1 (2015): 81.
+
+%GHIS - haploinsuff. scores
+\bibitem[Steinberg {\it et~al}., 2015]{steinberg2015} Steinberg, Julia, \textit{et~al.} "Haploinsufficiency predictions without study bias." {\it Nucleic acids research} 43.15 (2015): e101-e101.
+
+%HGMD - 2003, this is what they ask to cite though
+\bibitem[Stenson {\it et~al}., 2003]{stenson2003} Stenson, Peter D., \textit{et~al.} "Human gene mutation database (HGMD®): 2003 update." {\it Human mutation} 21.6 (2003): 577-581.
+
+%ANNOVAR - variant annotation
+\bibitem[Wang {\it et~al}., 2010]{wang2010} Wang, Kai, Mingyao Li, and Hakon Hakonarson.
+"ANNOVAR: functional annotation of genetic variants from high-throughput sequencing data." {\it Nucleic acids research} 38.16 (2010): e164-e164.
+
+%PubTator
+\bibitem[Wei {\it et~al}., 2013]]{wei2013} Wei, Chih-Hsuan, Hung-Yu Kao, and Zhiyong Lu.
+"PubTator: a web-based text mining tool for assisting biocuration." {\it Nucleic acids research} 41.W1 (2013): W518-W522.
+
+%PyxisMap
+\bibitem[Wilk {\it et~al}., 2018]{wilk2018} Wilk, Brandon, James M.
+Holt, and Elizabeth A.
+Worthey.
+"PyxisMap." {\it HudsonAlpha Institute for Biotechnology.}  URL: \href{https://github.com/HudsonAlpha/LayeredGraph}{https://github.com/HudsonAlpha/LayeredGraph}.
+
+%Worthey WGS pipeline
+\bibitem[Worthey, 2017]{worthey2017} Worthey, Elizabeth A.
+"Analysis and Annotation of Whole-Genome or Whole-Exome Sequencing Derived Variants for Clinical Diagnosis." {\it Current protocols in human genetics} 95.1 (2017): 9-24.
+
+%Phenolyzer - phenotype based ranking
+\bibitem[Yang {\it et~al}., 2015]{yang2015} Yang, Hui, Peter N.
+Robinson, and Kai Wang.
+"Phenolyzer: phenotype-based prioritization of candidate genes for human diseases." {\it Nature methods} 12.9 (2015): 841.
+
+%PhenIX
+\bibitem[Zemojtel {\it et~al}., 2014]{zemojtel2014} Zemojtel, Tomasz, \textit{et~al.} "Effective diagnosis of genetic disease by computational phenotype analysis of the disease-associated genome." {\it Science translational medicine} 6.252 (2014): 252ra123-252ra123.
 

### 2019-01-14 scripts/ExomiserUtil.py

@@ -1,0 +1,57 @@
+'''
+This file should contain helper functions to parse and load Exomiser ranking files
+'''
+
+import json
+import os
+
+def loadAllExomiserRanks(sl):
+    '''
+    Loads a .json file from Exomiser
+    @param sl - the SL number to load
+    '''
+    fn = '/Users/matt/githubProjects/VarSight/exomiser_results/'+sl+'-hiphive-genome-PASS_ONLY.json'
+    fp = open(fn, 'rt')
+    j = json.load(fp)
+    fp.close()
+
+    #the file is a little weird, ranks are done by gene first, and the variants are a sub-group inside the gene
+    orderedRet = []
+    for gDict in j:
+        #inside the gene dictionary, we want variant evaluations
+        #print(gDict['geneSymbol'])
+        for vEval in gDict['variantEvaluations']:
+            chrom = vEval['chromosome']
+            pos = vEval['position']
+            ref = vEval['ref']
+            alt = vEval['alt']
+            #these are not the normal gDots we see typically; changing to chrom pos ref alt and figuring it out
+            #gDot = vEval['transcriptAnnotations'][0]['hgvsGenomic']
+            #print('\t', chrom, pos, ref, alt)
+            
+            #make sure there are no doubles for overlapping genes
+            if (chrom, pos, ref, alt) not in orderedRet:
+                orderedRet.append((str(chrom), str(pos), ref, alt))
+    
+    return orderedRet
+
+def getTargetRanks(sl, foundPrimaries):
+    '''
+    Simple utility to load the SL file and find the rank of each primary that Exomiser actually ranked; None for any it didn't
+    @param sl - the identifier for the patient
+    @param foundPrimaries - list of primaries in tuple form (chrom, position, ref allele, alt allele)
+    @return - a list of their 0-based rank from the Exomiser JSON output
+    '''
+    fullRanks = loadAllExomiserRanks(sl)
+    ret = []
+    for fp in foundPrimaries:
+        try:
+            ind = fullRanks.index(fp)
+        except:
+            ind = len(fullRanks)
+        ret.append(ind)
+    return ret
+
+if __name__=='__main__':
+    #for testing purposes
+    pass

### 2019-01-14 scripts/TestLearners.py

@@ -11,6 +11,7 @@ import os
 import re
 
 import CodiDumpUtil
+import ExomiserUtil
 import HPOUtil
 import OntologyUtil
 import PyxisMapUtil
@@ -193,7 +194,8 @@ def loadFormattedData():
         ("percent reads", -1.0),
         ("total depth", -1.0)
     ]
-    seqFieldsOnly = set([l for l, d in seqValues])
+    
+    seqFieldsOnly = set([l for l, d in seqValues]) | set(['chromosome', 'position', 'ref allele', 'alt allele'])
     
     #numerical variant-transcript information
     #"ADA Boost Splice Prediction" - looks like 0-1 with "NA" in there; TODO: what is good/bad?
@@ -253,6 +255,8 @@ def loadFormattedData():
     #this will track the breaks between cases
     startIndices = [0]
 
+    exomiserRanks = []
+
     for sl in sorted(caseData.keys()):
         #we only need to load those which actually have a return
         if len(caseData[sl]['primaries']) == 0:
@@ -311,6 +315,7 @@ def loadFormattedData():
         
         #print('Primary set: ', primarySet)
         foundPrimaries = set([])
+        fpCPRA = []
         
         #go through each variant now and format the inputs to be seen by a classifier
         for variant in codiDump:
@@ -322,6 +327,9 @@ def loadFormattedData():
             isPrimary = (1.0 if (vName in primarySet) else 0.0)
             if isPrimary:
                 foundPrimaries.add(vName)
+                
+                #fpCPRA is the same order as the repDicts (needed for later)
+                fpCPRA.append((str(variant['chromosome']), str(variant['position']), variant['ref allele'], variant['alt allele']))
                 repDicts.append(primaryDict[vName])
             else:
                 repDicts.append(None)
@@ -430,6 +438,11 @@ def loadFormattedData():
                 allCatBreak[k] += catBreak[k]
             allClassifications.append(classifications)
             allRepDicts += repDicts
+
+            #if we are using this case, load info from exomiser for comparison later
+            exRanks = ExomiserUtil.getTargetRanks(sl, fpCPRA)
+            exomiserRanks.append(exRanks)
+
         else:
             print(sl, 'ALL_MISSING', primarySet, foundPrimaries, sep=',')
         
@@ -441,7 +454,7 @@ def loadFormattedData():
[…]

### 2019-01-15 scripts/SummaryDBUtil.py
added option to only use path variants, doesn't work with paper output

@@ -5,9 +5,10 @@ This file should contain helper functions to load the UDN summary DB information
 import csv
 import openpyxl
 
-def loadSummaryDatabase(altIDMap):
+def loadSummaryDatabase(altIDMap, path_only):
     '''
     @param altIDMap - an alternate ID map for the HPO terms; dict of HPO term and the term to replace it with
+    @param path_only - if True, only load pathogenic and likely pathogenic variants as true positive
     @return - a JSON dictionary with the following structure:
     {
         "SL##" : {
@@ -91,7 +92,11 @@ def loadSummaryDatabase(altIDMap):
     fpcsv = open('/Users/matt/githubProjects/VarSight/CODI_metadata/3Jan2019_UDN_Variants_all.csv', 'rt')
     primarySheet = csv.DictReader(fpcsv)
     revOrderPath = ['PATHOGENIC', 'LIKELY_PATHOGENIC', 'VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'BENIGN']
-    ALLOWED_PATHOGENICITY = set(revOrderPath[0:3])
+    if path_only:
+        ALLOWED_PATHOGENICITY = set(revOrderPath[0:2])
+    else:
+        ALLOWED_PATHOGENICITY = set(revOrderPath[0:3])
+        
     for row in primarySheet:
         slid = row['Sample']
         genes = row['Gene']

### 2019-01-15 scripts/TestLearners.py

@@ -116,9 +116,10 @@ def getPyxisMapResults():
                 else:
                     print('Failed to retrieve results for '+sl+' from HPOUtil')
 
-def loadFormattedData():
+def loadFormattedData(args):
     '''
     This function is primarily about loading and formatting data prior to training/testing any classifiers.
+    @param args - arguments from the command line
     @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts)
         xFinal - an NxM matrix where N is the number of variants in our test/training set and M is the number of features; contains feature values
         yFinal - an N length array where N is the number of variants in our test/training set; 1 if the variant was reported
@@ -242,8 +243,8 @@ def loadFormattedData():
     ontFN = '/Users/matt/githubProjects/LayeredGraph/HPO_graph_data/hp.obo'
     nodes, edges, parents, altIDMap = OntologyUtil.loadGraphStructure(ontFN)
     
-    #1 - load the summary database
-    caseData = SummaryDBUtil.loadSummaryDatabase(altIDMap)
+    #1 - load the summary database 
+    caseData = SummaryDBUtil.loadSummaryDatabase(altIDMap, args.path_only)
     
     #2 - load each CODI dump that HAS a result from the summary database and reformat the data
     allValues = []
@@ -544,6 +545,17 @@ def runClassifiers(args, values, classifications, featureLabels, startIndices, a
     CASE_BASED_SPLIT = True
     TEST_SIZE = 0.5
 
+    pDict = {
+        'VARIANT_OF_UNCERTAIN_SIGNIFICANCE' : 3,
+        'LIKELY_PATHOGENIC' : 4,
+        'PATHOGENIC' : 5
+    }
+    if args.path_only:
+        pList = ['LIKELY_PATHOGENIC', 'PATHOGENIC']
+    else:
+        pList = ['VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 'LIKELY_PATHOGENIC', 'PATHOGENIC']
+    
+
     if CASE_BASED_SPLIT:
         #we decided to split the train/test by case, going to need some custom stuff here
         #we want to make sure ~25% of the true positives are in the test set and the remaining in the training set
@@ -751,14 +763,7 @@ def runClassifiers(args, values, classifications, featureLabels, startIndices, a
         if CASE_BASED_SPLIT:
[…]

### 2019-01-15 paper/data_template.tex
data_template will at least build a file for path only, just not formatted correctly later

--- 2019-01-04 paper/data_template.tex
@@
-\multirow{2}{*}{Classifier}&\multicolumn{4}{c||}{Performance}&\multicolumn{4}{c}{Rank Median(Mean)} \\ \cline{2-9}
-&TPR&FPR&AUC&CV10 Acc.&Overall (n={{ TEST_COUNTS.OVERALL }})&
+\multirow{2}{*}{Ranking System}&\multicolumn{4}{c||}{Case Rank - Median (Mean)}&\multicolumn{4}{c}{Percentage in Top X Variants - X=(1, 10, 20)} \\ \cline{2-9}
+&All (n={{ TEST_COUNTS.OVERALL }})&
 VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
-Likely Path. (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Pathogenic (n={{ TEST_COUNTS.PATHOGENIC }}) \\ \hline
+LP (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Path. (n={{ TEST_COUNTS.PATHOGENIC }})&
+All (n={{ TEST_COUNTS.OVERALL }})&
+VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
+LP (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Path. (n={{ TEST_COUNTS.PATHOGENIC }}) \\ \hhline{=|=|=|=|=||=|=|=|=}
+{% for cs_label in CS_LABELS %}
+    {{ cs_label }}&
+    {% for LEVEL in LEVELS %}
+        {{ '%0.1f (%0.2f)' % (np.median(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]), np.mean(COMPARISON[cs_label].TEST_RANKINGS[LEVEL])) }}&
+    {% endfor %}
+    {% for LEVEL in LEVELS %}
+        {{ '%d, %d, %d' % 
+            (100.0*np.sum(np.array(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]) <= 1) / len(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]) <= 10) / len(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]) <= 20) / len(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]))
+        }}
+        {% if LEVEL != 'PATHOGENIC' %}
+            &
+        {% endif %}
+    {% endfor %}
+    \\
+{% endfor %}
+\hhline{=|=|=|=|=||=|=|=|=}
+
+Exomiser$^*$&
+{% for LEVEL in LEVELS %}
+    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER.TEST_RANKINGS[LEVEL])) }}&
+{% endfor %}
+{% for LEVEL in LEVELS %}
+    {{ '%d, %d, %d' % 
+        (100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER.TEST_RANKINGS[LEVEL]))
+    }}
+    {% if LEVEL != 'PATHOGENIC' %}
+        &
+    {% endif %}
+{% endfor %}
+\\ \hhline{=|=|=|=|=||=|=|=|=}
+
 {% for clf_label in CLF_LABELS %}
     {{ clf_label }}&
-    {{ '%0.2f' % CLASSIFIERS[clf_label].TEST_TPR_RATE }}&
-    {{ '%0.2f' % CLASSIFIERS[clf_label].TEST_FPR_RATE }}&
-    {{ '%0.2f' % CLASSIFIERS[clf_label].ROC_AUC }}&
-    {{ '%0.2f+-%0.2f' % (CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[0], CLASSIFIERS[clf_label].CV10_BALANCED_ACCURACY[1]) }}&
-    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.OVERALL), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.OVERALL)) }}&
-    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE)) }}&
-    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.LIKELY_PATHOGENIC), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.LIKELY_PATHOGENIC)) }}&
-    {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS.PATHOGENIC), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS.PATHOGENIC)) }} \\
+    {% for LEVEL in LEVELS %}
+        {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL])) }}&
     {% endfor %}
+    {% for LEVEL in LEVELS %}
+        {{ '%d, %d, %d' % 
+            (100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 1) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 10) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 20) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]))
+        }}
+        {% if LEVEL != 'PATHOGENIC' %}
+            &
+        {% endif %}
+    {% endfor %}
+    \\
+{% endfor %}

### 2019-01-15 scripts/PyxisMapUtil.py
minor bug fix with args

@@ -17,7 +17,6 @@ def getPyxisMapResults(hpoTerms):
     '''
     #TODO: make this URL configurable?
     pyxisURL = 'http://0.0.0.0:4999/rank'
-    #pyxisURL = 'http://pyxismap.haib.org/rank'
     headers = {'Content-type': 'application/json'}
     data_json = json.dumps(sorted(hpoTerms))
     

### 2019-01-15 scripts/TestLearners.py

@@ -62,7 +62,7 @@ def getPyxisMapResults():
     hpoOnt = HPOUtil.HPOUtil(ontFN, annotFN)
     
     #load case data and go through them searching for HPO terms
-    caseData = SummaryDBUtil.loadSummaryDatabase(altIDMap)
+    caseData = SummaryDBUtil.loadSummaryDatabase(altIDMap, False)
     for sl in sorted(caseData.keys()):
         #TODO: figure out how to properly handle this
         if '/' in sl:

### 2019-01-15 paper/main.tex
cleaning up some TODOs and fixing a couple names I've heard back from

@@
 \author[Holt \textit{et~al}.]{James M.
-Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille Birch\,$^{\text{\sfb 1}}$, Donna Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa Wilk\,$^{\text{\sfb 1}}$, Julie Anderson\,$^{\text{\sfb 1}}$, Jeremy Harris\,$^{\text{\sfb 1}}$, Jacob Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A.
+Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille Birch\,$^{\text{\sfb 1}}$, Donna M.
+Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa A.
+Wilk\,$^{\text{\sfb 1}}$, Julie A.
+Anderson\,$^{\text{\sfb 1}}$, Jeremy Harris\,$^{\text{\sfb 1}}$, Jacob Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A.
 Worthey\,$^{\text{\sfb 1}}$}
@@
 In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.
-For the specific details on the filter used, refer to TODO:XXX (I think the source code?).
 In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).
+For the specific details on the filter used, please refer to the metadata available in the VarSight repository.
 
@@
 To reduce this dilution, we used principal component analysis \citep{jolliffe2011} to reduce the dimensions of each category and stored at most two features per categorical feature.
-For specifics on variance explained by the top two PCA components per category, please see Supplementary document XXX (TODO: are we going to make this or no?)
+%For specifics on variance explained by the top two PCA components per category, please see Supplementary document X (If we ever do this, it will be at the request of a user or reviewers, it doesn't seem super important at this juncture)
 

### 2019-01-17 paper/supplements/supplement.tex
supplementary doc v1

\documentclass{article}
\usepackage{graphicx}
\usepackage{rotating}

\begin{document}

\title{Supplementary Document}
\author{James M. Holt {\it et~al.}}

\maketitle

\tableofcontents

\section{Filtering}
The following sections describes the filter that was passed to Codicem for generating the test and training set variants.  Figure \ref{fig:filter_overview} shows the sequential order of filters, and the following subsections briefly describes each filter unit's purpose.

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{filter_supplement_overall.png}
\caption{Filter overview.  This high level image shows the sequential order of filtering, descriptions of sub-filters can be found in other sections.}
\label{fig:filter_overview}
\end{figure}

\subsection{Total Depth filter}

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{filter_supplement_layer0.png}
\caption{Total Depth filter.  This filter is primarily a QC-related filter that only passes variants with at least 8 reads covering the locus.}
\label{fig:filter0}
\end{figure}

The total depth filter is primarily a QC-related filter intended to remove variant calls with total coverage.  This filter only passes variants that have at least 8 reads overlapping the locus.  Figure \ref{fig:filter0} shows a visualization of the filter.

\subsection{Percentage of Reads Supporting Allele filter.}

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{filter_supplement_layer1.png}
\caption{Percentage of Reads Supporting Allele filter.  This filter is primarily a QC-related filter that only passes variants with at least 15\% of the locus' reads supporting the alternate allele.}
\label{fig:filter1}
\end{figure}

This filter is also a QC-related filter intended to remove variant calls with low support for the alternate allele.  The filter only passes variants with more than 15\% of the reads supporting the alternate allele.  Figure \ref{fig:filter1} shows a visualization of the filter.

\subsection{Population Allele Frequency filter.}

\begin{sidewaysfigure}
\centering
\includegraphics[width=\textwidth]{filter_supplement_layer2.png}
\caption{Population Allele Frequency filter.  This filter is primarily designed to filter out variants with a high population frequency.  It relies mostly on GnomAD and ExAC to determine allele frequencies.  Variants that are only semi-rare are allowed through if there are appropriate HGMD or ClinVar annotations supporting them.  Additionally, one variant in gene {\it F5} (rs6025) is allowed through because the reference allele is actually the rare, pathogenic allele.}
\label{fig:filter2}
\end{sidewaysfigure}

This filter removes variants that have a high population frequency.  There are three general ways for a variant to pass this filter: 1) be a rare variant in GnomAD AND a rare variant in ExAC; 2) be semi-rare in GnomAD, semi-rare in ExAC, AND have a high-confidence damaging annotation from HGMD; or 3) be semi-rare in GnomAD, semi-rare in ExAC, AND have a ClinVar classification labeling it as pathogenic, likely pathogenic, or a drug response.  Additionally, there is one variant from the gene {\it F5} (rs6025) allowed through this filter.  Note that this specific variant has a very high alternate allele frequency because the reference allele is actually the disease-causing mutation.  Figure \ref{fig:filter2} shows a visualization of the filter.

\subsection{HGMD, ClinVar, CADD, and Effects filter}

\begin{figure}
\centering
\includegraphics[width=.80\textwidth]{filter_supplement_layer3.png}
\caption{HGMD, ClinVar, CADD, and Effects filter.  This filter is primarily designed to filter out variants that are not predicted or annotated to have an effect on a transcript.}
\label{fig:filter3}
\end{figure}

This filter removes variants that are not predicted or annotated to have an effect on a transcript.  A variant can pass this filter in any of four ways: 1) have an HGMD accession, 2) have a ClinVar accession, 3) have an effect that is predicted to modify a transcript, or 4) have a very high CADD score.  Figure \ref{fig:filter3} shows a visualization of the filter.

\subsection{Gene Has Associated Disease filter}

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{filter_supplement_layer4.png}
\caption{Gene Has Associated Disease filter.  This filter removes variants for which there is no annotated disease name in HGMD, OMIM, or ClinVar.}
\label{fig:filter4}
\end{figure}

This filter removes variants for which there is no annotated disease name in HGMD, OMIM, or ClinVar.  If any of those annotations for the variant has a disease name, then the variant will pass this filter.  Figure \ref{fig:filter4} shows a visualization of the filter.

\subsection{Red Herring filter}

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{filter_supplement_layer5.png}
\caption{Red Herring filter.  This filter removes variants that are commonly found through sequencing, but do not appear in population databases like gnomAD or ExAC.  These variants are believed to be sequencing artifacts, so they are labeled as ``Red Herring" variants and filtered out.}
\label{fig:filter5}
\end{figure}

This filter removes variants that are sequencing artifacts.  Our analysts found that some variants are not found in population databases, but show up frequently ($>20$\% of samples) in our sequencing data.  These variants are believed to be sequencing artifacts, so Codicem maintains a database of these variants to filter out during analysis.  Figure \ref{fig:filter5} shows a visualization of the filter.

\subsection{Repeats filter}

\begin{figure}
\centering
\includegraphics[width=.65\textwidth]{filter_supplement_layer6.png}
\caption{Repeats filter.  This filter removes variants found in repeat tracks if there is no associated entry in HGMD or ClinVar.  These variants are generally considered polymorphic and/or sequencing artifacts unless there are annotations from disease databases.}
\label{fig:filter6}
\end{figure}

This filter removes variants that are polymorphic repeats and/or artifacts from sequencing.  It filters out any variants that are found in repeat tracks if there is no associated entry in HGMD or ClinVar.  Figure \ref{fig:filter6} shows a visualization of the filter.

\subsection{Unknowns filter}

\begin{figure}
\centering
\includegraphics[width=.65\textwidth]{filter_supplement_layer7.png}
\caption{Unknowns filter.  This filter removed variants with an ``Unknown" effect on transcription if there is not additional support for that variant through either CADD or HGMD.}
\label{fig:filter7}
\end{figure}

This filter removes variants that have an ``Unknown" effect on transcription if there isn't additional support for that variant through either CADD or HGMD.  If CADD is greater than 10 or there is an HGMD accession tied to the variant, it will pass this filter.  Figure \ref{fig:filter7} shows a visualization of the filter.

\subsection{Near Splice filter}

\begin{sidewaysfigure}
\centering
\includegraphics[width=\textwidth]{filter_supplement_layer8.png}
\caption{Near Splice filter.  This filter removes ``Near Splice Site Alteration" variants that are not supported to be deleterious by splice prediction algorithms, ClinVar annotations, or CADD predictions.}
\label{fig:filter8}
\end{sidewaysfigure}

This filter removes near splice variants that do not have additional annotated or predicted support indicating that the variant would impact splicing.  If a variant is a ``Near Splice Site Alteration", that variant will only pass the filter if one of the following is true: 1) both splice predictions algorithms predict an impact on splicing, 2) ClinVar has a pathogenic or likely pathogenic annotation, or 3) CADD is relatively high.  Figure \ref{fig:filter8} shows a visualization of the filter.

\end{document}

### 2019-01-17 scripts/SupplementGen.py

@@ -1,0 +1,194 @@
+
+import json
+import os
+
+def createFilterImages(filterFN, prefix):
+    '''
+    This script will generate DOT graph files and then images for a given filter.  It first draws the overall filter
+    layout and then an image for each sub-filter (which may be composed of ANDs or ORs or some combo)
+    '''
+    fp = open(filterFN, 'rt')
+    j = json.load(fp)
+    fp.close()
+
+    overallLayers = ['All Variants']
+    for i, filt in enumerate(j['queries']):
+        layerPrefix = prefix+'layer'+str(i)
+        createLayerImage(layerPrefix, filt)
+        overallLayers.append(filt['name'])
+    overallLayers.append('Passed Variants')
+
+    #draw the overall
+    overallDotFN = prefix+'overall.dot'
+    overallPngFN = prefix+'overall.png'
+
+    fp = open(overallDotFN, 'wt+')
+    fp.write('digraph {\n')
+    prev = None
+    for i, n in enumerate(overallLayers):
+        nodeLabel = 'node'+str(i)
+        if n in ['All Variants', 'Passed Variants']:
+            shape = 'Mdiamond'
+        else:
+            shape='rect'
+        fp.write('\t'+nodeLabel+' [label="'+n+'", shape='+shape+'];\n')
+
+        if prev != None:
+            fp.write('\t'+prev+' -> '+nodeLabel+';\n')
+        prev = nodeLabel
+    fp.write('}\n')
+    fp.close()
+
+    #cmd: dot -Tpng -o <filename.png> <input.dot>
+    os.system('dot -Tpng -o '+overallPngFN+' '+overallDotFN)
+
+def createLayerImage(layerPrefix, layerDict):
+    '''
+    Creates a DOT and image file for a given layer
+    @param layerPrefix - the prefix of the filename, creates a <layerPrefix>.dot and <layerPrefix>.png file
+    @param layerDict - the dictionary corresponding to the layer
+    '''
+    dotFN = layerPrefix+'.dot'
+    pngFN = layerPrefix+'.png'
+    fp = open(dotFN, 'wt+')
+    fp.write('digraph {\n')
+    
+    print(layerDict['name'])
+    nodes = [('source', 'All Variants'), ('dest', 'Pass Variants')]
+    for n, l in nodes:
+        fp.write('\t'+n+' [label="'+l+'" shape=Mdiamond];\n')
+
+    #create a subgraph for labeling purposes
[…]

### 2019-01-17 paper/main.tex
updating .gitignore and adding Julies edits

@@
 
-\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the major goal is to identify one or more variants that cause disease.
+\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause disease.
 Typically, this is done through filtering and then prioritization of variants for manual curation.
 However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.
-Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability of critical importance.
+Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.
 \\
-\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will be ultimately clinically reported and returned to a patient.
+\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.
 To test the classifiers, we performed a retrospective study on variants that were clinically reported to patients in the Undiagnosed Diseases Network.
@@
 After sequencing, we followed GATK best practices \citep{depristo2011} to align to the hg19 reference with BWA-mem \citep{li2013}.
-Aligned sequences we processed via GATK for base quality score recalibration, indel realignment, and duplicate removal.
+Aligned sequences were processed via GATK for base quality score recalibration, indel realignment, and duplicate removal.
 Finally, SNV and indel variants were joint genotyped, again according to GATK best practices \citep{depristo2011}.
@@
 Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.
-We used the latest version of their primary clinical filter to replicate the standard filtering process of the analysts.
+We used the latest version of the primary clinical filter to replicate the standard filtering process of the analysts.
 In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.
@@
 The second method, a tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.
-The PyxisMap method has the added benefit of incorporating fresh gene-phenotype connections from newer papers, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.
+The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.
 Each method generates a single numerical feature that is used in the following analyses.
@@
 Note that secondary and/or incidental findings are specifically not included in this list.
-The analysts assigned each primary variant a rank from variant of uncertain significance (VUS), likely pathogenic, or pathogenic based on the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.
+The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic based on the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.
 
@@
 \subsection{Model training and tuning}
-As noted earlier, there are generally hundreds of variants per proband that make it through the filter, but only a few are ever clinically reported.
+As noted earlier, there are generally hundreds of variants per proband that pass the filter, but only a few are ever clinically reported.
+Across all 237 proband samples, there were a total of 378 clinically reported variants and another 87819 variants that were seen but not reported.
 As a result, there is a major imbalance in the number of true positives (variants clinically reported) and true negatives (variants seen, but not clinically reported).
-Across all 237 proband samples, there were a total of 378 clinically reported variants and another 87819 variants that were seen but not reported.
+
 We split the data into training and test sets on a per-proband basis with the primary goal of roughly balancing the total number of true positives in each set.
@@
 For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).
-We used followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.
+We followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.
 For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.
@@
 For ``Pathogenic" variants, the median rank was 2.0 and 97\% of those variants were ranked in the top 20 for the case.
-While not perfect, the ability for the algorithms to push reported variants to very early in the list is quite promising.
+While these algorithms are not perfect classifiers, their use as a prioritization system is quite promising.
 
-We expect these classification algorithms could be improved or refined in a variety of ways.
+We expect these classification algorithms could be refined in a variety of ways.
 First, adding new features and/or removing unused features could lead to improvements in the algorithm.
@@
 
-Ultimately, we believe the trained classifiers in VarSight are a significant step forward in handling real-world data for rare, undiagnosed disease patients.
-Despite the variability in phenotype and allowance of variants of uncertain significance in reporting, the models trained in VarSight still ranked reported variants highly, especially variants labeled as ``Pathogenic".
-Going forward, we expect the standard for variant ranking systems will behave similar to those in VarSight, fully integrating all available phenotype and variant annotations into one model.
+We believe the trained classifiers in VarSight are a significant step forward in reducing the problem of variant fatigue.
+The models improve our ability to prioritize variants despite the variability and uncertainty injected by real-world data.
+Ultimately, we believe implementing these models will enable analysts to assess the best candidate variants first, reducing the time to analyze a case and return molecular diagnoses to patients.
 
@@
 We would like to acknowledge all of the teams within the UDN including the coordinating center and all of the clinical sites working hard to provide definitive diagnoses for UDN patients.
-We would also like to acknowledge the UDN WGS core headed by Dr.
+We would also like to acknowledge the UDN whole genome sequencing core headed by Dr.
 Shawn Levy.
@@
 %HPO paper to cite
-\bibitem[Köhler {\it et~al}., 2018]]{kohler2018} Köhler, Sebastian, \textit{et~al.} "Expansion of the Human Phenotype Ontology (HPO) knowledge base and resources." {\it Nucleic acids research} (2018).
+\bibitem[Köhler {\it et~al}., 2018]{kohler2018} Köhler, Sebastian, \textit{et~al.} "Expansion of the Human Phenotype Ontology (HPO) knowledge base and resources." {\it Nucleic acids research} (2018).
 
@@
 %PubTator
-\bibitem[Wei {\it et~al}., 2013]]{wei2013} Wei, Chih-Hsuan, Hung-Yu Kao, and Zhiyong Lu.
+\bibitem[Wei {\it et~al}., 2013]{wei2013} Wei, Chih-Hsuan, Hung-Yu Kao, and Zhiyong Lu.
 "PubTator: a web-based text mining tool for assisting biocuration." {\it Nucleic acids research} 41.W1 (2013): W518-W522.

### 2019-01-17 scripts/SupplementGen.py
minor changes

@@ -75,6 +75,7 @@ def createLayerImage(layerPrefix, layerDict):
 
 REP_DICT = {
     "GREATER_THAN_OR_EQUAL" : "&ge;",
+    "GREATER_THAN" : "&gt;",
     "EQUALS" : "==",
     "DOES_NOT_EQUAL" : "&ne;",
     "LESS_THAN_OR_EQUAL" : "&le;",
@@ -123,21 +124,6 @@ def recurseCriteria(fp, currDict, nodeLabelList):
         
         if allSingles:
             #this is all single values comparisons, make one box
-            recordRows = []
-            '''
-            for subFilter in currDict['filters']:
-                assert(len(subFilter['criteria']['op']) == 1)
-                relat = REP_DICT.get(subFilter['criteria']['op'][0], subFilter['criteria']['op'][0])
-                rowEntry = [subFilter['logic'], subFilter['field'], relat, '\\n'.join(subFilter['criteria']['val'])]
-                recordRows.append('{'+'|'.join(rowEntry)+'}')
-
-            recordRows.append('JOIN('+currDict['join']+')')
-            recordRows.append(currDict['logic'])
-            recordLabel = "{"+'|'.join(recordRows)+"}"
-            nodeJoin = 'join'+'v'.join([str(x) for x in nodeLabelList])
-            fp.write('\t\t'+nodeJoin+' [shape=record label="'+recordLabel+'"];\n')
-            '''
-            
             recordLabel = '<TABLE BORDER="0" CELLBORDER="1" CELLSPACING="0" CELLPADDING="4">'
             recordLabel += '<TR><TD COLSPAN="4">'+currDict['logic']+'</TD></TR>'
             recordLabel += '<TR><TD COLSPAN="4">BEGIN: JOIN('+currDict['join']+')</TD></TR>'

### 2019-01-19 paper/main.tex
updated with mana feedback

@@
 \author[Holt \textit{et~al}.]{James M.
-Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille Birch\,$^{\text{\sfb 1}}$, Donna M.
+Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille L.
+Birch\,$^{\text{\sfb 1}}$, Donna M.
 Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa A.
@@
 \textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.
-To test the classifiers, we performed a retrospective study on variants that were clinically reported to patients in the Undiagnosed Diseases Network.
-We measured the classifier against another variant prioritization algorithm along with two single-measure controls to show how the classifiers tend to outperform the other methods.
-The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.
+To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.
+We treated the classifiers as a ranking system and compared them to another variant prioritization algorithm and two single-measure controls.
+% to show how the classifiers tend to outperform the other methods.
+We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
+%this was probably too detailed, need a different sentence
+%The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  
 \\
@@
 This variant fatigue means that variants at the end of the prioritized list are often regarded with less scrutiny and/or a less accurate patient model than those near the beginning.
-Additionally, if causative variants can be identified earlier due to a high rank, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.
+Methods that can prioritize these variants accurately can reduce the impact of variant fatigue, reducing the chances that variants are overlooked or mis-identified.
+Additionally, if causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.
 Finally, accurate prioritization is a step towards the ultimate goal of automatically identifying all variants that cause a patient's primary phenotypes.
@@
 Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as phenotypes caused by old age).
-Finally, analysts have the discretion to send back variants with different levels of pathogenicity as specified by the ACMG guidelines \cite{richards2015}.
-The summation of these real-world problems means that prioritizing variants is challenging due to noisy phenotype inputs and variation in the pathogenicity of clinically reported outputs.
+%Finally, analysts have the discretion to send back variants with different levels of pathogenicity as specified by the ACMG guidelines \citep{richards2015}.  
+Because the UDN is a research collaboration, there is also variability in reported variants that range in pathogenicity from ``variant of uncertain significance" (VUS) through ``pathogenic" as defined by the ACMG guidelines \citep{richards2015}.
+The summation of this real-world variation means that accurately prioritizing variants is challenging due to noise and variation in phenotype inputs and variation in pathogenicity of reported variant outputs.
 
@@
 We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.
-Each classifier calculated probabilities of a variant belong to the ``returned" class, allowing for both classification analysis and ranking of the variants by their calculated probabilities.
+Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing for both classification analysis and ranking of the variants by their calculated probabilities.
 After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant ranking perspective.
@@
 \subsection{Data sources}
-All samples were selected from the cohort of genome sequencing samples originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB) Sequencing Center of the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.
+All samples were selected from the cohort of Undiagnosed Diseases Network (UDN) \citep{ramoni2017} genome sequencing samples that were originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB).
 In short, the UDN accepts patients with rare, undiagnosed diseases that are believed to have a genetic origin.
 The UDN is not restricted to a particular disease, so there are a diverse set of diseases and phenotypes represented across the whole population.
-The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) some patients have multiple diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, and 3) some patients exhibit more or less phenotypes than are associated with the classic disease presentation.
+The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) some patients have multiple diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, and 3) some patients exhibit more or fewer phenotypes than are associated with the classic disease presentation.
 For more details on the UDN, refer to \citealp{ramoni2017}.
@@
 \subsection{Alignment and variant calling}
-After sequencing, we followed GATK best practices \citep{depristo2011} to align to the hg19 reference with BWA-mem \citep{li2013}.
+After sequencing, we followed GATK best practices \citep{depristo2011} to align to the GRCh37 human reference genome with BWA-mem \citep{li2013}.
 Aligned sequences were processed via GATK for base quality score recalibration, indel realignment, and duplicate removal.
 Finally, SNV and indel variants were joint genotyped, again according to GATK best practices \citep{depristo2011}.
-The end result of this pipeline is a Variant Call Format (VCF) file used in the following sections.
+The end result of this pipeline is one Variant Call Format (VCF) file per patient sample.
+This collection of VCF files is used in the following sections.
 
@@
 Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.
-We used the latest version of the primary clinical filter to replicate the standard filtering process of the analysts.
+We used the latest version of the primary clinical filter for rare disease variants to replicate the standard filtering process for patients in the UDN.
 In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.
 In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).
-For the specific details on the filter used, please refer to the metadata available in the VarSight repository.
+For the specific details on the filter used, please refer to Supplementary Documents.
+%metadata available in the VarSight repository.
 
@@
 This method tends to be more conservative because it relies solely on curated annotations from the HPO.
-The second method, a tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.
+The second method, an internally-developed tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.
 The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.
+We used PyxisMap v1.2, and we ran the standard installation script that downloads all required data sources on December 19, 2018.
 Each method generates a single numerical feature that is used in the following analyses.
@@
 We required the following for each proband sample included in our analyses: 1) at least one clinically reported primary variant that came through the primary clinical filter (i.e.
-it was not found through some other targeted search) and 2) a set of phenotypes annotated using the Human Phenotype Ontology \citep{kohler2018}.
+it was not found through some other targeted search) and 2) a set of phenotypes annotated with Human Phenotype Ontology \citep{kohler2018} terms using the Phenotips software \citep{girdea2013}.
 At the time of writing, this amounted to 378 primary, reported variants spanning a total of 237 proband samples.
@@
 For the purposes of classification, all annotations needed to be cleaned and stored as numerical features.
-For numerical annotations, we simply copied the annotation over as a single value feature.
-Default values were assigned to missing annotations that were outside the expected value range.
-Additionally, these values were always on the less impactful side of the spectrum (e.g.
+For numerical annotations (e.g.
+float values like CADD or GERP), we simply copied the annotation over as a single value feature.
+Missing annotations were assigned a default value that was outside the expected value range for that feature.
+Additionally, these default values were always on the less impactful side of the spectrum (e.g.
 a default conservation score would err on the side of not being conserved).
@@
 For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.
-We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data.
-The values for each model are shown in Table \ref{tab:classifier_results}.
-Additionally, receiver operator curves for FPR v.
+We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}.
+%The values for each model are shown in Table \ref{tab:classifier_results}.  
+Figure \ref{fig:roc} show receiver operator curves for FPR v.
 TPR and recall v.
-precision were created for the four models.
-We compared this to two single-measure statistics (CADD Scaled and HPO-cosine) in Figure \ref{fig:roc}.
+precision for the four models.
+Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.
 
-From these metrics, the two random forest models and the EasyEnsembleClassifier seem to perform similarly whereas LogisticRegression performs slightly worse across the board.
+From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.
 All four classifiers have higher AUROC than either single-measure statistic.
@@
 \processtable{Classifier performance statistics.
-For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy, true positive rate, false positive rate, and area under the receiver operator curve.
+For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC).
 In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.
@@
 The bottom four rows are the tuned, binary classification methods tested in this paper.
-The ``Case Rank" columns show the median and mean ranks for all variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).
+The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).
 The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.
@@
 We then calculated median and mean rank statistics for the reported variants.
-Additionally, we quantified the percentage of variants that were ranked in the top 1, 10, and 20 variants in each case.
+Additionally, we quantified the percentage of reported variants that were ranked in the top 1, 10, and 20 variants in each case.
 While the models were trained as a binary classification system, we broke down the results further to demonstrate differences between variants that were clinically reported as a variant of uncertain significance (VUS), likely pathogenic, and pathogenic.
@@
 For primary variants that were unranked by Exomiser, we conservatively assumed that the Exomiser ranked those variants as just outside the range of what was returned (e.g.
-if Exomiser ranked 15 variants, we assumed a missing variant was ranked 16th).
+if Exomiser ranked 15 variants, we assumed all missing variants were ranked 16th).
 
 Finally, we added two control scores for comparison: CADD scaled and HPO-cosine.
-These scores were inputs to each classifier, but also represent some common ways variants one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).
+These scores were inputs to each classifier, but also represent two common ways variants one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).
 The results for the two control scores, Exomiser, and all four classifiers are shown in Table \ref{tab:performance}.
@@
 In the overall data, all four classifiers outperform the single-measure statistics and Exomiser across the board, with the one exception being Exomiser's mean (this measure is likely heavily biased due to our conservative ranking of unranked variants, harsher penalties for unranked variants would drastically change this result).
-As expected, all methods perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.
-Of the methods tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.
+As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.
+Of the classifiers tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.
 
@@
 After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
-Both random forest models calculate a feature importance array after training that sum to 1.0.
+Both random forest models calculate a feature importance array after training that sum to 1.0 (the other two classifiers do not have this information readily available).
 By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.
@@
 We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report.
-We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process used for the Undiagnosed Diseases Network (UDN).
+We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process obtained from the Undiagnosed Diseases Network (UDN).
 From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.
@@
 URL: \href{http://envisiongenomics.com/codicem-analysis-platform/}{http://envisiongenomics.com/codicem-analysis-platform/}.
+
+%Phenotips
+\bibitem[Girdea {\it et~al}., 2013]{girdea2013} Girdea, Marta, \textit{et~al.} "PhenoTips: Patient Phenotyping Software for Clinical and Research Use." {\it Human mutation} 34.8 (2013): 1057-1065.
 

### 2019-01-22 paper/data_template.tex
changed up some of the output to support mana's queries

@@
 
-Exomiser$^*$&
+Exomiser(best)&
 {% for LEVEL in LEVELS %}
-    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER.TEST_RANKINGS[LEVEL])) }}&
+    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER_BEST.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER_BEST.TEST_RANKINGS[LEVEL])) }}&
 {% endfor %}
@@
     {{ '%d, %d, %d' % 
-        (100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
-        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
-        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER.TEST_RANKINGS[LEVEL]))
+        (100.0*np.sum(np.array(EXOMISER_BEST.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER_BEST.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER_BEST.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER_BEST.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER_BEST.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER_BEST.TEST_RANKINGS[LEVEL]))
     }}
@@
 {% endfor %}
+\\
+
+Exomiser(avg)&
+{% for LEVEL in LEVELS %}
+    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL])) }}&
+{% endfor %}
+{% for LEVEL in LEVELS %}
+    {{ '%d, %d, %d' % 
+        (100.0*np.sum(np.array(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]),
+        100.0*np.sum(np.array(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]))
+    }}
+    {% if LEVEL != 'PATHOGENIC' %}
+        &
+    {% endif %}
+{% endfor %}
+
 \\ \hhline{=|=|=|=|=||=|=|=|=}

### 2019-01-22 paper/main.tex
changed up some of the output to support mana's queries

@@
 Birch\,$^{\text{\sfb 1}}$, Donna M.
-Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa A.
+Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alexander C.
+Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa A.
 Wilk\,$^{\text{\sfb 1}}$, Julie A.
-Anderson\,$^{\text{\sfb 1}}$, Jeremy Harris\,$^{\text{\sfb 1}}$, Jacob Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A.
+Anderson\,$^{\text{\sfb 1}}$, Jeremy M.
+Harris\,$^{\text{\sfb 1}}$, Jacob M.
+Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina E.
+Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A.
 Worthey\,$^{\text{\sfb 1}}$}
@@
 
-\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause disease.
+\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.
 Typically, this is done through filtering and then prioritization of variants for manual curation.
@@
 To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.
-We treated the classifiers as a ranking system and compared them to another variant prioritization algorithm and two single-measure controls.
+We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls.
 % to show how the classifiers tend to outperform the other methods.
@@
 \section{Introduction}
-Genome and exome sequencing are both currently being used as molecular diagnostic tools for patients with rare, undiagnosed diseases \citep{ramoni2017} (TODO: XXX: CITE MORE, CSER maybe?).
-Typically, these technologies are applied clinically following the workflows consisting of blood draw, sequencing, alignment, variant calling, variant annotation, variant filtering, and variant prioritization \citep{worthey2017, roy2018}.
+Genome and exome sequencing are both currently being used as molecular diagnostic tools for patients with rare, undiagnosed diseases \citep{ramoni2017, bagnall2018, sweeney2018}.
+Typically, these technologies are applied clinically following workflows consisting of blood draw, sequencing, alignment, variant calling, variant annotation, variant filtering, and variant prioritization \citep{worthey2017, roy2018}.
 Then, clinical analysts usually perform the more manual processes of inspecting and then clinically reporting variants based on the known set of patient phenotypes.
@@
 These methods use a wide range of input sources including but not limited to population allele frequencies \citep{lek2016}, conservation scores \citep{cooper2005, siepel2006, petrovski2013}, haploinsufficiency scores \citep{huang2010, steinberg2015}, deleteriousness scores \citep{steinberg2015, rentzsch2018}, transcript impact scores \citep{kumar2009, choi2012, adzhubei2013, dong2014, jian2014} , and previously associated disease annotation \citep{stenson2003, hamosh2005, landrum2015}.
-Variant prioritization is also quite diverse with some methods relying on the variant annotations to order variants \citep{hu2013} and some relying on patient phenotype to order the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.
+Variant prioritization is also quite diverse with some methods relying on the variant annotations to prioritize variants \citep{hu2013} and some relying on patient phenotype to rank the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.
 There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.
@@
 In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.
-In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).
+In general, the filter reduces the number of variants from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).
 For the specific details on the filter used, please refer to Supplementary Documents.
@@
 The Codicem annotations are all agnostic of the patient phenotype.
-As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases and phenotypes tied to UDN patients.
+As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases, phenotypes, and genetic heritage tied to UDN patients.
 In order to incorporate patient phenotype information, we used two distinct methods to rank genes based on the Human Phenotype Ontology (HPO) \citep{kohler2018}.
@@
 Note that secondary and/or incidental findings are specifically not included in this list.
-The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic based on the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.
+The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic adhering to the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.
 
@@
 
-\section{Discussion}
+\section{Results}%mana said results!
 \subsection{Classifier Statistics}
@@
 \caption{Receiver operator curves.  
-This figure shows the receiver operator curves for the test data for each single-measure statistic and for each trained classifier.
+%This figure shows the receiver operator curves for the test data for each single-measure statistic and for each trained classifier.  
+This figure shows the receiver operator curves for the test data for each trained classifier.  
 On the left, we show the false positive rate against the true positive rate.
@@
 Area under the receiver-operator curve (AUROC) is reported beside each method in the legend.
-In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly.
-LogisticRegression tends to perform slightly worse overall, but still better than either single-measure statistic.}
+In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly with LogisticRegression performing slightly worse overall.}
 \label{fig:roc}
@@
 precision for the four models.
-Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.
+%Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.
 
 From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  
-All four classifiers have higher AUROC than either single-measure statistic.
-However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was only 0.2109).
+%All four classifiers have higher AUROC than either single-measure statistic.  
+However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was 0.2109).
 This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.
@@
 Exomiser is an external tool that only reported ranks for a subset of the filtered variants.
+``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.
+In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e.
+middle) for all unranked variants.
 The bottom four rows are the tuned, binary classification methods tested in this paper.
@@
 \hline
-\end{tabular}}{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
+\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
 \end{table*}
@@
 We then parsed the output JSON file created by Exomiser into a ranked order of variants.
-Of note, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 20-50 variants from the pre-filtered VCF.
-For primary variants that were unranked by Exomiser, we conservatively assumed that the Exomiser ranked those variants as just outside the range of what was returned (e.g.
-if Exomiser ranked 15 variants, we assumed all missing variants were ranked 16th).
+
+Unfortunately, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 10-50 variants from the pre-filtered VCF.
+For the reported variants, Exomiser did not rank 98 of the 189 variants in the test set (58 variant of uncertain significance, 21 likely pathogenic, and 19 pathogenic).
+For these unranked variants, we assigned a rank using two different methods.
+The ``Exomiser(best)" method conservatively assumed that unranked primary variants were found at the next best position in the rank order (e.g.
+if Exomiser ranked 15 variants, we assumed all unranked variants were ranked 16th).
+In contrast, the ``Exomiser(avg)" method assumed that unranked primary variants were ranked at the average position for all unranked variants (e.g.
+if Exomiser ranked 10 of 100 variants, we assumed all unranked variants were ranked 55th).
+``Exomiser(best)" represents the best possible ordering that Exomiser could give us for unranked variants where ``Exomiser(avg)" is a more realistic expectation.
 
 Finally, we added two control scores for comparison: CADD scaled and HPO-cosine.
-These scores were inputs to each classifier, but also represent two common ways variants one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).
-The results for the two control scores, Exomiser, and all four classifiers are shown in Table \ref{tab:performance}.
+These scores were inputs to each classifier, but also represent two common ways one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).
+The results for the two control scores, both Exomiser approaches, and all four classifiers are shown in Table \ref{tab:performance}.
 
-In the overall data, all four classifiers outperform the single-measure statistics and Exomiser across the board, with the one exception being Exomiser's mean (this measure is likely heavily biased due to our conservative ranking of unranked variants, harsher penalties for unranked variants would drastically change this result).
+In the overall data, all four classifiers outperform the single-measure statistics and both Exomiser approaches across the board, with the one exception being the mean of ``Exomiser(best)" (a measure that is heavily biased by the conservative ranking of unranked variants).
 As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.
@@
 After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
-Both random forest models calculate a feature importance array after training that sum to 1.0 (the other two classifiers do not have this information readily available).
+Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).
+The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.
 By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.
-Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ as reported by each classifier.
+Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.
 
@@
 The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may improve the result.
+Finally, training the model on different patient populations may yield different models.
+We trained on a fairly diverse patient population
 
@@
 \section*{Acknowledgements}
-We are grateful for the participation of patients and family members and all of their referring clinicians.
+We are grateful for the participation of patients and family members of the UDN and all of their referring clinicians.
 We would like to acknowledge all of the teams within the UDN including the coordinating center and all of the clinical sites working hard to provide definitive diagnoses for UDN patients.
@@
 
+%clinical WGS example
+\bibitem[Bagnall {\it et~al}., 2018]{bagnall2018} Bagnall, Richard D., \textit{et~al.} "Whole genome sequencing improves outcomes of genetic testing in patients with hypertrophic cardiomyopathy." {\it Journal of the American College of Cardiology} 72.4 (2018): 419-429.
+
 %PROVEAN - variant-transcript prediction
@@
 %GERP - conservation
-\bibitem[Cooper {\it et~al}., 2005]{cooper2005} Cooper, Gregory M., \textit{et al.} "Distribution and intensity of constraint in mammalian genomic sequence." {\it Genome research} 15.7 (2005): 901-913.
+\bibitem[Cooper {\it et~al}., 2005]{cooper2005} Cooper, Gregory M., \textit{et~al.} "Distribution and intensity of constraint in mammalian genomic sequence." {\it Genome research} 15.7 (2005): 901-913.
 
@@
 
+%Example WGS in clinic
+\bibitem[Sweeney {\it et~al}., 2018]{sweeney2018} Sweeney, Nathaly M., \textit{et~al.} "The case for early use of rapid whole genome sequencing in management of critically ill infants: Late diagnosis of Coffin-Siris syndrome in an infant with left congenital diaphragmatic hernia, congenital heart disease and recurrent infections." {\it Molecular Case Studies} (2018): mcs-a002469.
+
 %ANNOVAR - variant annotation

### 2019-01-22 scripts/ExomiserUtil.py

@@ -48,7 +48,8 @@ def getTargetRanks(sl, foundPrimaries):
         try:
             ind = fullRanks.index(fp)
         except:
-            ind = len(fullRanks)
+            assert(len(fullRanks) != 0)
+            ind = -1*len(fullRanks)
         ret.append(ind)
     return ret
 

### 2019-01-22 scripts/TestLearners.py

@@ -120,12 +120,13 @@ def loadFormattedData(args):
     '''
     This function is primarily about loading and formatting data prior to training/testing any classifiers.
     @param args - arguments from the command line
-    @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts)
+    @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts, exomiserRanks)
         xFinal - an NxM matrix where N is the number of variants in our test/training set and M is the number of features; contains feature values
         yFinal - an N length array where N is the number of variants in our test/training set; 1 if the variant was reported
         featureLabels - an M length array containing feature labels for our output benefit
         startIndices - a (C+1) length array containing the start indices of variants from an individual case
         allRepDicts - an N length array of None or dictionaries; if a dictionary, it contains data on a returned variant
+        exomiserRanks - a list of lists containing exomiser ranks on a per-case basis; missing values are -1*len(ranked variants)
     '''
     pyxisRootDir = '/Users/matt/githubProjects/VarSight/pyxis_ranks_'+PYXIS_DATE
     
@@ -528,16 +529,12 @@ def runClassifiers(args, values, classifications, featureLabels, startIndices, a
     @param featureLabels - C length array containing labels (strings) for the features
     @param startIndices - the startIndices of individual cases in the training set
     @param allRepDicts - a list of dictionaries for variants that were reported; non-reported vars are None
-    @param exomiserRanks - paired exomiser ranks for each case; needs to be split up with train/test
+    @param exomiserRanks - paired exomiser ranks for each case; needs to be split up with train/test; missing values are -1*len(ranked variants)
     @return resultsDict - a dictionary containing many results we wish to include in a paper
     '''
     resultsDict = {}
     resultsDict['FEATURE_LABELS'] = featureLabels
     print('Values:', values.shape)
-    
-    #V1
-    #Classifications (and bincount): (36346,) [36165   181]
-    #Num cases: 108
     print('Classifications (and bincount):', classifications.shape, np.bincount(classifications))
     print('Num cases:', len(startIndices)-1)
 
[…]

### 2019-01-24 paper/main.tex
added a preprint toggle for biorxiv and easy switch to normal

@@
+\newif\ifpreprint
+\preprinttrue
+
+\ifpreprint
+\documentclass{article}
+\else
 \documentclass{bioinfo}
+\fi
+
 \usepackage{caption}
 \usepackage{graphicx}
+\usepackage{hhline}
 \usepackage{multirow}
-\usepackage{hhline}
+
+\ifpreprint
+\usepackage{hyperref}
+\usepackage{natbib}
+\usepackage{amsmath}
+%need these to rotate
+\usepackage{geometry}
+\usepackage{pdflscape}
+\usepackage{scrextend}
+\usepackage{authblk}
+
+\begin{document}
+\title{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
+\author[1]{James M. Holt}
+\author[1]{Brandon Wilk}
+\author[1]{Camille L. Birch}
+\author[1]{Donna M. Brown}
+\author[1]{Manavalan Gajapathy}
+\author[1]{Alexander C. Moss}
+\author[1]{Nadiya Sosonkina}
+\author[1]{Melissa A. Wilk}
+\author[1]{Julie A. Anderson}
+\author[1]{Jeremy M. Harris}
+\author[1]{Jacob M. Kelly}
+\author[1]{Fariba Shaterferdosian}
+\author[1]{Angelina E. Uno-Antonison}
+\author[1]{Arthur Weborg}
+\author[ ]{Undiagnosed Diseases Network}
+\author[1]{Elizabeth A. Worthey}
+\affil[1]{Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA}
+\maketitle
+
+\begin{abstract}
+\noindent\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.
+Typically, this is done through filtering and then prioritization of variants for manual curation.
+However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.
+Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.
+\\
+\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.
+To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.
+We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls.
+We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
+\\
+\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
+\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
+\end{abstract}
+
+\else%begin non-preprint
+
 \copyrightyear{2015} \pubyear{2015}
@@
 
+%end non-preprint
+\fi
+
 \section{Introduction}
@@
 
+\ifpreprint
+
+\else
 \begin{methods}
+\fi
 \section{Methods}
@@
 These four tuned models were then tested against the unseen set of test proband cases.
+\ifpreprint
 
+\else
 \end{methods}
+\fi
 
@@
 For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.
-We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}.
+We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}).
 %The values for each model are shown in Table \ref{tab:classifier_results}.  
@@
 This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.
+
+\ifpreprint
+\begin{table}[!t]
+\centering
+\begin{tabular}{c|c|c|c|c}
+\hline
+\input{classifier_rendered.tex}
+\hline
+\end{tabular}
+\caption{Classifier performance statistics.
+For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC).
+In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.}
+\label{tab:classifier_results}
+\end{table}
+
+\else
 
@@
 \end{table}
+\fi
+
+\ifpreprint
+\newgeometry{margin=1cm} % modify this if you need even more space
+\begin{landscape}
+\begin{table}[!t]
+\centering
+\begin{tabular}{c|c|c|c|c||c|c|c|c}
+\hline
+\input{data_rendered.tex}
+\hline
+\end{tabular}
+\caption{Ranking performance statistics.
+This table shows the ranking performance statistics for all methods evaluated on our test set.
+CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.
+Exomiser is an external tool that only reported ranks for a subset of the filtered variants.
+``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.
+In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e.
+middle) for all unranked variants.
+The bottom four rows are the tuned, binary classification methods tested in this paper.
+The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).
+The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.
+All values were generated using only the test data that was unseen during training.}
+\label{tab:performance}
+\end{table}
+\end{landscape}
+\restoregeometry
+\subsection{Ranking Statistics}
+
+\else
 
@@
 \end{table*}
+\fi
 
@@
 
+\ifpreprint
+\begin{table}[!t]
+\centering
+\begin{tabular}{l|c|c}
+\hline
+\input{features_rendered.tex}
+\hline
+\end{tabular}
+\caption{Random forest feature importances.
+This table shows the top feature importances reported by the two random forest algorithms we tested.
+We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance.}
+\label{tab:features}
+\end{table}
+
+\else
+
 \begin{table}[!t]
@@
 \end{table}
+\fi
 

### 2019-01-25 scripts/TestLearners.py
updating with best performers in code

@@ -627,7 +627,7 @@ def runClassifiers(args, values, classifications, featureLabels, startIndices, a
     resultsDict['TRAINING_MODE'] = ['EXACT_MODE', 'GRID_MODE', 'RANDOM_MODE'][currentMode]
 
     classifiers = [
-        ('RandomForest(sklearn)', RandomForestClassifier(random_state=0, class_weight='balanced', max_depth=4, n_estimators=300, min_samples_split=2, max_features='sqrt'), 
+        ('RandomForest(sklearn)', RandomForestClassifier(random_state=0, class_weight='balanced', max_depth=3, n_estimators=100, min_samples_split=2, max_features='sqrt'), 
         {
             'random_state' : [0],
             'class_weight' : ['balanced'],
@@ -638,7 +638,7 @@ def runClassifiers(args, values, classifications, featureLabels, startIndices, a
         }),
         #('svc_bal', SVC(probability=True, class_weight="balanced")),#this one doesn't scale well past 10k samples
         #('mlp', MLPClassifier()),#this one doesn't work, presumably because there is no balanced option
-        ('LogisticRegression(sklearn)', LogisticRegression(random_state=0, class_weight='balanced', penalty='l2', C=0.01, solver='liblinear', max_iter=200), 
+        ('LogisticRegression(sklearn)', LogisticRegression(random_state=0, class_weight='balanced', penalty='l2', C=10.0, solver='newton-cg', max_iter=200), 
         {
             'random_state' : [0],
             'class_weight' : ['balanced'],

### 2019-03-13 scripts/ExomiserUtil.py
added better comparisons to scripts and refactored some for later

@@ -5,12 +5,14 @@ This file should contain helper functions to parse and load Exomiser ranking fil
 import json
 import os
 
-def loadAllExomiserRanks(sl):
+def loadAllExomiserRanks(sl, subtype):
     '''
     Loads a .json file from Exomiser
     @param sl - the SL number to load
+    @param subtype - the particular run of exomiser
     '''
-    fn = '/Users/matt/githubProjects/VarSight/exomiser_results/'+sl+'-hiphive-genome-PASS_ONLY.json'
+    #fn = '/Users/matt/githubProjects/VarSight/exomiser_results/'+sl+'-hiphive-genome-PASS_ONLY.json'
+    fn = '/Users/matt/githubProjects/VarSight/exomiser_results_v2/%s-%s.json' % (sl, subtype)
     fp = open(fn, 'rt')
     j = json.load(fp)
     fp.close()
@@ -35,21 +37,34 @@ def loadAllExomiserRanks(sl):
     
     return orderedRet
 
-def getTargetRanks(sl, foundPrimaries):
+def getTargetRanks(sl, foundPrimaries, subtype):
     '''
     Simple utility to load the SL file and find the rank of each primary that Exomiser actually ranked; None for any it didn't
     @param sl - the identifier for the patient
     @param foundPrimaries - list of primaries in tuple form (chrom, position, ref allele, alt allele)
     @return - a list of their 0-based rank from the Exomiser JSON output
     '''
-    fullRanks = loadAllExomiserRanks(sl)
+    fullRanks = loadAllExomiserRanks(sl, subtype)
     ret = []
-    for fp in foundPrimaries:
+    for fpRaw in foundPrimaries:
+        #X is changed to 23 for some reason
+        if fpRaw[0] == 'X':
+            fp = ('23', )+fpRaw[1:]
+        else:
+            fp = fpRaw
+        
         try:
             ind = fullRanks.index(fp)
         except:
             assert(len(fullRanks) != 0)
             ind = -1*len(fullRanks)
+            
+            for fr in fullRanks:
+                print('fr', fr)
+            print('ERROR: missing reported variant in Exomiser output')
+            print('Missing:', fp)
+            raise Exception('See above')
+            
         ret.append(ind)
     return ret
 

### 2019-03-13 scripts/PVPUtil.py

@@ -1,0 +1,64 @@
+'''
+This file should contain helper functions to parse and load Exomiser ranking files
+'''
+
+import os
+
+def loadAllPVPRanks(sl):
+    '''
+    Loads a .res file from PVP
+    @param sl - the SL number to load
+    '''
+    #fn = '/Users/matt/githubProjects/VarSight/exomiser_results/'+sl+'-hiphive-genome-PASS_ONLY.json'
+    fn = '/Users/matt/githubProjects/VarSight/pvp_results/%s_results.txt.res' % (sl, )
+    fp = open(fn, 'rt')
+
+    #first line is the header
+    header = fp.readline().rstrip()
+    expectedHeader = ['Chr', 'Start', 'Ref', 'Alt', 'GT', 'Gene', 'CADD', 'GWAVA', 'DANN', 'Sim_Score', 'Prediction_Score']
+    #print(header.split('\t'))
+    #print(expectedHeader)
+    assert(header.split('\t') == expectedHeader)
+
+    #the file is a little weird, ranks are done by gene first, and the variants are a sub-group inside the gene
+    orderedRet = []
+    for l in fp:
+        pieces = l.split('\t')
+        chrom = pieces[0]
+        pos = pieces[1]
+        ref = pieces[2]
+        alt = pieces[3]
+        orderedRet.append((chrom, pos, ref, alt))
+    
+    fp.close()
+    
+    return orderedRet
+
+def getTargetRanks(sl, foundPrimaries):
+    '''
+    Simple utility to load the SL file and find the rank of each primary that Exomiser actually ranked; None for any it didn't
+    @param sl - the identifier for the patient
+    @param foundPrimaries - list of primaries in tuple form (chrom, position, ref allele, alt allele)
+    @return - a list of their 0-based rank from the Exomiser JSON output
+    '''
+    fullRanks = loadAllPVPRanks(sl)
+    ret = []
+    for fp in foundPrimaries:
+        try:
+            ind = fullRanks.index(fp)
+        except:
+            assert(len(fullRanks) != 0)
+            ind = -1*len(fullRanks)
+            
+            for fr in fullRanks:
+                print('fr', fr)
+            print('ERROR: missing reported variant in PVP output')
+            print('Missing:', fp)
+            raise Exception('See above')
+            
+        ret.append(ind)
+    return ret
+
+if __name__=='__main__':
+    #for testing purposes
+    pass

### 2019-03-13 scripts/PhenGenUtil.py

@@ -1,0 +1,90 @@
+'''
+This file should contain helper functions to parse and load Exomiser ranking files
+'''
+
+import os
+import vcf
+
+def loadAllPhenGenRanks(sl):
+    '''
+    Loads the output from PhenGen
+    @param sl - the SL number to load
+    '''
+    geneFN = '/Users/matt/githubProjects/VarSight/phengen_results/%s_hpo_%s_results_genescores.txt' % (sl, sl)
+    vcfFN = '/Users/matt/githubProjects/VarSight/phengen_results/%s_hpo_%s_results_variants_for_topgenes.vcf' % (sl, sl)
+
+    fp = open(geneFN, 'rt')
+
+    #first line is the header
+    header = fp.readline().rstrip()
+    expectedHeader = ['#GENE_ID', 'PROBABILITY_DAMAGING']
+    assert(header.split('\t') == expectedHeader)
+    
+    gDict = {}
+    for l in fp:
+        pieces = l.rstrip().split('\t')
+        geneName = pieces[0]
+        score = float(pieces[1])
+        gDict[geneName] = score
+    fp.close()
+
+    #now go through the VCF
+    vcfReader = vcf.Reader(filename=vcfFN, compressed=False)
+    unsorted = []
+    for variant in vcfReader:
+        chrom = str(variant.CHROM)
+        pos = str(variant.POS)
+        ref = variant.REF
+        alt = str(variant.ALT[0])
+        #print(chrom, pos, ref, alt, variant.INFO)
+        geneList = variant.INFO['GNID']
+        for i, g in enumerate(geneList):
+            if g.endswith('_neighboring'):
+                geneList[i] = g[:-len('_neighboring')]
+        dcod = float(variant.INFO.get('DCOD', 0.0)) #can be absent if non-coding
+        dreg = float(variant.INFO['DREG'])
+        
+        maxGeneVal = max([gDict[g] for g in geneList])
+        #print(maxGeneVal, dcod, dreg, chrom, pos, ref, alt)
+        unsorted.append((maxGeneVal, dcod, dreg, chrom, pos, ref, alt))
+    
+    #sorting by gene, then DCOD, then DREG (bigger is more deleterious for all of them)
+    #for v in unsorted:
+    #    print(v)
+    unsorted.sort(reverse=True)
+
+    #now pull out the part we care about
+    orderedRet = []
+    for e in unsorted:
+        orderedRet.append(e[3:])
+    
+    return orderedRet
+
+def getTargetRanks(sl, foundPrimaries):
+    '''
[…]

### 2019-03-13 scripts/TestLearners.py

@@ -14,6 +14,8 @@ import CodiDumpUtil
 import ExomiserUtil
 import HPOUtil
 import OntologyUtil
+import PhenGenUtil
+import PVPUtil
 import PyxisMapUtil
 import SummaryDBUtil
 
@@ -23,9 +25,11 @@ from sklearn.decomposition import PCA
 from sklearn.ensemble import RandomForestClassifier
 from sklearn.linear_model import LogisticRegression
 from sklearn.metrics import auc
+from sklearn.metrics import average_precision_score
 from sklearn.metrics import balanced_accuracy_score
 from sklearn.metrics import confusion_matrix
 from sklearn.metrics import f1_score
+from sklearn.metrics import make_scorer
 from sklearn.metrics import precision_recall_curve
 from sklearn.metrics import roc_curve
 from sklearn.model_selection import cross_validate
@@ -120,13 +124,13 @@ def loadFormattedData(args):
     '''
     This function is primarily about loading and formatting data prior to training/testing any classifiers.
     @param args - arguments from the command line
-    @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts, exomiserRanks)
+    @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts, externalRanks)
         xFinal - an NxM matrix where N is the number of variants in our test/training set and M is the number of features; contains feature values
         yFinal - an N length array where N is the number of variants in our test/training set; 1 if the variant was reported
         featureLabels - an M length array containing feature labels for our output benefit
         startIndices - a (C+1) length array containing the start indices of variants from an individual case
         allRepDicts - an N length array of None or dictionaries; if a dictionary, it contains data on a returned variant
-        exomiserRanks - a list of lists containing exomiser ranks on a per-case basis; missing values are -1*len(ranked variants)
+        externalRanks - a dictionary where key is a label and value is a list of lists containing exomiser ranks on a per-case basis; missing values are -1*len(ranked variants)
     '''
     pyxisRootDir = '/Users/matt/githubProjects/VarSight/pyxis_ranks_'+PYXIS_DATE
     
@@ -143,8 +147,8 @@ def loadFormattedData(args):
     catLabels = []
     
     #treat all categorical things the same other than where it is sourced from
-    labelBlocks = ['allelicInformation', 'genes', 'variantTranscripts']
-
[…]

### 2019-03-13 paper/data_template.tex
updating with some feedback

--- 2019-01-15 paper/data_template.tex
@@
 
-Exomiser$^*$&
+{% for ext_label in EXT_LABELS %}
+    {{ ext_label.replace('_', '\\_') }}&
     {% for LEVEL in LEVELS %}
-    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER.TEST_RANKINGS[LEVEL])) }}&
+        {{ '%0.1f (%0.2f)' % (np.median(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]), np.mean(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL])) }}&
     {% endfor %}
@@
         {{ '%d, %d, %d' % 
-        (100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
-        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
-        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER.TEST_RANKINGS[LEVEL]))
+            (100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 1) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 10) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]),
+            100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 20) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]))
         }}
@@
     {% endfor %}
-\\ \hhline{=|=|=|=|=||=|=|=|=}
+    \\
+{% endfor %}
+\hhline{=|=|=|=|=||=|=|=|=}
 

### 2019-03-13 paper/main.tex
removing date so it doesn't act crazy later
updating with some feedback

@@
 \newif\ifpreprint
-\preprinttrue
+%\preprinttrue
 
@@
 \affil[1]{Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA}
+\date{}
 \maketitle
@@
 Variant prioritization is also quite diverse with some methods relying on the variant annotations to prioritize variants \citep{hu2013} and some relying on patient phenotype to rank the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.
-There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.
+There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a, boudellioua2019}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.
 
@@
 As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases, phenotypes, and genetic heritage tied to UDN patients.
+Additionally, we made no effort to alter or condense the set of phenotypes provided by the corresponding clinical sites.
 In order to incorporate patient phenotype information, we used two distinct methods to rank genes based on the Human Phenotype Ontology (HPO) \citep{kohler2018}.
@@
 
-The first method uses base annotations provided by the HPO to calculate a simple cosine score \citep{kohler2017} between the patient's phenotypes and each gene.
+The first method uses phenotype-to-gene annotations provided by the HPO to calculate a cosine score \citep{kohler2017} between the patient's phenotypes and each gene.
+In short, this method builds a vector from the phenotype terms (and ancestor terms in the ontology) for both the patient and each gene.
+Then, the cosine between the patient vector and each gene vector is calculated as a representation of similarity.
 This method tends to be more conservative because it relies solely on curated annotations from the HPO.
-The second method, an internally-developed tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.
+
+The second method, an internally-developed tool called PyxisMap \citep{wilk2018}, uses the same phenotype-to-gene annotations from the HPO, but adds in automatically text-mined annotations from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.
 The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.
@@
 
-We then selected four readily available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.
-For each model, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.
+We then selected four publicly available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.
+The primary reason for selecting these four classifiers is that they are all capable of handling imbalanced data either by default or with a single parameter specifying to balance the data during training.
+The two random forest models build collections of decision trees that weight each training input by its class frequency.
+Logistic regression calculates the probability of a value belonging to a particular class, again weighting by the class frequency.
+In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.
+
+Initially, we also tried the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.
+Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.
+While we did not achieve good results using these out-of-the-box implementations, it may be possible to use the methods with appropriate changes made to the underlying implementations.
+
+For each of our tested models, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.
 For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e.
@@
 For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.
-We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}).
-%The values for each model are shown in Table \ref{tab:classifier_results}.  
-Figure \ref{fig:roc} show receiver operator curves for FPR v.
-TPR and recall v.
-precision for the four models.
-%Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.
+We then calculated the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC) using the unseen test set (see Table \ref{tab:classifier_results}).
+Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the four models.
 
 From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.
-%All four classifiers have higher AUROC than either single-measure statistic.  
 However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was 0.2109).
@@
 \centering
-\begin{tabular}{c|c|c|c|c}
+\begin{tabular}{c|c|c|c}
 \hline
@@
 \caption{Classifier performance statistics.
-For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC).
+For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC).
 In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.}
@@
 \processtable{Classifier performance statistics.
-For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC).
+For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC).
 In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.
 \label{tab:classifier_results}}{
-\begin{tabular}{c|c|c|c|c}
+\begin{tabular}{c|c|c|c}
 \hline
@@
 CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.
-Exomiser is an external tool that only reported ranks for a subset of the filtered variants.
+%Exomiser is an external tool that only reported ranks for a subset of the filtered variants.
 ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.
@@
 middle) for all unranked variants.
+The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.
+Phen-Gen was the only external tool that did not rank every variant in the set, so we conservatively assumed unranked variants were at the next best position despite being unranked.
 The bottom four rows are the tuned, binary classification methods tested in this paper.
@@
 The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.
-All values were generated using only the test data that was unseen during training.}
+All values in this table were generated using only the test data that was unseen during training.}
 \label{tab:performance}
@@
 The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.
-All values were generated using only the test data that was unseen during training.
+All values in this table were generated using only the pre-filtered variants from testing data.
 \label{tab:performance}}{
@@
 
-For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).
-We followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.
-For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.
-We then parsed the output JSON file created by Exomiser into a ranked order of variants.
+For comparison, we selected to run Exomiser \citep{smedley2015a}, Phen-Gen \citep{javed2014}, and DeepPVP \citep{boudellioua2019}.
+For each tool, we input the exact same set of phenotype terms used by the classifiers we tested.
+Additionally, we used the pre-filtered variant set from Codicem as input to each ranking algorithm.
+As a result, all external tools and our trained classifiers are ranking on identical phenotype and variant information.
 
-Unfortunately, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 10-50 variants from the pre-filtered VCF.
-For the reported variants, Exomiser did not rank 98 of the 189 variants in the test set (58 variant of uncertain significance, 21 likely pathogenic, and 19 pathogenic).
-For these unranked variants, we assigned a rank using two different methods.
-The ``Exomiser(best)" method conservatively assumed that unranked primary variants were found at the next best position in the rank order (e.g.
-if Exomiser ranked 15 variants, we assumed all unranked variants were ranked 16th).
-In contrast, the ``Exomiser(avg)" method assumed that unranked primary variants were ranked at the average position for all unranked variants (e.g.
-if Exomiser ranked 10 of 100 variants, we assumed all unranked variants were ranked 55th).
-``Exomiser(best)" represents the best possible ordering that Exomiser could give us for unranked variants where ``Exomiser(avg)" is a more realistic expectation.
+For Exomiser, we followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.
+We ran Exomiser twice, once using the default hiPhive prioritizer (incorporates knowledge from human, mouse, and fish) and once using the human only version of the hiPhive prioritizer (this was recommended in place of the deprecated PhenIX algorithm \citep{zemojtel2014}).
+Phen-Gen V1 was run using the pre-compiled binary using the ``dominant'' and ``genomic'' modes to maximize the output.
+Of note, Phen-Gen was the only external method that did not fully rank all variants, so we conservatively assumed that any absent variants were at the next best possible rank.
+Thus, the reported Phen-Gen comparisons are an optimistic representation for this test data.
+Finally, DeepPVP v2.1 was run using the instructions available on their website.
+Details on the exact execution for each external tool can be found in the Supplementary Documents (TODO: XXX).
 
@@
 These scores were inputs to each classifier, but also represent two common ways one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).
-The results for the two control scores, both Exomiser approaches, and all four classifiers are shown in Table \ref{tab:performance}.
+The results for the two control scores, all four external tools, and all four trained classifiers are shown in Table \ref{tab:performance}.
 
-In the overall data, all four classifiers outperform the single-measure statistics and both Exomiser approaches across the board, with the one exception being the mean of ``Exomiser(best)" (a measure that is heavily biased by the conservative ranking of unranked variants).
+In the overall data, all four classifiers outperform the single-value measures and external tools across the board.
 As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.
@@
 \subsection{Random Forest Feature Importances}
-After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
-Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).
-The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.
-By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.
-Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.
-
-Interestingly, the two strongest individual features by far were the phenotype-based measures (HPO-cosine and PyxisMap) making up $\sim$33-40\% of the feature importance.
-These were followed by CADD Scaled ($\sim$7-10\%) and three conservation scores ($\sim$14\%).
-GnomAD-related fields ($\sim$8\%), GERP rsScore ($\sim$2-3\%), and HGMD-related fields ($\sim$4\%) made up the rest of features with higher than expected importance.
-These results would suggest that phenotype-based metrics, deleteriousness scores, and conservation scores are of very high importance for predicting if a variant will be clinically returned to a rare disease patient.
-
 \ifpreprint
@@
 \fi
+
+After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
+Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).
+The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.
+By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.
+Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.
+
+The two strongest individual features by far were the phenotype-based measures (HPO-cosine and PyxisMap) making up $\sim$33-40\% of the feature importance.
+This is not a major surprise given that phenotype information was not used during the pre-filtering steps provided by Codicem.
+These were followed by CADD Scaled ($\sim$7-10\%) and three conservation scores ($\sim$14\%).
+These results would suggest that phenotype-based metrics, deleteriousness scores, and conservation scores are of very high importance for the classifiers we tested.
 
@@
 
+%DeepPVP - variant ranker
+\bibitem[Boudellioua {\it et~al}., 2019]{boudellioua2019} Boudellioua, Imane, \textit{et~al.} "DeepPVP: phenotype-based prioritization of causative variants using deep learning." {\it BMC bioinformatics} 20.1 (2019): 65.
+
 %PROVEAN - variant-transcript prediction
@@
 \bibitem[Jäger {\it et~al}., 2014]{jager2014} Jäger, Marten, \textit{et~al.} "Jannovar: A Java Library for Exome Annotation." {\it Human mutation} 35.5 (2014): 548-555.
+
+%Phen-Gen - variant ranker
+\bibitem[Javed {\it et~al}., 2014]{javed2014} Javed, Asif, Saloni Agrawal, and Pauline C.
+Ng.
+"Phen-Gen: combining phenotype and genotype to analyze rare disorders." {\it Nature methods} 11.9 (2014): 935.
 

### 2019-03-16 scripts/TestLearners.py
reverted to OHE-encoding and feature selection, seems to work well

@@ -23,6 +23,9 @@ from imblearn.ensemble import BalancedRandomForestClassifier, RUSBoostClassifier
 
 from sklearn.decomposition import PCA
 from sklearn.ensemble import RandomForestClassifier
+from sklearn.feature_selection import SelectKBest
+from sklearn.feature_selection import mutual_info_classif
+from sklearn.feature_selection import RFE
 from sklearn.linear_model import LogisticRegression
 from sklearn.metrics import auc
 from sklearn.metrics import average_precision_score
@@ -181,93 +184,22 @@ def loadFormattedData(args):
                 raise Exception("Unknown interpretation: %s" % (d['interpret'], ))
 
     #These are the fields used from CODICEM dumps
-    '''
-    floatValues = [
-        ('CADD Scaled', -1.0),
-        ('phylop conservation', -30.0),
-        ('phylop100 conservation', -30.0),
-        ('phastcon100 conservation', -1.0),
-        ('Mappability', -1.0),
-        ('GERP rsScore', -13.0),
-        ('Gnomad Exome AF', -1.0),
-        ('Gnomad Exome Hom alt allele count', -1.0),
-        ('Gnomad Exome Hemi alt allele count', -1.0),
-        ('Gnomad Exome total allele count', -1.0),
-        ('Gnomad Genome AF', -1.0),
-        ('Gnomad Genome Hom alt allele count', -1.0),
-        ('Gnomad Genome Hemi alt allele count', -1.0),
-        ('Gnomad Genome total allele count', -1.0),
-    ]
-    '''
     floatValues = floatDict['allelicInformation']
     fieldsOnly = set([l for l, d in floatValues]) | set([d['key'] for d in catMeta['allelicInformation']])
     
     #(label, reduce function, default)
-    '''
-    geneValues = [
-        ('RVIS Score', min, 10.0),
-        ('GHIS Score', max, -10.0),
-        ('HIS Score', max, -10.0)
-    ]
-    '''
     geneValues = floatDict['genes']
     geneFieldsOnly = set([l for l, r, d in geneValues]) | set([d['key'] for d in catMeta['genes']])
     
     #(label, default)
-    '''
-    seqValues = [
-        ("percent reads", -1.0),
-        ("total depth", -1.0)
-    ]
-    '''
     seqValues = floatDict['sequencingFields']
     seqFieldsOnly = set([l for l, d in seqValues]) | set(['chromosome', 'position', 'ref allele', 'alt allele'])
     
     #numerical variant-transcript information
[…]

### 2019-03-17 scripts/CodiDumpUtil.py
beta version of standalone VarSight script for running on trained models

@@ -3,9 +3,14 @@ This file should contain helper functions to parse and load CODICEM filter files
 '''
 
 import json
+import numpy as np
 import os
+import re
 
-def loadCodiDump(sl, EXTRA_FIELDS, GENE_FIELDS, SEQ_FIELDS, TRANSCRIPT_FIELDS):
+import HPOUtil
+import PyxisMapUtil
+
+def loadCodiDump(codiDumpFN, EXTRA_FIELDS, GENE_FIELDS, SEQ_FIELDS, TRANSCRIPT_FIELDS):
     '''
     This file should open the latest CODI dump file from a pre-determined directory and load particular fields to be returned
     @return - a list of variants:
@@ -19,13 +24,13 @@ def loadCodiDump(sl, EXTRA_FIELDS, GENE_FIELDS, SEQ_FIELDS, TRANSCRIPT_FIELDS):
     ]
     '''
     #make sure the file exists
-    fn = '/Users/matt/githubProjects/VarSight/CLI_primary_V6/'+sl+'_results.json'
-    if not os.path.exists(fn):
+    #fn = '/Users/matt/githubProjects/VarSight/CLI_primary_V6/'+sl+'_results.json'
+    if not os.path.exists(codiDumpFN):
         #print('Missing file: '+fn)
         return []
     
     #load the json file
-    fp = open(fn, 'rt')
+    fp = open(codiDumpFN, 'rt')
     j = json.load(fp)
     fp.close()
     
@@ -42,7 +47,7 @@ def loadCodiDump(sl, EXTRA_FIELDS, GENE_FIELDS, SEQ_FIELDS, TRANSCRIPT_FIELDS):
         vDict = {}
         
         #the higher this number is, the bigger it is in the circle pack
-        variantOrderedRank = variant['variantOrderedRank']
+        #variantOrderedRank = variant['variantOrderedRank']
         
         #change this to the format we recognize from the table
         vd = variant['variantDisplayName']
@@ -94,3 +99,224 @@ def loadCodiDump(sl, EXTRA_FIELDS, GENE_FIELDS, SEQ_FIELDS, TRANSCRIPT_FIELDS):
         ret.append(vDict)
         
     return ret
+
+def prepDataFromHpo(catMeta, codiDumpFN, hpoTerms):
+    '''
+    This method is intended to be used by VarSight.py. As such, this is basically a wrapper around prepDataShared(...)
+    such that code isn't duplicated all over the place.
[…]

### 2019-03-17 scripts/PyxisMapUtil.py

@@ -7,7 +7,7 @@ import requests
 
 import pronto
 
-def getPyxisMapResults(hpoTerms):
+def getPyxisMapResults(hpoTerms, pyxisRoot='http://0.0.0.0:4999'):
     '''
     This function hits the PyxisMap /rank endpoint to get a single ranking of the genes
     @param hpoTerms - a set of core HPO terms of interest
@@ -16,7 +16,7 @@ def getPyxisMapResults(hpoTerms):
         number of genes ranked
     '''
     #TODO: make this URL configurable?
-    pyxisURL = 'http://0.0.0.0:4999/rank'
+    pyxisURL = pyxisRoot+'/rank'
     headers = {'Content-type': 'application/json'}
     data_json = json.dumps(sorted(hpoTerms))
     

### 2019-03-17 scripts/TestLearners.py

@@ -8,6 +8,7 @@ import json
 import matplotlib.pyplot as plt
 import numpy as np
 import os
+import pickle
 import re
 
 import CodiDumpUtil
@@ -130,6 +131,7 @@ def loadFormattedData(args):
     @return - tuple (xFinal, yFinal, featureLabels, startIndices, allRepDicts, externalRanks)
         xFinal - an NxM matrix where N is the number of variants in our test/training set and M is the number of features; contains feature values
         yFinal - an N length array where N is the number of variants in our test/training set; 1 if the variant was reported
+        catMeta - the categorical metadata 
         featureLabels - an M length array containing feature labels for our output benefit
         startIndices - a (C+1) length array containing the start indices of variants from an individual case
         allRepDicts - an N length array of None or dictionaries; if a dictionary, it contains data on a returned variant
@@ -137,67 +139,11 @@ def loadFormattedData(args):
     '''
     pyxisRootDir = '/Users/matt/githubProjects/VarSight/pyxis_ranks_'+PYXIS_DATE
     
-    #load the categorical metadata
+    #load the feature metadata
     metaFN = '/Users/matt/githubProjects/VarSight/CODI_metadata/fields_metadata.json'
     fp = open(metaFN, 'rt')
     catMeta = json.load(fp)
     fp.close()
-
-    #build some lookup dicts
-    indexDict = {}
-    indexLen = {}
-    ignoreDict = {}
-    catLabels = []
-    
-    #treat all categorical things the same other than where it is sourced from
-    labelBlocks = ['allelicInformation', 'genes', 'sequencingFields', 'variantTranscripts']
-    floatDict = {k : [] for k in labelBlocks}
-    for lg in labelBlocks:
-        for d in catMeta[lg]:
-            #make sure we never get the same category twice (shouldn't happen)
-            assert(d['key'] not in indexDict)
-
-            if d['interpret'] == 'single':
-                #singles are easy
-                catLabels.append(d['key'])
-            elif d['interpret'] == 'multiple':
-                #multiples, store the number of possiblities in indexLen and the index in indexDict
-                indexDict[d['key']] = {}
-                indexLen[d['key']] = len(d['accepted'])
-                for i, v in enumerate(sorted(d['accepted'])):
-                    #we are .lower()-ing everything, make sure that doesn't introduce a conflict
-                    assert(v not in indexDict[d['key']])
-                    indexDict[d['key']][v.lower()] = i
-                ignoreDict[d['key']] = set([v.lower() for v in d.get('ignored', [])])
[…]

### 2019-03-17 scripts/VarSight.py

@@ -1,0 +1,114 @@
+
+import argparse as ap
+import numpy as np
+import os
+import pickle
+
+import CodiDumpUtil
+
+def loadClassifier(args):
+    '''
+    @param args - the argparse values
+    @return - tuple
+        classifier - the trained model from the pickle file
+        catMeta - the categorical metadata for all possible features
+        features - the features as ordered in the model
+    '''
+    #pull out any args we care about
+    modelsFN = args.modelsFN
+    modelName = "LogisticRegression(sklearn)" #TODO: make a parameter
+    
+    #load the pickle file
+    fp = open(modelsFN, 'rb')
+    modelDict = pickle.load(fp)
+    fp.close()
+
+    allowedModels = modelDict['MODELS'].keys()
+    if modelName not in allowedModels:
+        raise Exception('No model named "%s" in "%s"' % (modelName, modelsFN))
+    classifier = modelDict['MODELS'][modelName]
+    features = modelDict['FEATURES']
+    catMeta = modelDict['CATEGORICAL_METADATA']
+
+    return (classifier, catMeta, features)
+
+def loadHpoTerms(fn):
+    '''
+    @param fn - a text file with one HPO term per line in HP:####### format
+    @return - a set containing all HPO terms in the file
+    '''
+    hpoTerms = set([])
+    fp = open(fn, 'r')
+    for l in fp:
+        t = l.rstrip()
+        hpoTerms.add(t)
+    fp.close()
+    return hpoTerms
+
+def extractRelevantFeatures(xValuesFull, featureLabelsFull, features):
+    '''
+    @param xValuesFull - the full list of features
+    @param featureLabelsFull - the labels for each column
+    @param features - the features we want to extract
+    @return - an array containing only the features we care about (aka, fewer columns)
+    '''
+    columnsKeep = []
+    for f in features:
+        columnsKeep.append(featureLabelsFull.index(f))
+    
+    colArr = np.array(columnsKeep)
+    return xValuesFull[:, colArr]
+
+def runPredictions(classifier, xFeatureVals):
+    '''
+    @param classifier - the classifier to run
+    @param xFeatureVals - the feature values to test on
+    '''
+    probs = classifier.predict_proba(xFeatureVals)[:, 1]
+    ordered = np.argsort(probs)[::-1]
+    return (probs, ordered)
+
+if __name__ == "__main__":
+    #first set up the arg parser
+    DESC = "VarSight - prioritizing Codicem filtered variants using trained classifiers"
[…]

### 2019-03-18 scripts/HPOUtil.py
fixed a comment mistake

@@ -59,7 +59,9 @@ def loadHPOAnnot(fn, hpoOnt):
     This will load the HPO annotations file
     @param fn - the file to load, expected is ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt
     @param hpoOnt - the MicaDS hpoOnt used for filtering out things that are absent
-    @return - dictionary where key is a gene name and value is a set of associated MONDO terms
+    @return - tuple (g2p, p2g)
+        g2p - dictionary where key is a gene symbol and value is a phenotype set
+        p2g - dictionary where key is a HPO ID and value is a gene set
     '''
     g2p = {}
     p2g = {}

### 2019-03-18 paper/main.tex
finalizing changes based on feature selection

@@
 We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls.
-We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
+We showed that these classifiers outperformed the other methods with the best classifiers ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
 \\
@@
 % to show how the classifiers tend to outperform the other methods.
-We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
+We showed that these classifiers outperformed the other methods with the best classifier ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
 %this was probably too detailed, need a different sentence
@@
 Given a prioritized list of variants, analysts manually inspect those variants and curate a list of variants to ultimately report to the ordering physician.
-Unfortunately, manual curation is time consuming and exhausting.
-An analyst must inspect each variant and the associated metadata while simultaneously maintaining a mental picture of the patient's phenotype, leading to what we colloquially refer to as ``variant fatigue".
-This variant fatigue means that variants at the end of the prioritized list are often regarded with less scrutiny and/or a less accurate patient model than those near the beginning.
-Methods that can prioritize these variants accurately can reduce the impact of variant fatigue, reducing the chances that variants are overlooked or mis-identified.
-Additionally, if causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.
-Finally, accurate prioritization is a step towards the ultimate goal of automatically identifying all variants that cause a patient's primary phenotypes.
+Unfortunately, manual curation is a time consuming process where analysts must inspect each variant while maintaining a mental picture of the patient's phenotype.
+\citealp{bick2017} report an average of 600 variants per case analyzed by two people (one analyst and one director) over three hours ($\approx$100 variants per man-hour).
+If causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.
+Finally, accurate prioritization is a step towards the ultimate goal of automatic identification of rare variants that may cause a patient's primary phenotypes.
 
 One of the issues with previously published ranking methods is that they were primarily tested on simulated datasets with known, single-gene, pathogenic variants injected into real or simulated background genomic datasets.
-Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few noisy terms added or removed.
+Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few (typically 2-3) noisy terms added or removed.
+Unfortunately, rare disease patient data is often far noisier due to a wide variety of reasons.
 
@@
 Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as phenotypes caused by old age).
-%Finally, analysts have the discretion to send back variants with different levels of pathogenicity as specified by the ACMG guidelines \citep{richards2015}.  
 Because the UDN is a research collaboration, there is also variability in reported variants that range in pathogenicity from ``variant of uncertain significance" (VUS) through ``pathogenic" as defined by the ACMG guidelines \citep{richards2015}.
@@
 \section{Approach}
-In this paper, we tested the application of classification algorithms for two purposes: 1) predicting whether a variant observed by an analyst would be clinically reported and 2) prioritizing all variants seen by clinical analysts.
+We tested the application of classification algorithms for identifying clinically reported variants in real world patient in two ways: 1) predicting whether a variant observed by an analyst would be clinically reported and 2) prioritizing all variants seen by the clinical analysts.
 In particular, we focused our analyses on real patients with a diverse collection of rare, undiagnosed diseases that were admitted to the Undiagnosed Diseases Network (UDN) \citep{ramoni2017}.
@@
 We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.
-Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing for both classification analysis and ranking of the variants by their calculated probabilities.
+Of note, our focus was not on picking the ``best'' classifier, but rather on analyzing their overall ability of classification algorithms to handle the noise of real-world patient cases from the UDN.
+
+Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing us to measure their performance as a classifier and as a ranking system.
 After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant ranking perspective.
+Additionally, we tested four publicly available variant ranking algorithms and two single-value ranking methods for comparison.
 All of the scripts to train classifiers, test classifiers, and format results are contained in the VarSight repository.
@@
 In short, this method builds a vector from the phenotype terms (and ancestor terms in the ontology) for both the patient and each gene.
-Then, the cosine between the patient vector and each gene vector is calculated as a representation of similarity.
+Then, the cosine of the angle between the patient vector and each gene vector is calculated as a representation of similarity.
 This method tends to be more conservative because it relies solely on curated annotations from the HPO.
@@
 The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.
-We used PyxisMap v1.2, and we ran the standard installation script that downloads all required data sources on December 19, 2018.
+%We used PyxisMap v1.2, and we ran the standard installation script that downloads all required data sources on December 19, 2018. 
 Each method generates a single numerical feature that is used in the following analyses.
@@
 \subsection{Patient selection}
-In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believed the variant to be directly related to the patient's phenotype.
+In the clinical analysis, each patient was fully analyzed by one director and one analyst.
+After the initial analysis, the full team of directors and analysts review flagged variants and determine their reported pathogenicity.
+In our analysis, we focused on variants that were clinically reported as ``primary", meaning the team of analysts believed the variant to be directly related to the patient's phenotype.
 Note that secondary and/or incidental findings are specifically not included in this list.
-The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic adhering to the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.
+The team of analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic adhering to the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.
 
@@
 
-\subsection{Data cleaning}
+\subsection{Feature Selection}
+\ifpreprint
+\begin{table}[!t]
+\centering
+\begin{tabular}{l|c|c}
+\hline
+\input{features_rendered.tex}
+\hline
+\end{tabular}
+\caption{Feature selection.
+This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.
+After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).
+We average the two importance values, and order them from most to least important.
+Feature labels with an `\_' represent a single category of a multi-category feature (i.e.
+``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem).}
+\label{tab:features}
+\end{table}
+
+\else
+
+\begin{table}[!t]
+\processtable{Feature selection.
+This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.
+After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).
+We average the two importance values, and order them from most to least important.
+Feature labels with an `\_' represent a single category of a multi-category feature (i.e.
+``HGMD assessment type\_DM'' means the ``DM'' bin-count feature from the ``HGMD assessment type'' annotation in Codicem).
+\label{tab:features}}{
+\begin{tabular}{l|c|c}
+\hline
+\input{features_rendered.tex}
+\hline
+\end{tabular}}{}
+\end{table}
+\fi
+
 For the purposes of classification, all annotations needed to be cleaned and stored as numerical features.
-For numerical annotations (e.g.
+For single-value numerical annotations (e.g.
 float values like CADD or GERP), we simply copied the annotation over as a single value feature.
@@
 The one exception to this rule was for variant allele frequencies where a variant absent from the database was considered to have an allele frequency of 0.0.
+For multi-value numerical annotations, we reduced the values (using minimum or maximum) to a single value that was used as the feature.
 
-For categorical data, we used a two step approach to cleaning the data: bin-count encoding and principal component analysis.
+%this was the old way before feature selection
+%For categorical data, we used a two step approach to cleaning the data: bin-count encoding and principal component analysis.
 First, we chose to use a bin-count because there are many categories where multiple categorical labels may be present at different quantities.
@@
 To reduce this dilution, we used principal component analysis \citep{jolliffe2011} to reduce the dimensions of each category and stored at most two features per categorical feature.
-%For specifics on variance explained by the top two PCA components per category, please see Supplementary document X (If we ever do this, it will be at the request of a user or reviewers, it doesn't seem super important at this juncture)
+The specific features used and the feature types are detailed in the Supplementary Documents.
+
+For categorical data, we relied on bin-count encoding to store the features.
+We chose to bin-count because there are many annotations where multiple categorical labels may be present at different quantities.
+For example, a single ClinVar variant may have multiple entries where different sites have selected different levels of pathogenicity.
+In this situation, we desired to capture not only the categorical label as a feature, but also the number of times that label occurred in the annotations.
+
+After converting all annotations to numerical features or multiple bin-count encoded features, we had a total of 95 features per variant.
+We then pruned down to only the top 20 features using univariate feature selection (specifically the \texttt{SelectKBest} method of \texttt{sklearn}).
+This method evaluates how well an individual feature performs as a classifier and keeps only the top 20 features for the full classifiers.
+Table \ref{tab:features} shows the list of retained features ordered by feature importance after training.
+Feature importance was derived from the random forest models which automatically report how important each feature was to classification.
+The entire set of annotations along with descriptions of how each was processed prior to feature selection are detailed in the Supplementary Documents.
 
@@
 We then selected four publicly available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.
-The primary reason for selecting these four classifiers is that they are all capable of handling imbalanced data either by default or with a single parameter specifying to balance the data during training.
+These classifiers were chosen for three main reasons: 1) their ability to handle imbalanced data (i.e.
+far more unreported variants than reported variants), 2) their ability to scale to the size of the training and testing datasets, and 3) they are freely available implementations that can be tuned and tested with relative ease.
 The two random forest models build collections of decision trees that weight each training input by its class frequency.
@@
 In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.
+For more details on each classifier, please refer to the {\it sklearn} and {\it imblearn} documentations.
 
-Initially, we also tried the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.
+Initially, we also tested the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.
 Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.
-While we did not achieve good results using these out-of-the-box implementations, it may be possible to use the methods with appropriate changes made to the underlying implementations.
+While we did not achieve positive results using these three implementations, it may be possible to use the methods through another implementation.
 
 For each of our tested models, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.
-For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e.
-a weighted accuracy score where inputs are weighted by their inverse class frequency).
-For each model type, we saved the hyperparameters and model with the best average balanced accuracy score.
-These four tuned models were then tested against the unseen set of test proband cases.
+For each model and set of hyperparameters, we performed stratified 10-fold cross validation on the training variants and recorded the balanced accuracy (i.e.
+weighted accuracy based on inverse class frequency) and the F1 scores (i.e.
+harmonic mean between precision and recall).
+For each model type, we saved the hyperparameters and model with the best average F1 score.
+These four tuned models were then trained on the full training set and tested against the unseen set of test proband cases.
+The set of hyperparameters tested along with the highest performance setting for each hyperparameter can be found in the Supplementary Documents.
+
 \ifpreprint
@@
 }
-\caption{Receiver operator curves.  
-%This figure shows the receiver operator curves for the test data for each single-measure statistic and for each trained classifier.  
-This figure shows the receiver operator curves for the test data for each trained classifier.  
-On the left, we show the false positive rate against the true positive rate.
-On the right, we show the recall against the precision.
-Area under the receiver-operator curve (AUROC) is reported beside each method in the legend.
-In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly with LogisticRegression performing slightly worse overall.}
+\caption{Receiver operator and precision-recall curves.  
+These figures show the performance of the four classifiers on the testing set after hyperparameter tuning and fitting to the training set.
+On the left, we show the receiver operator curve (false positive rate against the true positive rate).
+On the right, we show the precision recall curve.
+Area under the curve (AUROC and AUPRC) is reported beside each method in the legend.}
 \label{fig:roc}
@@
 
-For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.
-We then calculated the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC) using the unseen test set (see Table \ref{tab:classifier_results}).
-Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the four models.
+Each model was tuned using 10-fold cross validation and the resulting average and standard deviation of balanced accuracy is reported in Table \ref{tab:classifier_results}.
+After fitting the tuned model to the full training set, we evaluated the models on the testing set by calculating the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC).
+Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the results from the testing set on all four models.
 
 From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.
-However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was 0.2109).
+However, these classifier all have relatively poor performance from a precision-recall perspective (best AUPRC was 0.2109).
 This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.
@@
 For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC).
-In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.}
+The CV10 Acc.
+was gathered during hyperparameter tuning by calculating the average and standard deviation of the 10-fold cross validation.
+AUROC and AUPRC was evaluated on the testing set after hyperparameter tuning and fitting to the full training set.}
 \label{tab:classifier_results}
@@
 For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), area under the receiver operator curve (AUROC), and area under the precision-recall curve (AUPRC).
-In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.
+The CV10 Acc.
+was gathered during hyperparameter tuning by calculating the average and standard deviation of the 10-fold cross validation.
+AUROC and AUPRC was evaluated on the testing set after hyperparameter tuning and fitting to the full training set.
 \label{tab:classifier_results}}{
@@
 CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.
-%Exomiser is an external tool that only reported ranks for a subset of the filtered variants.
-``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.
-In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e.
-middle) for all unranked variants.
 The middle four rows (two Exomiser runs, Phen-Gen, and DeepPVP) represent external tools that ranked the same set of variants as the classifier algorithms.
@@
 The bottom four rows are the tuned, binary classification methods tested in this paper.
-The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).
+The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (variant of uncertain significance (VUS), likely pathogenic (LP), or pathogenic (Path.)) derived from the ACMG guidelines.
 The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.
-All values in this table were generated using only the test data that was unseen during training.}
+All values in this table were generated using only the Codicem-filtered variants from testing set.}
 \label{tab:performance}
@@
 The bottom four rows are the tuned, binary classification methods tested in this paper.
-The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).
+The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (variant of uncertain significance (VUS), likely pathogenic (LP), or pathogenic (Path.)) derived from the ACMG guidelines.
 The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.
-All values in this table were generated using only the pre-filtered variants from testing data.
+All values in this table were generated using only the Codicem-filtered variants from testing set.
 \label{tab:performance}}{
@@
 For each tool, we input the exact same set of phenotype terms used by the classifiers we tested.
-Additionally, we used the pre-filtered variant set from Codicem as input to each ranking algorithm.
+Additionally, we used the same set of pre-filtered variants from Codicem as input to each ranking algorithm.
 As a result, all external tools and our trained classifiers are ranking on identical phenotype and variant information.
@@
 For Exomiser, we followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.
-We ran Exomiser twice, once using the default hiPhive prioritizer (incorporates knowledge from human, mouse, and fish) and once using the human only version of the hiPhive prioritizer (this was recommended in place of the deprecated PhenIX algorithm \citep{zemojtel2014}).
+We ran Exomiser twice, once using the default hiPhive prioritizer (incorporates knowledge from human, mouse, and fish) and once using the human only version of the hiPhive prioritizer (this was recommended instead of the PhenIX algorithm \citep{zemojtel2014}).
 Phen-Gen V1 was run using the pre-compiled binary using the ``dominant'' and ``genomic'' modes to maximize the output.
@@
 Finally, DeepPVP v2.1 was run using the instructions available on their website.
-Details on the exact execution for each external tool can be found in the Supplementary Documents (TODO: XXX).
+Details on the exact installation and execution for each external tool can be found in the Supplementary Documents.
 
@@
 In the overall data, all four classifiers outperform the single-value measures and external tools across the board.
-As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.
-Of the classifiers tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.
+As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases ranking 33-52\% of pathogenic variants in the first position and 80-94\% of pathogenic variants in the top 20.
 
-\subsection{Random Forest Feature Importances}
-\ifpreprint
-\begin{table}[!t]
-\centering
-\begin{tabular}{l|c|c}
-\hline
-\input{features_rendered.tex}
-\hline
-\end{tabular}
-\caption{Random forest feature importances.
-This table shows the top feature importances reported by the two random forest algorithms we tested.
-We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance.}
-\label{tab:features}
-\end{table}
-
-\else
-
-\begin{table}[!t]
-\processtable{Random forest feature importances.
-This table shows the top feature importances reported by the two random forest algorithms we tested.
-We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance.
-\label{tab:features}}{
-\begin{tabular}{l|c|c}
-\hline
-\input{features_rendered.tex}
-\hline
-\end{tabular}}{}
-\end{table}
-\fi
-
-After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
+%\subsection{Random Forest Feature Importances}
+%After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
 Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).
@@
 
-The two strongest individual features by far were the phenotype-based measures (HPO-cosine and PyxisMap) making up $\sim$33-40\% of the feature importance.
+%The two strongest individual features by far were the phenotype-based measures (HPO-cosine and PyxisMap) making up $\sim$33-40\% of the feature importance.
 This is not a major surprise given that phenotype information was not used during the pre-filtering steps provided by Codicem.
@@
 \section{Conclusion}
-We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report.
+We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report for rare disease patients.
 We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process obtained from the Undiagnosed Diseases Network (UDN).
 From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.
-However, when evaluated as a ranking system, all four methods out-performed single-measure ranking systems and Exomiser.
-We consider the best classifier to be the EasyEnsembleClassifier from {\it imblearn} that had a median rank of 7.0 for all reported variants while ranking 73\% in the top 20 for the case.
-For ``Pathogenic" variants, the median rank was 2.0 and 97\% of those variants were ranked in the top 20 for the case.
+However, when evaluated as a ranking system, all four methods out-performed the single-measure ranking systems and external tools that were tested.
+The classifiers had median ranks of 6.0-10.0 for all reported variants while ranking 65-72\% of those variants in the top 20 for the case.
+For ``Pathogenic" variants, the median ranks were 1.0-4.0 and 80-94\% of those variants were ranked in the top 20 for the case.
 While these algorithms are not perfect classifiers, their use as a prioritization system is quite promising.
 
+There are two major factors that we believe is influencing the classifiers' performance relative to the externally tested tools.
+First, all results were generated using real-world patients from the UDN, but only our four classifiers were trained on real-world patients from the UDN.
+In contrast, the four external tools were primarily evaluated and/or trained using simulations that do not capture the variation and/or noise that is apparent in UDN patients.
+Second, the four classifiers we tested have far more information (i.e.
+features) available to them than the external tools.
+As noted in our methods, we tried to reflect an analyst's view of each variant as much as possible, ultimately resulting in 50 features incorporated into the model.
+Incorporating the same set of features and/or training on real-world patients may improve the externally tested tools with respect to these classifiers.
+
 We expect these classification algorithms could be refined in a variety of ways.
-First, adding new features and/or removing unused features could lead to improvements in the algorithm.
-The top two most important features were both phenotype related suggesting that a highly accurate phenotype scoring system would greatly benefit these algorithms.
-Many features had a low importance, so pruning them may improve the overall results.
-Additionally, some of the features represent data that is not freely available to the research community, so pruning or replacing those features with publicly accessible sources would likely influence the results.
-Second, there may be a better classification algorithm for this type of data.
-The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may improve the result.
-Finally, training the model on different patient populations may yield different models.
-We trained on a fairly diverse patient population
+First, adding new features could lead to increased performance in the classifiers.
+Additionally, some of the features represent data that is not freely available to the research community, so replacing those features with publicly accessible sources would likely influence the results.
+Second, there may be a better classification algorithms for this type of data.
+The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may have better performance.
+Finally, training the classifier on different patient populations will likely yield different results, especially in terms of feature selection and feature importances.
 
-We believe the trained classifiers in VarSight are a significant step forward in reducing the problem of variant fatigue.
-The models improve our ability to prioritize variants despite the variability and uncertainty injected by real-world data.
-Ultimately, we believe implementing these models will enable analysts to assess the best candidate variants first, reducing the time to analyze a case and return molecular diagnoses to patients.
+Overall, we believe the classifiers trained in VarSight represent a significant step forward in tackling real clinical data.
+The tested models improved our ability to prioritize variants despite the variability and uncertainty injected by real-world patients.
+Ultimately, we believe implementing these classifiers will enable analysts to assess the best candidate variants first, reducing the time to analyze a case and return molecular diagnoses to patients.
 
@@
 \bibitem[Bagnall {\it et~al}., 2018]{bagnall2018} Bagnall, Richard D., \textit{et~al.} "Whole genome sequencing improves outcomes of genetic testing in patients with hypertrophic cardiomyopathy." {\it Journal of the American College of Cardiology} 72.4 (2018): 419-429.
+
+%paper describing some aspects of analysis
+\bibitem[Bick {\it et~al}., 2017]{bick2017} Bick, David, \textit{et~al.} "Successful application of whole genome sequencing in a medical genetics clinic." {\it Journal of pediatric genetics} 6.02 (2017): 061-076.
 
@@
 %highly cite PCA from book
-\bibitem[Jolliffe, 2011]{jolliffe2011} Jolliffe, Ian. "Principal component analysis." {\it International encyclopedia of statistical science}. Springer, Berlin, Heidelberg, 2011. 1094-1096.
+%\bibitem[Jolliffe, 2011]{jolliffe2011} Jolliffe, Ian. "Principal component analysis." {\it International encyclopedia of statistical science}. Springer, Berlin, Heidelberg, 2011. 1094-1096.
 

### 2019-03-18 paper/supplements/supplement.tex
finalizing changes based on feature selection

@@
 \usepackage{graphicx}
+\usepackage{longtable}
+\usepackage{multirow}
 \usepackage{rotating}
+\usepackage{url}
 
@@
 
-\section{Filtering}
+\section{Model Features}
+\subsection{Overview}
+There are a total of 50 features used as inputs to the models.
+Two of those features are phenotype-based features and the remaining 48 are derived from Codicem (\url{http://envisiongenomics.com/codicem-analysis-platform/}) extractions.
+
+\subsection{Phenotype Features}
+For both methods, we relied on annotations and ontological structures provided by the Human Phenotype Ontology (HPO).
+Specifically, we used the \texttt{hp.obo} and \url{ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt} corresponding to data version \texttt{releases/2018-10-09}.
+
+\subsubsection{HPO-cosine}
+The first phenotype feature is called the ``HPO-cosine'' score.
+Given $N$ terms in the HPO, a phenotype profile is represented by an $N$-dimensional vector where dimension $p$ is set to 1.0 if the corresponding phenotype term is present in the profile.
+Note that if a phenotype is present in the profile, all ancestor terms as defined by the ontology are automatically set as present as well.
+Given two profiles, we calculate the ``HPO-cosine'' score by performing the dot product calculation on the vectors and normalizing by their lengths.
+Using this measure, values close to 1.0 are highly similar (the vectors are ``pointing'' in similar directions) whereas those close to 0.0 are not (the vectors are mostly perpendicular).
+
+For each gene, we created a phenotype profile based on the HPO annotations relating genes to phenotypes.
+For each patient, we created a phenotype profile and calculated the cosine score between the patient profile and all genes to create an ordered rank from 1 to $G$, where $G$ is the total number of genes annotated by the HPO.
+For each variant, we then found the smallest rank, $R$, of genes tied to the variant and stored the normalized value (${R \over G}$) as the ``HPO-cosine'' feature tied to the variant.
+For precise implementation details, please refer to the source code (\url{https://github.com/HudsonAlpha/VarSight/blob/master/scripts/HPOUtil.py}).
+
+\subsubsection{PyxisMap}
+The second phenotype is derived from PyxisMap (\url{https://github.com/HudsonAlpha/LayeredGraph}).
+We derived our data from PyxisMap v1.2 and we ran the standard installation script that downloads all required data sources on December 19, 2018.
+
+To run PyxisMap, we passed in the patient phenotype terms and obtained an ordered rank of all genes.
+For each variant, we then selected the best rank from genes tied to the variant, normalized the rank, and stored it as the ``PyxisMap'' feature tied to the variant.
+Note that this process is identical to how ranks were handled in the ``HPO-cosine'' feature.
+For precise implementation details, please refer to the source code (\url{https://github.com/HudsonAlpha/VarSight/blob/master/scripts/PyxisMapUtil.py}).
+
+\subsection{Codicem Features}
+Codicem annotates each variant using many different data sources.
+We selected the annotations that we believed were most frequently used by analysts when reviewing a case.
+Each annotation is assigned a feature type corresponding to how that annotation was turned into a feature for the models:
+
+\begin{enumerate}
+\item float - The feature is a single value floating-point decimal that is copied from the annotation.
+\item single - The annotation is a categorical field with a single value per variant.  Each allowed value is assigned a numerical value, and that value is used as the feature.
+%\item multiple - The annotation is a categorical field with multiple values per variant.
+The total number of occurrences of each allowed value is calculated per variant and stored in an array.
+Then, the collection of variant-occurrence arrays is fed into a Principal Component Analysis.
+Each array is then transformed into principal component space, and only the top two component values are added as features to the model.
+This is the only feature type resulting in multiple features (i.e.
+two features) per variant.
+\item multiple - The annotation is a categorical field with multiple values per variant.
+The total number of occurrences of each allowed value is calculated per variant and each is stored as a feature.
+This is the only annotation type resulting in multiple features per variant.
+\item float\_reduce - The annotation is a floating-point decimal with multiple values per variant.
+These float values are combined using a reduction method (typically, minimum or maximum as defined by the JSON configuration) and the single-value result of this reduction is used as the feature.
+\end{enumerate}
+
+The JSON describing features is available on GitHub (\url{https://github.com/HudsonAlpha/VarSight/blob/master/CODI_metadata/fields_metadata.json}), and a selection of that information is reproduced here.
+Table \ref{tab:features} contains a list of each Codicem annotations that was automatically generated from the JSON.
+
+\begin{longtable}{p{.20\linewidth}|p{.20\linewidth}|p{.50\linewidth}}
+\textbf{Feature Label}&\textbf{Feature Type}&\textbf{Description} \\ \hline
+\endhead
+\input{feature_info_rendered.tex}
+\caption{Codicem feature metadata.  This table shows the feature name, feature type, and a brief description of each feature that was fed into the models from Codicem.}
+\label{tab:features}
+\end{longtable}
+
+\section{Hyperparameter Tuning}
+\subsection{Tuning method}
+For each classifier, a selection of hyperparameters were tested based on recommendations from \texttt{sklearn} and \texttt{imblearn}.
+We performed a brute-force search over the entire hyperparameter space using the \texttt{GridSearchCV} method of \texttt{sklearn}.
+This method tries every combination of hyperparameters and selects the method with the best performance as defined by the user.
+For defining performance, we performed stratified 10-fold cross validation (\texttt{StratifiedKFold} in \texttt{sklearn}) on the training data and optimized for best F1-score (see \url{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.f1_score.html} for more info).
+
+\subsection{Tuning results}
+Table \ref{tab:hyperparams} shows the classifiers, hyperparameters, tested values, and the final tuned value for our tests.
+
+\begin{table}
+\begin{tabular}{p{.3\linewidth}|p{.25\linewidth}|p{.2\linewidth}|p{.15\linewidth}}
+\textbf{Classifier}&\textbf{Hyperparameter}&\textbf{Tested Values}&\textbf{Tuned Value} \\ \hline
+\input{hyperparameter_rendered.tex}
+\end{tabular}
+\caption{Hyperparameters tested.
+Each classifier is shown on the left.
+For each classifier, we selected hyperparameters and test values for those parameters based on recommendations provided by the developers of each method.
+The selected tuned value (based on 10-fold cross validation optimizing for F1 score) for each hyperparameter is shown on the right.}
+\label{tab:hyperparams}
+\end{table}
+
+\section{External Tools}
+\subsection{Exomiser}
+\subsubsection{Installation}
+We followed the instructions on the Exomiser website (\url{http://exomiser.github.io/Exomiser/manual/7/installation/}) to install Exomiser CLI v.11.0.0.
+We downloaded the latest files for hg19 (\url{https://data.monarchinitiative.org/exomiser/latest/1811_hg19.zip}).
+Additionally, we downloaded data stores for ReMM v0.3.1 (\url{https://zenodo.org/record/1197579/files/ReMM.v0.3.1.tsv.gz}) and CADD v1.3 (\url{https://krishna.gs.washington.edu/download/CADD/v1.3/whole_genome_SNVs.tsv.gz} and \url{https://krishna.gs.washington.edu/download/CADD/v1.3/InDels.tsv.gz}) for use by Exomiser.
+
+\subsubsection{Execution}
+Exomiser was run twice using two different final prioritization algorithm: hiPhive and hiPhive (human only).  Here is the templated command we used to run exomiser:
+\begin{verbatim}
+java -Xms2g -Xmx4g -jar exomiser-cli-11.0.0.jar --analysis [ymlFN]
+\end{verbatim}
+
+The \texttt{ymlFN} parameter is a templated file with parameters for the VCF filename, the HPO terms, and the output path.
+The template for the hiPhive execution is available at \url{https://github.com/HudsonAlpha/VarSight/blob/master/ExomiserTemplates/hiPhive-template.yml} and the template for the hiPhive (human only) is available at \url{https://github.com/HudsonAlpha/VarSight/blob/master/ExomiserTemplates/hiPhive_human-template.yml}.
+These templates were populated with the corresponding information for each patient using the same variants and HPO terms as were used in the VarSight classifiers.
+
+\subsection{Phen-Gen}
+\subsubsection{Installation}
+We followed the instructions on the Phen-Gen website (\url{http://phen-gen.org}, click on ``DOWNLOAD'') to download Phen-GenV1 and followed directions in the README file to install and run Phen-Gen.
+
+\subsubsection{Execution}
+Phen-Gen takes the HPO terms and VCF file as input to the algorithm.
+For the HPO terms, we created a plain-text file with one HPO term per line as instructed.
+We used the same pre-filtered VCF file as was used in all other external tools.
+
+Phen-Gen also requires extra parameters that were not need by other tools, specifically requiring a pedigree file with sex information, mode of inheritance, and the predictor type.
+For the pedigree file, we simply created a one-member pedigree with the corresponding sex field populated with the annotated sex of the patient.
+For mode of inheritance, we selected dominant as this tended to rank more variants than recessive.
+Finally, we selected the predictor type to be ``genomic'' as opposed to ``coding'' only.
+Here is the templated command we used to run Phen-Gen:
+
+\begin{verbatim}
+perl phen-gen.pl input_phenotype=[hpoFN] input_vcf=[vcfFN] \
+    input_ped=[pedFN] inheritance=0 predictor=0
+\end{verbatim}
+
+\texttt{hpoFN} is the filename of the HPO file, \texttt{vcfFN} is the filename of the VCF file, and \texttt{pedFN} is the filename of the pedigree file.
+
+\subsection{DeepPVP}
+\subsubsection{Installation}
+We followed the instructions on the GitHub README (\url{https://github.com/bio-ontology-research-group/phenomenet-vp}) to download the distribution file for DeepPVP v2.1.1 (\url{https://github.com/bio-ontology-research-group/phenomenet-vp/releases/download/v2.1.1/phenomenet-vp-2.1.zip}).
+We downloaded all datasets required by their instructions including the v2.1 data file (\url{http://bio2vec.net/pvp/data-v2.1.tar.gz}), CADD v1.3 with annotations (\url{http://krishna.gs.washington.edu/download/CADD/v1.3/whole_genome_SNVs_inclAnno.tsv.gz}), and DANN (\url{https://cbcl.ics.uci.edu/public_data/DANN/data/DANN_whole_genome_SNVs.tsv.bgz}).
+We then ran their pre-processing of CADD script (\url{http://www.bio2vec.net/pvp/generate.sh}) and moved data to the correct directories.
+
+We created a fresh installation of Python v2.7.16 and followed the instructions to \texttt{pip install -r requirements.txt} to install all required Python packages.
+However, we found the program was crashing due to incompatibilities between the versions of Keras and Theano.
+To fix the issue, we updated to Keras v2.2.4 and the issue was resolved.
+Due to the relatively high performance of DeepPVP compared to other external tools, we do not believe this change influenced DeepPVP's performance.
+
+\subsubsection{Execution}
+DeepPVP requires a VCF file and the HPO terms to execute.  Here is the templated command used to run DeepPVP:
+
+\begin{verbatim}
+phenomenet-vp -y [pythonPath] -p [csv-HPO] -f [vcfFN] -of [outFN]
+\end{verbatim}
+
+\texttt{pythonPath} is the path to our fresh Python install, \texttt{csv-HPO} is a comma-separated list of HPO terms for the patient, \texttt{vcfFN} is the filename of the VCF file, and \texttt{outFN} is the filename for storing the output.
+
+\section{Codicem Filtering}
 The following sections describes the filter that was passed to Codicem for generating the test and training set variants.

### 2019-03-18 scripts/TestLearners.py

@@ -992,7 +992,7 @@ def generateLaTeXResult(args, d):
     orderedFeatures = []
     classWithFeat = ['RandomForest(sklearn)', 'BalancedRandomForest(imblearn)']
     shortClassNames = ['RF(sklearn)', 'BRF(imblearn)']
-    CUTOFF_IMPORTANCE = 0.02
+    CUTOFF_IMPORTANCE = 0.00
     assert(len(classWithFeat) == len(shortClassNames))
     for i, label in enumerate(d['FEATURE_LABELS']):
         orderedFeatures.append(np.mean([d['CLASSIFIERS'][c]['FEATURE_IMPORTANCE'][i] for c in classWithFeat]))

### 2019-03-19 paper/main.tex
adding supplements template files

@@
 \newif\ifpreprint
-%\preprinttrue
+\preprinttrue
 
@@
 \\
-\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.
+\textbf{Results:} We tested the application of classification algorithms that ingest variant annotations along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.
 To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.
-We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls.
+We treated the classifiers as variant prioritization systems and compared them to four variant prioritization algorithms and two single-measure controls.
 We showed that these classifiers outperformed the other methods with the best classifiers ranking 72\% of all reported variants and 94\% of reported pathogenic variants in the top 20.
 \\
-\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
+\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight} release v1.1.
+\\
 \textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
@@
 \\
-\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.
+\textbf{Results:} We tested the application of classification algorithms that ingest variant annotations along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.
 To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.
-We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls.
+We treated the classifiers as variant prioritization systems and compared them to four variant prioritization algorithms and two single-measure controls.
 % to show how the classifiers tend to outperform the other methods.
@@
 \\
-\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
+\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight} release v1.1. \\
 \textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
@@
 Unfortunately, manual curation is a time consuming process where analysts must inspect each variant while maintaining a mental picture of the patient's phenotype.
-\citealp{bick2017} report an average of 600 variants per case analyzed by two people (one analyst and one director) over three hours ($\approx$100 variants per man-hour).
-If causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited to reduce the time needed to analyze a case.
-Finally, accurate prioritization is a step towards the ultimate goal of automatic identification of rare variants that may cause a patient's primary phenotypes.
+\citealp{bick2017} report an average of 600 variants per case analyzed by two people (one analyst and one director) over three hours, meaning a thoughput of $\approx$100 variants per man-hour.
+If causative variants can be identified earlier due to a high rank from prioritization, it's possible that the full filtered variant list can be short-circuited, reducing the total number of variants reviewed and therefor the time to analyze a case.
+Additionally, accurate prioritization is a step towards the ultimate goal of automatic identification of rare variants that may cause a patient's primary phenotypes.
 
@@
 Additionally, when phenotype terms were used, they tended to use all available phenotype terms paired with the simulated disease with a few (typically 2-3) noisy terms added or removed.
-Unfortunately, rare disease patient data is often far noisier due to a wide variety of reasons.
+In practice, rare disease patient data is often far noisier due to a wide variety of reasons.% such as multiple diseases, incomplete penetrant phenotypes, or age-related phenotypes not caused by the primary disease.
 
@@
 Because the UDN is not focused on a single particular disease, the patient population has a diverse range of phenotypes represented.
-Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as phenotypes caused by old age).
+Additionally, the phenotypes associated to an individual patient can be quite noisy for a variety of reasons: multiple genetic diseases, phenotype collection differences, and/or unrelated non-genetic diseases (such as age- or pathogen-related disease).
 Because the UDN is a research collaboration, there is also variability in reported variants that range in pathogenicity from ``variant of uncertain significance" (VUS) through ``pathogenic" as defined by the ACMG guidelines \citep{richards2015}.
@@
 We tested four classifiers that are readily available in the {\it sklearn} \citep{pedregosa2011} and {\it imblearn} \citep{lemaitre2017} Python modules.
-Of note, our focus was not on picking the ``best'' classifier, but rather on analyzing their overall ability of classification algorithms to handle the noise of real-world patient cases from the UDN.
+Of note, our focus was not on picking the ``best'' classifier, but rather on analyzing their overall ability to handle the noise of real-world patient cases from the UDN.
 
-Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing us to measure their performance as a classifier and as a ranking system.
-After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant ranking perspective.
-Additionally, we tested four publicly available variant ranking algorithms and two single-value ranking methods for comparison.
+Each classifier calculated probabilities of a variant belonging to the ``returned" class, allowing us to measure their performance as both a classifier and a prioritization/ranking system.
+After tuning each classifier, we generated summaries of the performance of each method from both a binary classification perspective and a variant prioritization perspective.
+Additionally, we tested four publicly available variant prioritization algorithms and two single-value ranking methods for comparison.
 All of the scripts to train classifiers, test classifiers, and format results are contained in the VarSight repository.
@@
 \subsection{Data sources}
-All samples were selected from the cohort of Undiagnosed Diseases Network (UDN) \citep{ramoni2017} genome sequencing samples that were originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB).
+All samples were selected from the cohort of Undiagnosed Diseases Network (UDN) \citep{ramoni2017} genome sequencing samples that were sequenced at HudsonAlpha Institute for Biotechnology.
 In short, the UDN accepts patients with rare, undiagnosed diseases that are believed to have a genetic origin.
 The UDN is not restricted to a particular disease, so there are a diverse set of diseases and phenotypes represented across the whole population.
-The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) some patients have multiple diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, and 3) some patients exhibit more or fewer phenotypes than are associated with the classic disease presentation.
+The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) patients may have multiple genetic diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, 3) patients may exhibit more or fewer phenotypes than are associated with the classic disease presentation, and 4) patients may have phenotypes of non-genetic origin such as age- or pathogen-related phenotypes.
 For more details on the UDN, refer to \citealp{ramoni2017}.
@@
 The first method uses phenotype-to-gene annotations provided by the HPO to calculate a cosine score \citep{kohler2017} between the patient's phenotypes and each gene.
-In short, this method builds a vector from the phenotype terms (and ancestor terms in the ontology) for both the patient and each gene.
+Given $P$ terms in the HPO, this method builds a binary, $P$-dimensional vector for each patient such that only the phenotype terms and all ancestral phenotype terms tied to the patient are set to 1.
+Similarly, a $P$-dimensional vector for each gene is built using the phenotype-to-gene annotations.
 Then, the cosine of the angle between the patient vector and each gene vector is calculated as a representation of similarity.
@@
 it was not found through some other targeted search) and 2) a set of phenotypes annotated with Human Phenotype Ontology \citep{kohler2018} terms using the Phenotips software \citep{girdea2013}.
-At the time of writing, this amounted to 378 primary, reported variants spanning a total of 237 proband samples.
+At the time of writing, this amounted to 378 primary-reported variants and 87819 unreported variants spanning a total of 237 proband samples.
 
@@
 This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.
-After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).
+After training, the two random forest classifiers report the importance of each feature in the classifier (total is 1.00 per classifier).
 We average the two importance values, and order them from most to least important.
@@
 \processtable{Feature selection.
-This table shows the top 20 features that were used to train the classifiers ordered from most important to least important.
-After training, the two random forest models report the importance of each feature in the model (total is 1.00 per classifier).
-We average the two importance values, and order them from most to least important.
+This table shows the 20 features gathered using univariate feature selection that were used to train the classifiers.
+After training, the two random forest classifiers report the importance of each feature in the classifier (total is 1.00 per classifier).
+We average the two importance values, and order them from most to least important according to this average.
 Feature labels with an `\_' represent a single category of a multi-category feature (i.e.
@@
 The one exception to this rule was for variant allele frequencies where a variant absent from the database was considered to have an allele frequency of 0.0.
-For multi-value numerical annotations, we reduced the values (using minimum or maximum) to a single value that was used as the feature.
+For multi-value numerical annotations, we reduced the values (using minimum or maximum) to a single value corresponding to the ``worst'' value (i.e.
+most deleterious value, most conserved value, etc.) that was used as the feature.
 
@@
 After converting all annotations to numerical features or multiple bin-count encoded features, we had a total of 95 features per variant.
-We then pruned down to only the top 20 features using univariate feature selection (specifically the \texttt{SelectKBest} method of \texttt{sklearn}).
+We then pruned down to only the top 20 features using univariate feature selection (specifically the \texttt{SelectKBest} method of \textit{sklearn} \citep{pedregosa2011}).
 This method evaluates how well an individual feature performs as a classifier and keeps only the top 20 features for the full classifiers.
 Table \ref{tab:features} shows the list of retained features ordered by feature importance after training.
-Feature importance was derived from the random forest models which automatically report how important each feature was to classification.
+Feature importance was derived from the random forest classifiers which automatically report how important each feature was for classification.
 The entire set of annotations along with descriptions of how each was processed prior to feature selection are detailed in the Supplementary Documents.
 
-\subsection{Model training and tuning}
+\subsection{Classifier training and tuning}
 As noted earlier, there are generally hundreds of variants per proband that pass the filter, but only a few are ever clinically reported.
@@
 In the test set, there were a total of 189 returned variants and 43226 not returned variants spanning 117 different probands.
-In our results, the returned test variants are further stratified in their reported levels of pathogenicity.
+In our results, the returned test variants are further stratified by their reported levels of pathogenicity.
 
-We then selected four publicly available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.
+We then selected four publicly available binary-classification models that are capable of training on imbalanced datasets: the \texttt{RandomForest} model by {\it sklearn} \citep{pedregosa2011}, the \texttt{LogisticRegression} model by {\it sklearn}, the \texttt{BalancedRandomForest} model by {\it imblearn} \citep{lemaitre2017}, and the \texttt{EasyEnsembleClassifier} model by {\it imblearn}.
 These classifiers were chosen for three main reasons: 1) their ability to handle imbalanced data (i.e.
-far more unreported variants than reported variants), 2) their ability to scale to the size of the training and testing datasets, and 3) they are freely available implementations that can be tuned and tested with relative ease.
-The two random forest models build collections of decision trees that weight each training input by its class frequency.
+far more unreported variants than reported variants), 2) their ability to scale to the size of the training and testing datasets, and 3) they are freely available implementations that can be tuned, trained, and tested with relative ease in the same Python framework.
+The two random forest classifiers build collections of decision trees that weight each training input by its class frequency.
 Logistic regression calculates the probability of a value belonging to a particular class, again weighting by the class frequency.
 In contrast to the other three tested methods, the ensemble classification balances the training input using random under-sampling and then trains an ensemble of AdaBoost learners.
-For more details on each classifier, please refer to the {\it sklearn} and {\it imblearn} documentations.
+For more details on each classifier, please refer to the {\it sklearn} and {\it imblearn} documentations \citep{pedregosa2011, lemaitre2017}.
 
-Initially, we also tested the support vector classifier by {\it sklearn}, the multi-layer perceptron by {\it sklearn}, and the random under-sampling AdaBoost classifier by {\it imblearn}.
+Initially, we also tested the support vector classifier by {\it sklearn} (\texttt{SVC}), the multi-layer perceptron by {\it sklearn} (\texttt{MLPClassifier}), and the random under-sampling AdaBoost classifier by {\it imblearn} (\texttt{RUSBoostClassifier}).
 Each of these was excluded from our results due to, respectively, scaling issues with the training size, failure to handle the data imbalance, and overfitting to the training set.
@@
 
-For each of our tested models, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.
-For each model and set of hyperparameters, we performed stratified 10-fold cross validation on the training variants and recorded the balanced accuracy (i.e.
+For each of our tested classifiers, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.
+For each classifier and set of hyperparameters, we performed stratified 10-fold cross validation on the training variants and recorded the balanced accuracy (i.e.
 weighted accuracy based on inverse class frequency) and the F1 scores (i.e.
 harmonic mean between precision and recall).
-For each model type, we saved the hyperparameters and model with the best average F1 score.
-These four tuned models were then trained on the full training set and tested against the unseen set of test proband cases.
+For each classifier type, we saved the hyperparameters and classifier with the best average F1 score (this is recommended for imbalanced datasets).
+These four tuned classifiers were then trained on the full training set and tested against the unseen set of test proband cases.
 The set of hyperparameters tested along with the highest performance setting for each hyperparameter can be found in the Supplementary Documents.
@@
 On the right, we show the precision recall curve.
-Area under the curve (AUROC and AUPRC) is reported beside each method in the legend.}
+Area under the curve (AUROC or AUPRC) is reported beside each method in the legend.}
 \label{fig:roc}
@@
 
-Each model was tuned using 10-fold cross validation and the resulting average and standard deviation of balanced accuracy is reported in Table \ref{tab:classifier_results}.
-After fitting the tuned model to the full training set, we evaluated the models on the testing set by calculating the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC).
-Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the results from the testing set on all four models.
+The hyperparameters for each classifier were tuned using 10-fold cross validation and the resulting average and standard deviation of balanced accuracy is reported in Table \ref{tab:classifier_results}.
+After fitting the tuned classifiers to the full training set, we evaluated the classifiers on the testing set by calculating the area under the receiver operator curve (AUROC) and area under the precision-recall curve (AUPRC) (also shown in Table \ref{tab:classifier_results}).
+Figure \ref{fig:roc} shows the corresponding receiver operator curves and precision-recall curves for the results from the testing set on all four classifiers.
 
-From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.
-However, these classifier all have relatively poor performance from a precision-recall perspective (best AUPRC was 0.2109).
-This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.
+From these metrics, we can see that all four classifiers have a similar performance with regards to AUROC.
+However, all classifiers have a relatively poor performance from a precision-recall perspective (best AUPRC was only 0.2458).
+This indicates that from a classification perspective, these classifiers would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.
 
@@
 
-In addition to the model performance statistics, we also quantified the performance of each classifier as a ranking system.
+In addition to the classifier performance statistics, we also quantified the performance of each classifier as a ranking system.
 For each proband, we calculated the probability of each class (reported or not reported) for each variant and ordered them from highest to lowest probability of being reported.
@@
 Additionally, we quantified the percentage of reported variants that were ranked in the top 1, 10, and 20 variants in each case.
-While the models were trained as a binary classification system, we broke down the results further to demonstrate differences between variants that were clinically reported as a variant of uncertain significance (VUS), likely pathogenic, and pathogenic.
+While the classifiers were trained as a binary classification system, we broke down the results further to demonstrate differences between variants that were clinically reported as a variant of uncertain significance (VUS), likely pathogenic, and pathogenic.
 
@@
 %After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.
-Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).
-The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.
-By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.
+Both random forest classifiers calculate a feature importance array that stores how important each feature is in the trained classifier (the other two classifiers do not have this information readily available).
+The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained classifier.
+By chance, we had exactly 50 parameters as input to our classifier, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.
 Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.
@@
 We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report for rare disease patients.
-We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process obtained from the Undiagnosed Diseases Network (UDN).
+We trained and tested these algorithms using real patient variants and phenotype terms obtained from the Undiagnosed Diseases Network (UDN).
 From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.
 However, when evaluated as a ranking system, all four methods out-performed the single-measure ranking systems and external tools that were tested.
-The classifiers had median ranks of 6.0-10.0 for all reported variants while ranking 65-72\% of those variants in the top 20 for the case.
+The classifiers had median ranks of 6.0-10.0 for all reported variants and ranked 65-72\% of those variants in the top 20 for the case.
 For ``Pathogenic" variants, the median ranks were 1.0-4.0 and 80-94\% of those variants were ranked in the top 20 for the case.
@@
 features) available to them than the external tools.
-As noted in our methods, we tried to reflect an analyst's view of each variant as much as possible, ultimately resulting in 50 features incorporated into the model.
+As noted in our methods, we tried to reflect an analyst's view of each variant as much as possible, starting with 95 features that were pruned down to 20 features used by each classifier.
 Incorporating the same set of features and/or training on real-world patients may improve the externally tested tools with respect to these classifiers.
@@
 Overall, we believe the classifiers trained in VarSight represent a significant step forward in tackling real clinical data.
-The tested models improved our ability to prioritize variants despite the variability and uncertainty injected by real-world patients.
-Ultimately, we believe implementing these classifiers will enable analysts to assess the best candidate variants first, reducing the time to analyze a case and return molecular diagnoses to patients.
+The tested classifiers improved our ability to prioritize variants despite the variability and uncertainty injected by real-world patients.
+Ultimately, we believe implementing these classifiers will enable analysts to assess the best candidate variants first, allowing for faster clinical throughput and increased automation in the future.
 

### 2019-03-19 paper/supplements/hyperparameter_template.tex
adding supplements template files

{% for clf_label in CLF_LABELS %}
    {{ '\\multirow{%d}{*}{\\parbox{1.0\\linewidth}{%s}}' % (len(CLASSIFIERS[clf_label]['BEST_PARAMS']), clf_label.replace('(', ' (')) }} 
    {% for k in CLASSIFIERS[clf_label]['BEST_PARAMS'] %}
        &{{ k.replace('_', '\\_') }}&{{ CLASSIFIERS[clf_label]['HYPERPARAMETER_SPACE'][k] }}&{{ CLASSIFIERS[clf_label]['BEST_PARAMS'][k] }} \\ \cline{2-4}
    {% endfor %}
    \hline
{% endfor %}

%resultsDict['CLASSIFIERS'][clf_label]['BEST_PARAMS'] = clf.best_params_
%resultsDict['CLASSIFIERS'][clf_label]['HYPERPARAMETER_SPACE'] = hyperparam

### 2019-03-19 paper/supplements/supplement.tex
adding supplements template files

@@
 The first phenotype feature is called the ``HPO-cosine'' score.
-Given $N$ terms in the HPO, a phenotype profile is represented by an $N$-dimensional vector where dimension $p$ is set to 1.0 if the corresponding phenotype term is present in the profile.
+Given $P$ terms in the HPO, a phenotype profile is represented by an $P$-dimensional vector where dimension $p$ is set to 1.0 if the corresponding phenotype term is present in the profile.
 Note that if a phenotype is present in the profile, all ancestor terms as defined by the ontology are automatically set as present as well.

### 2019-04-10 paper/data_template.tex
fixed templating with pathogenic only

--- 2019-01-15 paper/data_template.tex
@@
 &All (n={{ TEST_COUNTS.OVERALL }})&
-VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
+VUS (n={{ TEST_COUNTS.get('VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 0) }})&
 LP (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Path. (n={{ TEST_COUNTS.PATHOGENIC }})&
 All (n={{ TEST_COUNTS.OVERALL }})&
-VUS (n={{ TEST_COUNTS.VARIANT_OF_UNCERTAIN_SIGNIFICANCE }})&
+VUS (n={{ TEST_COUNTS.get('VARIANT_OF_UNCERTAIN_SIGNIFICANCE', 0) }})&
 LP (n={{ TEST_COUNTS.LIKELY_PATHOGENIC }})&Path. (n={{ TEST_COUNTS.PATHOGENIC }}) \\ \hhline{=|=|=|=|=||=|=|=|=}
@@
     {% for LEVEL in LEVELS %}
+        {% if TEST_COUNTS.get(LEVEL, 0) == 0 %}
+            -- (--)&
+        {% else %}
             {{ '%0.1f (%0.2f)' % (np.median(COMPARISON[cs_label].TEST_RANKINGS[LEVEL]), np.mean(COMPARISON[cs_label].TEST_RANKINGS[LEVEL])) }}&
+        {% endif %}
     {% endfor %}
     {% for LEVEL in LEVELS %}
+        {% if TEST_COUNTS.get(LEVEL, 0) == 0 %}
+            --
+        {% else %}
             {{ '%d, %d, %d' % 
@@
             }}
+        {% endif %}
         {% if LEVEL != 'PATHOGENIC' %}
@@
 
-Exomiser$^*$&
+{% for ext_label in EXT_LABELS %}
+    {{ ext_label.replace('_', '\\_') }}&
     {% for LEVEL in LEVELS %}
-    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER.TEST_RANKINGS[LEVEL])) }}&
+        {% if TEST_COUNTS.get(LEVEL, 0) == 0 %}
+        -- (--)&
+        {% else %}
+            {{ '%0.1f (%0.2f)' % (np.median(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]), np.mean(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL])) }}&
+        {% endif %}
     {% endfor %}
     {% for LEVEL in LEVELS %}
+        {% if TEST_COUNTS.get(LEVEL, 0) == 0 %}
+            --
+        {% else %}
             {{ '%d, %d, %d' % 
-        (100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
-        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
-        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER.TEST_RANKINGS[LEVEL]))
+                (100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 1) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]),
+                100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 10) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]),
+                100.0*np.sum(np.array(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]) <= 20) / len(EXTERNAL[ext_label].TEST_RANKINGS[LEVEL]))
             }}
-    {% if LEVEL != 'PATHOGENIC' %}
-        &
         {% endif %}
-{% endfor %}
-\\ \hhline{=|=|=|=|=||=|=|=|=}
-
-{% for clf_label in CLF_LABELS %}
-    {{ clf_label }}&
-    {% for LEVEL in LEVELS %}
-        {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL])) }}&
-    {% endfor %}
-    {% for LEVEL in LEVELS %}
-        {{ '%d, %d, %d' % 
-            (100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 1) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
-            100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 10) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
-            100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 20) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]))
-        }}
         {% if LEVEL != 'PATHOGENIC' %}
@@
 {% endfor %}
+\hhline{=|=|=|=|=||=|=|=|=}
+
+{% for clf_label in CLF_LABELS %}
+    {{ clf_label }}&
+    {% for LEVEL in LEVELS %}
+        {% if TEST_COUNTS.get(LEVEL, 0) == 0 %}
+        -- (--)&
+        {% else %}
+            {{ '%0.1f (%0.2f)' % (np.median(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]), np.mean(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL])) }}&
+        {% endif %}
+    {% endfor %}
+    {% for LEVEL in LEVELS %}
+        {% if TEST_COUNTS.get(LEVEL, 0) == 0 %}
+            --
+        {% else %}
+            {{ '%d, %d, %d' % 
+                (100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 1) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
+                100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 10) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]),
+                100.0*np.sum(np.array(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]) <= 20) / len(CLASSIFIERS[clf_label].TEST_RANKINGS[LEVEL]))
+            }}
+        {% endif %}
+        {% if LEVEL != 'PATHOGENIC' %}
+            &
+        {% endif %}
+    {% endfor %}
+    \\
+{% endfor %}

### 2019-08-06 paper/supplements/supplement.tex
updating some info in the supplements

@@
 \documentclass{article}
+\usepackage[margin=1in]{geometry}
 \usepackage{graphicx}
@@
 \tableofcontents
+
+\pagebreak
+\section{Overall Workflow}
+Figure \ref{fig:workflow_outline} contains a visualization of the overall workflow for training VarSight models.  For details, refer to the main paper.
+
+\begin{figure}[!b]
+\centering
+\includegraphics[width=.50\textwidth]{workflow_outline.png}
+\caption{Training Workflow. This is a visualization of the workflow for training VarSight models.}
+\label{fig:workflow_outline}
+\end{figure}
 
@@
 
+\pagebreak
+\section{Results Extended}
+\subsection{Ranking Boxplot}
+Figure \ref{fig:results_boxplot} shows the rankings from our experimental results in a boxplot format.
+
+\begin{figure}[!b]
+\centering
+\includegraphics[width=.9\textwidth]{results_boxplot.png}
+\caption{Results Boxplot.
+This image shows the rankings from our experiments in a boxplot format (see Table 3 and 4 in main paper for summarized results).
+Each reported variant rank is a red circle on the corresponding method with an overlayed, standard boxplot for all rankings for the method.
+Note, the values in the figure represent only the ranking of reported variants in the test set.}
+\label{fig:results_boxplot}
+\end{figure}
+
+\pagebreak
 \section{Codicem Filtering}

### 2019-08-06 scripts/TestLearners.py

@@ -24,9 +24,10 @@ from imblearn.ensemble import BalancedRandomForestClassifier, RUSBoostClassifier
 
 from sklearn.decomposition import PCA
 from sklearn.ensemble import RandomForestClassifier
-from sklearn.feature_selection import SelectKBest
+from sklearn.ensemble import ExtraTreesClassifier
+from sklearn.feature_selection import SelectKBest, SelectFdr, SelectFpr, SelectFwe
+from sklearn.feature_selection import SelectFromModel
 from sklearn.feature_selection import mutual_info_classif
-from sklearn.feature_selection import RFE
 from sklearn.linear_model import LogisticRegression
 from sklearn.metrics import auc
 from sklearn.metrics import average_precision_score
@@ -43,6 +44,8 @@ from sklearn.model_selection import RandomizedSearchCV
 from sklearn.model_selection import StratifiedKFold
 from sklearn.model_selection import train_test_split
 from sklearn.neural_network import MLPClassifier
+from sklearn.preprocessing import StandardScaler
+from sklearn.svm import LinearSVC
 
 #CONSTANTS
 PYXIS_DATE='01042019'   
@@ -453,20 +456,66 @@ def runClassifiers(args, values, classifications, featureLabels, startIndices, a
         (or maybe amount) of pruning.  I think leave it out for now unless this becomes a deal-breaker with some
         ornery reviewer.
         '''
-        #'''
         #more details here: https://scikit-learn.org/stable/modules/feature_selection.html
         #doing this seems to help with LogisticRegression only, but doesn't bring it up to our current results
         #select the best parameters based on training only
-        skb = SelectKBest(k=20)
-        skb.fit(train_x, train_y)
-        selected = skb.get_support(False)
-        selected2 = skb.get_support(True)
-        train_x = train_x[:, selected]
-        test_x = test_x[:, selected]
-        featureLabels = [featureLabels[x] for x in selected2]
-        resultsDict['FEATURE_LABELS'] = featureLabels
-        #'''
+        FS_NONE = 0
+        FS_SELECT_K_BEST = 1
+        #FS_RFECV = 2 #this one took forever to run, and underperformed
+        FS_MODEL_SELECTION = 3
+        FEATURE_SELECTION_TYPE = FS_SELECT_K_BEST
 
+        if FEATURE_SELECTION_TYPE == FS_NONE:
+            pass
+
[…]
